### 2022-11-22 main.tex
Initial Overleaf Import

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts

\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{tabularx}

\newcommand{\chameleon}{ChameleonIDE}
\newcommand{\pjs}[1]{\textcolor{blue}{\textsc{Pjs}: #1}}
\newcommand{\jw}[1]{\textcolor{orange}{\textsc{JW}: #1}}
\newcommand{\td}[1]{\textcolor{teal}{\textsc{td}: #1}}
\newcommand{\tf}[1]{\textcolor{magenta}{\textsc{tf}: #1}}

\def\BibTeX{{\rm B\kern-.05em{\sc i\kern-.025em b}\kern-.08em
    T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
\begin{document}

\title{\chameleon{}: Untangling Type Errors Through Interactive
  Visualization and Exploration
}

\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
\and
\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
\textit{name of organization (of Aff.)}\\
City, Country \\
email address or ORCID}
}

\maketitle

\begin{abstract}
In recent years, many programming communities have been undergoing a transition to writing strongly typed programs. To benefit from type-safe codebases, programmers must first overcome the challenge of resolving type errors where text messages can often be unhelpful and misleading. Modern programming tools do not concentrate on making type errors easy to understand. We demonstrate \chameleon{}, a type debugging tool capable of displaying the full context of a type error and allowing programmers to search for where the type error actually occurs based on their prior knowledge and to incrementally verify their assumptions. \chameleon{} was developed through a user-centred, iterative process with four participatory experiments. Through the experiments, we showed that programmers using \chameleon{} fix type errors faster than traditional text-based error messages.  This difference is more significant in harder tasks. Further, programmers actively using \chameleon{} interactive features are shown to be more efficient in fixing type errors than simply reading the type error output.
\end{abstract}

\begin{IEEEkeywords}
component, formatting, style, styling, insert
\end{IEEEkeywords}

\input{sections/introduction.tex}
\input{sections/related-work.tex}
\input{sections/chameleon.tex}
\input{sections/walkthrough.tex}
\input{sections/evaluation.tex}
\input{sections/discussion.tex}
\input{sections/conclusion.tex}

\bibliographystyle{IEEEtran}
\bibliography{reference}

\end{document}

### 2022-11-22 sections/chameleon.tex
Initial Overleaf Import

\section{Chameleon IDE} \label{chameleon}

\chameleon{} comprises two parts: a type inference engine and an original debugging interface. This separate architecture allows us to 1) adapt \chameleon{} into multiple front ends, such as IDE extensions, desktop applications, and web applications, and 2) reuse the debugging interface for different programming languages by replacing the back end type inference engine. The debugging interface is designed from the ground up; the type inference engine is a re-implementation of the original Chameleon with several novel improvements.

\subsection{The Type Inference Engine}
\label{sec:typeinferenceengine}

Chameleon was originally a command-line tool developed in the early 2000s to improve type error reporting and introduced a design for interactive type debugging for the Haskell programming language. Unlike traditional type errors produced by the Glasgow Haskell Compiler (GHC)~\cite{ghc}, which uses a Hindley–Milner type inference system, Chameleon infers types using constraint solving. In Chameleon, Constraints are generated from the source code based on typing rules. In addition, each constraint is labeled with the location (line and column numbers) where it is generated. This set of constraints is consistent if the program is well-typed, inconsistent if otherwise. When a type error occurs, an efficient algorithm is used to derive a minimal subset of the constraints that still contains inconsistencies. This subset is called a Minimal Unsatisfiable Subset (MUS). From this, Chameleon can report a list of locations, using the labels of constraints that are in the MUS. Stuckey et al.~\cite{stuckey2003interactive} showed that program locations linked to the constraints from a MUS are all relevant to the type error and must include the cause of the error. In this paper, we will refer to these locations as \textit{error locations}.

Despite successfully borrowing the underlying ideas, we could not reuse the implementation of Chameleon due to the project language standard and libraries used was out of date. In addition to what Chameleon can do, a few new type inference features were added in the \chameleon{} implementation.

\paragraph{Recovering concrete types from type errors}

Using only constraints from the MUS is sufficient to locating the type error, but to recover types from type errors we need the constraints from parts of the program that irrelevant to the type errors.  For instance, consider an ill-typed 2-tuple where two possible types can be assigned: \texttt{(Int, Int)} and \texttt{(Int, String)}. The types reconstructed from Chameleon may be \texttt{(a, Int)} and \texttt{(a, String)}. Although the recovered types are theoretically correct, they introduce new type variable \texttt{a} making the error message harder to understand.  To solve this issue in \chameleon{}, we first expand the MUS to a maximally satisfiable subset. We do so by removing one constraint from the MUS to get a consistent set, and then repeatedly adding constraints from the global constraint store while maintaining consistency. This process stops when no more constraints can be added. The resulting subset, while not helpful to inform error locations, will produce the most concrete conflicting types. This difference can be seen in figure~\ref{fig:compare-to-original}.

\begin{figure}[h]
    \centering
    \Description{Comparing Chameleon with \chameleon{}}
    \includegraphics[width=\linewidth]{images/compare-to-original.pdf}
    \caption{
Reporting the same type error, Chameleon uses
\texttt{Int -> a} and \texttt{Char -> a}, while \chameleon{} uses the 
concrete types \texttt{Int -> Bool} and \texttt{Char -> Bool}.
    }
    \label{fig:compare-to-original}
\end{figure}

This has a few benefits. First, concrete types provide extra information to programmers. With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space. However, this information is not preserved in \texttt{(a, Float)}. Second, with concrete types, the debugging front end has more flexibility to change the representation of types. If the front end
understands the type language in Haskell, it is possible to recognize and replace ground facts that are overlapping in both possible types and have the same effect. The other direction is impossible.

\paragraph{Type error explanation}

In addition, \chameleon{} provides support for type explanation. Similar to the type explanation system ~\cite{jun2002explaining} from Yang,  \chameleon{} is able to produce a human-readable explanation, but for type errors. This is achieved by expanding the metadata fields of each constraint to contain the information needed for type explanation.

\begin{lstlisting} [
    language=Haskell, caption={
    A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
}, label={listing:ifelse}
]
if a then b else c
a = "True"
\end{lstlisting}
For instance, the program in the listing \ref{listing:ifelse} \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition). Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed. \chameleon{} can construct a human-readable explanation from the MUS, for example: \texttt{a} has type \texttt{Bool}, because \texttt{a} is the condition of an if statement; however \texttt{a} has type \texttt{String}, because \texttt{a} is defined as the string literal \texttt{"True"}. This low-level explanation is used in a high-level user interface described in section \ref{sub:deduction-steps} to help programmers reason about the validity of error locations.

\begin{figure}[h]
    \centering
    \Description{An anatomy of \chameleon{} interface}
    \includegraphics[width=\linewidth]{images/atonomy.png}
    \caption{
        \textbf{The anatomy of \chameleon{}.}
        \chameleon{} consists of two windows: an editor window on the left and a
        debugging window on the right. The editor window is similar to a
        traditional code editor. Fragments of source code may have a highlight
        color (A). Additionally, an explanation layer (B) displays in the editor
        window, if deduction steps are enabled. The debugging window contains
        three panels. A message panel contains an error description (C), and, optionally,
        a list of candidate expression cards (D), a list of deduction steps (E), 
        and a control bar (F) to increment/decrement deduction step. 
        The conflicting types
        panel (G) shows two possible types inferred by the type inference
        engine. A relevant type information panel to show additional information
        that may help understand type errors.
    }
    \label{fig:anatomy}
\end{figure}

\subsection{The Debugging Interface}

The \chameleon{} debugging interface provides three main features to visualize the relation between an error statement and the error locations in the source code, and to explore the explanation of each error location given by the type inference engine. Further, programmers  can turn on and off features to suit their preferences and debugging needs.

\paragraph{Type compare tool} \label{sub:type-compare}

The type compare tool shows the two conflicting types in different colors, each associated with one or more error locations (Fig. \ref{fig:compare}).  The type error must be fixed by modifying at least one of these locations. The error locations are highlighted with the matching color of the conflicting types. The type compare tool is useful to quickly bisect the type error. If the programmers know the expression's intended type (they usually do), they will be able to eliminate half of the possible locations. To make the bisecting effect more pronounced, a user can hover on one of the possible types and only show the relevant locations that
contribute to that type. This is a convenient way to put the error "under the spotlight".

\begin{figure}[h]
    \centering
    \Description{Using \chameleon{} Type Compare Tool}
    \includegraphics[width=\linewidth]{images/intro-compare.pdf}
    \caption{
        \textbf{\chameleon{} minimal output (with candidate expression cards and deduction step disabled)}. \chameleon{} identified the conflicting types for the function \texttt{a} and associated the relevant locations with each type. In addition, \chameleon{} leaves out the right-hand side of the declaration '\texttt{0}' and '\texttt{1}' because they are unrelated.
}
    \label{fig:compare}
\end{figure}

\begin{figure}
    \centering
    \begin{subfigure}[b]{\linewidth}
        \centering
        \Description{Using \chameleon{} Candidate Expression Cards}
        \includegraphics[width=\linewidth]{images/intro-expression.pdf}
        \caption{
            \textbf{\chameleon{} with only candidate expression cards enabled.}
            \chameleon{} shows that the type error can
            occur in the definition of '\texttt{x}' or '\texttt{y}'.
        }
        \label{fig:expression}
    \end{subfigure}
    \hfill
    \begin{subfigure}[b]{\linewidth}
        \centering
        \Description{Using \chameleon{} Deduction Steps}
        \includegraphics[width=\linewidth]{images/intro-deductionstep.pdf}
        \caption{
            \textbf{\chameleon{} with both candidate expression cards and deduction steps
            enabled.}
            \chameleon{} explains the type error in four steps. In the screenshot, the active step is step 2, where \chameleon{} shows that the expression \texttt{x} and \texttt{y} should have the same type. 
        }
        \label{fig:deduction}
    \end{subfigure}
    \hfill
    \caption{ Candidate Expression Cards and Deduction Steps}
\end{figure}

\paragraph{Candidate Expression Cards}  \label{sub:candidate-expression}

A candidate expression is an expression that can be inferred to have two conflicting types. 
We propose candidate expression cards to make it clear to the user that there are multiple ways to resolve a type conflict. By contrast,  standard compiler error messages such as those of GHC arbitrarily focus the users attention on only one candidate expression, e.g.\ ``\textit{Couldn't match expected type ‘Char’ with actual type ‘Bool’ in expression x}" here x is the candidate expression. In practice this candidate expression often does not match programmers' intention (Fig. \ref{fig:single-candidate}).

When a type error is detected, \chameleon{} provides a list of all candidate expressions and programmers are free to choose the problem to resolve by clicking on one candidate expression card. In the example shown in Fig. \ref{fig:expression}, \texttt{x} and \texttt{y} are both candidate expressions. Indeed, fixing one type error will make both expressions well-typed.

\begin{figure}[h]
    \centering
    \Description{Showing an type error with a single candidate}
    \includegraphics[width=\linewidth]{images/single-error-location-example.png}
    \caption{
With traditional type error messages, one expression is blamed for causing the type error. This expression may not always match the programmers' intention. 
    }
    \label{fig:single-candidate}
\end{figure}

Selecting a candidate expression card is a way for programmers to communicate to the type checker the knowledge that the other candidate expressions are in fact well-typed. Once a card is selected, the information in the conflicting types panel changes according to the change of candidate expression. In the editor window, some error locations change highlight colors to reflect the updated knowledge of the program.

This feature can be effective in resolving the problem that traditional tools tend to report errors that happened in source code from third party libraries. With candidate expression cards, a programmer can freely change the context of the type error to the user defined variables and functions to gain a better understanding of why her own code does not match the library code instead of the other direction.

Programmers interact with the candidate expression cards by clicking on them to commit the change of meaning or hovering on them to peek at the effect of the change; the effect is reverted once the cursor moves away.

\paragraph{Deduction steps}  \label{sub:deduction-steps}

Deduction steps are motivated by the lack of explain why the program has  type errors. When compiler reject a program, it dumps the internal state of type checking.  This information may be a result of complex computing but this process is not reported by compilers. For a typical type error shown in Fig. \ref{fig:ghc-error-example}, clearly the evidence for type error are gathered from are generated from the previous two declarations. These have to be re-discovered by the programmers again, using harder and less sound methods.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/ghc-error-example.png}
    \caption{
        }
    \label{fig:ghc-error-example}
\end{figure}

Deduction steps allow programmers to explore all the error locations one at a time (Fig. \ref{fig:deduction}). Steps are shown as a list of sequentially  numbered circular buttons (step buttons) and an explanation layer in the editor window. In the explanation layer, two highlighted error locations are outlined, and a line is drawn to connect these two locations to a human-readable text explanation of
their semantic connection. Programmers are free to activate any step. The active step is shown as the green button. When activating a step, some highlights switch color. The message in the explanation layer changes as well. A program in Fig. \ref{fig:deduction} generates a list of steps shown in Fig. \ref{fig:step-interface} left.

Programmers use their mouse and keyboard shortcuts to increment or decrement the step number or jump to any step. Typically, a programmer resolves  type errors by navigating through all the deduction steps and verifying whether each explanation aligns with the intended meaning of the program. Eventually, she will find a step that does not match her intention, and the type error can be fixed by modifying one of the two outlined error locations.

\begin{figure}[h]
    \centering
    \Description{Chameleon Step}
    \includegraphics[width=\linewidth]{images/step-interface.pdf}
    \caption{
Deduction steps if they are shown all at once (left). In practice, steps are shown one at a time. Programmers increment or decrement the step number using the step control bar (Fig. \ref{fig:anatomy}-F) or by directly clicking on a step button (Fig. \ref{fig:anatomy}-E). To increment or decrement the deduction step can be intuitively thought of as to move the position of the \textit{split point} where the blue and orange highlights divide (right).
        }
    \label{fig:step-interface}
\end{figure}

Internally, deduction steps are different ways to group the error locations into two different highlighted colors. If all the highlights are placed in a list-like data structure, each increment of the step numbers moves the end of one color and the start of the other forward (Fig. \ref{fig:step-interface} right).

\paragraph{Multiple Modes}

Nielson pointed out \cite{nielsen1994usability}  that the two most important issues in designing for usability are understanding the users' tasks and the differences in users. From analyzing how users use \chameleon{}, we realized that the ideal debugging interface should adapt to the specific programmer and programming task. It is common that a programmer wants the debugger to just "show the answer", and it is also possible a programmer wants to dive deeper into the problem domain and search for the optimal solution. To accommodate the need to customize the level of information density and granularity of control, \chameleon{} provides three modes: basic, balanced, and advanced. Programmers can switch between modes by clicking on the mode switching toggles (Fig. \ref{fig:anatomy}-I). The features accessible from different modes are shown in table~\ref{tab:chameleon-features}.

\begin{table}
    \centering
    \begin{tabular}{ l l  }
     Mode & Features \\ \hline
     Basic Mode & Type Compare Tool \\ \hline
     Balanced mode & Type Compare Tool \\
     & Candidate Expression Card \\  \hline
     Advanced mode & Type Compare Tool \\
     & Candidate Expression Card \\
     & Deduction Steps \\
    \end{tabular}
    \caption{\chameleon{} modes and features}
    \label {tab:chameleon-features}
\end{table}

### 2022-11-22 sections/conclusion.tex
Initial Overleaf Import

\section{Conclusion}

We present \chameleon{}, a type debugging tool for the Haskell programming language. \chameleon{} facilitates a constraint-based type inference back end that provides accurate and comprehensive type error reports. \chameleon{} also facilitates an interactive front end with three different display modes, enabling programmers to navigate deep type relations and triangulate conflicting statements in source code. Four participatory experiments were conducted to evaluate the tool's design with programmers with varied experiences. We found that, particularly for more complex tasks, \chameleon{} helped programmers to fix type errors more quickly than traditional text-based error messages.  Further, programmers actively using \chameleon{} interactive features are shown to be more efficient in fixing type errors than simply reading the type error output.

### 2022-11-22 sections/discussion.tex
Initial Overleaf Import

\section{Discussion}
This paper presents the interactive type debugging tool \chameleon{} and charts the evolution of its design across several iterations in response to user evaluation and feedback, as well as examining the effectiveness of the general approach compared to traditional static type error messages. We found that programmers using \chameleon{} type compare tool are able to debug errors faster than using traditional text-based error messages. This effect is shown more clearly when the task is not trivial. We found that programmers who actively use ChameleonIDE's interactive features (candidate expression cards and deduction steps) are more efficient in fixing type errors than simply reading the type error output. In this section, we will discuss a few interpretations of the results.

From the results of user studies 1 and 2, we observed that the choice of debugging tool had little effect on how fast programmers solve type errors. Conversely, when facing more realistic problems (longer source code, error locations more scattered), programmers are more effective using \chameleon{}. We believe one factor of this effectiveness of \chameleon{} is to reduce the amount of reading time. Jbara \cite{jbara2017programmers} showed that reading source code is generally the initial step of solving programming problems, and is done in several passes. Although traditional compiler error message tools initially show fewer locations, it is unreliable, meaning that programmers have to expand the reading span without clear guidance. In contrast, Chameleon shows more error locations initially. However, the completeness of error locations assures programmers which part of the source code can be safely skipped.

From the results of user studies 3 and 4, we found that programmers who use interactive tool fix type errors faster than the ones who passively read the error output. This difference is stronger in harder tasks. We speculate that one factor of this result is that  \chameleon{} helps to develop debugging plans. We observed that when working with \chameleon{}, programmers form different debugging tactics to attack the problem. Among the high interactivity participants in user study 4, some programmers cycle through deduction steps as a guide to reading source code; some programmers navigate to both ends of the deduction steps where types are normally grounded and concrete. In contrast, minimal interactive participants generally form more similar plans, including carefully reading the program text and manually annotating expressions based on their understanding of the program.

\begin{lstlisting}[language=Haskell, float, caption={
    In this simple program, \chameleon{} report 
    the error happened in \texttt{f} and \texttt{z}.
}, label={listing:fz}]
f z     
    | z == 3 = False
    | z == '4' = True
\end{lstlisting}

We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information. With the program showing in the listing \ref{listing:fz}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}. Although, in theory, these two statements are to the same effect, programmers are often required to compute the latter from the former or vice versa. And this computation may carry out multiple layers. Programmers have to remember all the intermediate types and their reasoning throughout such mental gymnastics. Assisted by candidate expression cards and deduction steps, this intermediate information is preserved on screen and can be retrieved anytime. This may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.

In our studies, \chameleon{} shows a few different designs for type error visualization and interaction. However, we are far from exhausting the search space. One challenge we notice in the current design is that programmers have to shift their focus between the editor on the left and the \chameleon{} debugging interface on the right. Using the hover popup window in most mainstream IDEs may reduce the context switching during debugging.

On a similar note, the current implementation of \chameleon{} requires non-trivial adaptation for editors such as VS Code and IntelliJ due to the overlay explanation layer. This type of error visualization is non-standard in mainstream editors. However, there may exist alternative representations of \chameleon{} error reporting using only the features available in major editors and IDEs. It is especially beneficial to represent \chameleon{} errors using a universal debugging middle layer such as language server protocol (LSP). This will allow \chameleon{}  to be adapted into various coding environments which support an LSP back end.

Another area for future study is to extend on the evaluation process. In the user studies 2, 3, and 4, we used harder tasks compared to the ones in user study 1. However, the difficulty level is nowhere as hard as a professional Haskell programmer may face in the production Haskell codebase. It would be interesting to see how \chameleon{}  faces type errors that span multiple files and packages. Another dimension of difficulty is higher-level extractions. Haskell programs get more tricky to the debugger when more abstractions are added to the codebase, such as Monads, Monad transformers, and Lenses. These topics were not included in our study to accommodate novice users. However, observing how expert Haskell programmers use \chameleon{} features to solve advanced problems will gain valuable insight into the true usability of the tool.

Another direction to expand the work is in data collection. The online study limited the amount of qualitative data we could gather from the users. Although we handed out debriefing surveys after each study, it is no substitute for more in-depth qualitative study methods. Inviting users for speak-aloud walkthroughs and interviews are viable methods to unveil deeper usability issues about \chameleon{}.

Our study focuses on the Haskell language for its popularity in the academic world. However, the low-quality of textual type errors is a problem not limited to Haskell. Modern statically typed languages more or less all share the same problem. We believe the underlying type reporting technique and algorithms can be generalized to other languages. It will be exciting to see how interactive debugging features perform in other paradigms, such as  imperative languages and incremental type systems.

Our design of \chameleon{}  emphasizes allowing programmers to update the type error based on their prior knowledge. It unfolds to answering three questions: which expression should be considered un-typeable? Between multiple possible types, which one should be considered intended? Among all the possible locations, which one should be considered the cause? \chameleon{} answers all three questions by leaving it to the programmers to decide. However, it is possible and likely that some programmers enjoy conclusive error messages instead of flexible ones provided by \chameleon{}. Using heuristic methods may achieve the best of both worlds. It may produce effective results by guessing the possible un-typable expression based on the position it appears in the syntax tree and project structure or guessing the intended type signature by comparing the number of supporting slices in the source code.

### 2022-11-22 sections/evaluation.tex
Initial Overleaf Import

\section{Evaluation}

Over 12 months, several studies were conducted to answer new research questions or rephrase existing ones. These studies are grouped in 3 phases, each tested a different version of \chameleon{} with human participants. The number of participants and their haskell experience is shown in the table below.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/participants-experience.png}
    \caption{}
    \label{fig:participant-experience}
\end{figure}

A summary of the timeline of the user studies, their major questions and conclusions is given in Figure~\ref{fig:timeline}.

\begin{figure}[h]
    \centering
    \Description{A timeline of \chameleon{} evaluations}
    \includegraphics[width=\linewidth]{images/timeline.pdf}
    \caption{The timeline of \chameleon{}  development and 
    evaluation.}
    \label{fig:timeline}
\end{figure}

\subsection{Experiment Design}
\subsubsection*{\textbf{Recruitment}}

Participant recruiting channel was the social news aggregation website Reddit. Specifically, our user study advertisements were posted in the Haskell programming language community "r/haskell" and a general programming language-focused community "r/programminglanguages". Recruiting from social media allowed us to access a more diverse demographic that better represent the true population of Haskell programmers. The participation is fully anonymized. The detailed ethical implications of these experiments are reviewed and approved by the IRB of the institution of the authors.

One consequence of our recruiting approach is it is harder for previous participants to enter an later study. To minimise the lurking variable of previous experience, we use new code challenge every study and conduct trial runs in every study to bring new participants up to speed.

\subsubsection*{\textbf{Experiment setting}}
The experiment took place remotely and unsupervised. Participants took the study online via web browser and at the physical venue of their choosing. All user studies use a web-based debugging environment developed by the authors. Conducting the studies online helped us avoid variation when performing tasks in unfamiliar places and using different setups. The downside is that to intervene when a user encounters bugs or usability issues mid-study is impossible. To ensure there is no major usability issue that could lower the quality of data collection, we conducted cognitive walkthroughs and sandbox pilots before running each study.

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/study-process.png}
    \caption{}
    \label{fig:study-process}
\end{figure}

\subsubsection*{\textbf{Training and group assignment}}
After consent, participants received interactive training on every function of the tool interface and interactive features. Participants were also shown the important functionalities of the tool interface in a summarized cheat sheet. Participants had access to the cheat sheet at all times during the study. Participants were given 4 trial runs (2 for each setting) before the data collection starts.

All the user studies used a within-subject design to evaluate the effectiveness of different tools or feature sets while counterbalancing the difference in programming proficiency between participants. In each user study, participants were required to complete a series of programming tasks (8 for user studies 1 - 3, 9 for user study 4). At each task, a participant receives a single Haskell file that contains one or more type errors. The participant was asked to make the code type check with the help of the tool given for this task.

% During each study, a participant received two (user studies 1 - 3) or three (user study 4) different debugging tools. The debugging tools were different systems (user studies 1 - 3) or different feature sets of the same system (user study 4). For each participant, the debugging tools alternative through consecutive tasks to ensure the fatigue level does not impair the rigorousness of data collection.

\subsubsection*{\textbf{Measurement}}
Time is measured from the start of each task to the first time the program is successfully type-checked and also passes all the functional tests. The data is automatically recorded by the online debugging environment. To not introduce a barrier to completing the study, every task can be skipped if the participant made three attempts or spent over 1 minute on the task.

After completing all the tasks (8 in user studies 1-3, 9 in user study 4), users are prompted to fill in a debriefing survey. The survey questions include their Haskell programming experience and feedback on the tools and feature sets participants used during the study.

We used an browser session recording tool~\cite{openreplay} to record the user study sessions. These recordings are used to find usability issues in the study and recognize general patterns. However, the recording technology does not provide high enough sampling rate for us to be confidently used for rigorous analysis.

\subsection{\chameleon{} User Studies}

\subsubsection{\textbf{Version 1}}  
\chameleon{} version 1 provides the rewrite of the Chameleon type inference engine and the type compare tool to show the alternative types and possible error locations in matching colors.

\begin{figure}[h]
    \centering
    \Description{Chameleon v1}
    \includegraphics[width=\linewidth]{images/chameleon-v1.pdf}
    \caption{
        \chameleon{} Version 1
        }
    \label{fig:chameleon-v1}
\end{figure}

Two user studies were conducted to test \chameleon{} version 1, to compare the effectiveness of solving type errors using \chameleon{} and GHC compiler error messages. We choose GHC compiler error messages as the baseline because it is the canonical tool for working with type error in Haskell. Although high-level tools like Haskell Language Server exist, they generally relay the GHC error messages verbatim.

Eight tasks were given in both studies. In the first study, the tasks were taken from a exercises from the exercises of the Haskell programming units. These tasks cover a variety of type errors. \tf{break down the type error classes.} However these tasks, as shown from the result and feedback of user study 1, were too simple to fully evaluate the two options. In the second study, the tasks are sourced from the top 20 Haskell topics on GitHub~\cite{githubHaskell}. The authors manually added type errors.

% The study investigates the effectiveness of \chameleon{} compared to the GHC compiler error messages. We chose GHC compiler error messages as it is the canonical tool for debugging type errors in Haskell. Although high-level tools like Haskell Language Server exist, they relay the GHC error message verbatim for type errors. Participants are asked to complete 8 tasks. The tool participants used during each task alternated between \chameleon{} and GHC. Task programs were sourced from HaskellWiki \cite{haskellwiki}. The author manually added type errors. The errors cover a range of common Haskell type errors, including abstract data types, wrong arity, control expressions (if and case), infinite types, and tuples. The lines of code (LoC) range from 7 to 17 (mean = 11, median=10.5).

% In total 39 participants finished the study. Among them, 12 participants have over five years of Haskell experience, and 5 participants have three or four years of Haskell experience. And 8 participants have one or two years of experience, and 5 participants used Haskell for under a year. The rest left the question unanswered.

\begin{figure}[h]
    \centering
    \Description{Time to complete by task in user study 1 with confidence interval}
    \includegraphics[width=\linewidth]{images/r1-data.pdf}
    \caption{Time to complete by task in user study 1 with confidence interval}
    \label{fig:r1-analysis}
\end{figure}

\begin{figure}[h]
    \centering
    \Description{Time to complete by task in user study 2 confidence interval}
    \includegraphics[width=\linewidth]{images/r2-data.pdf}
    \caption{Time to complete by task in user study 2 confidence interval}
    \label{fig:r2-analysis}
\end{figure}

\subsubsection*{\textbf {Results}}

From the data collected during user study 1 (\ref{fig:r1-analysis}), we could not identify which group performed better (p-value =0.2041, Wilcoxin signed test). For the trivial challenges we set for users, the individual differences are generally more significant than differences between treatments. 
The most common feedback (7 out of 35) from the user study 1 was that the tasks were too trivial to invite meaningful evaluation. One participant put it, "Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice".

One interesting exception is task 8, where the \chameleon{} group outperformed the GHC group more consistently and with a larger effect size (p-value = 0.002456, Wilcoxin signed test). The discrepancy is thought to be related to the nature of Task 8:
\begin{itemize}
    \item {It has a longer source file than other tasks (only shorter than task 6, but task
    6 contains two independent type errors while task 8 is one connected task
    error);}
    \item {It is more complex than others (involves abstract data types and function application); and
    }
    \item {
        GHC struggles to produce a relevant error message for this type of error.
    }
  \end{itemize}
It can be further shown by examining the process some participants took to solve the problem.

% \begin{figure}[H]
%     \centering
%     \Description{A screenshot from user study 1}
%     \includegraphics[width=\textwidth]{round1-screenshot.pdf}
%     \caption{
% One participant working with \chameleon{}  managed to identify the most probable location of the error (the application of `fromMaybe'` on line 11) by hovering over the two alternative types. The participant then quickly realized that the first argument `0` (highlighted in orange) is inconsistent with the definition where it is case matched to a `Nothing` value (highlighted in purple). After two retries with short hesitation the participant found the correct fix (by reversing the order of `0` and `n`).
%     }
%     \label{fig:r1-task8}
% \end{figure}

% \begin{figure}[htb]
%     \centering
%     \Description{A screenshot from user study 1}
%     \includegraphics[width=\textwidth]{round1-screenshot-ghc.pdf}
%     \caption{
%         In task 8, participants using GHC took longer pauses at this task before committing to further investigation, either carefully reading the whole text or figuring out the meaning of the error message. The GHC error message is unfortunately unhelpful for this task.
%     }
%     \label{fig:task8-ghc}
% \end{figure}

In comparison, the \chameleon{} group solved the type error faster than the GHC group in almost all tasks (p =0.00742, Wilcoxin signed test) in study 2 (figure \ref{fig:r2-analysis}), barring task 1. The authors are unclear about the cause of low performance from \chameleon{} in task 1. It is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity. From the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program.

In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}. "Its most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant. "I think \chameleon{}  does a much better job than GHC's error messages. I like that it shows the sources for the type judgements. This makes it quite easy to figure out how to rectify errors." reported another participant.

\subsubsection{\textbf{Version 2}}  \label{sub:us4}

Version 2 is latest version of \chameleon{}. Not only the added features the deduction steps and candidate expressions are We designed an exploratory study to get better understanding of how users learn and use the multi-mode design of \chameleon{} version 3. During the study the initial mode of each task alternative through the three different modes, and repeat three cycles in nine tasks. The order of the three modes in each cycle is counterbalanced among all participants. However, participants are free to switch to other modes at any time.

\subsubsection*{\textbf{Results}}

The study of version 3 \chameleon{} is exploratory, in that we were hoping to let programmers discover their own way of using the tool. In post-hoc analysis of the collected log data, we were able to extrapolate some interesting patterns of how the tool was used. We share a few of our observations in this section.

The most striking feature of the data is users' tendency of using tools differ wildly. That is to say, some users used the feature extensively however some users completed the tasks without actively explore the given information. Based on this difference we divided the users into three groups.

\begin{tabularx}{\linewidth}{ 
  | >{\raggedright\arraybackslash}X 
  | >{\raggedright\arraybackslash}X  | }

    \hline
        Interaction level & Description \\ \hline
        Minimal Interaction & Users completed the tasks by making changes in source code, type checking, and reading error messages. \\ \hline
        Low Interaction & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
        High Interaction & Users did everything from the Common feature group but used features specific to Balanced and Advanced mode, such as activating steps and expression cards. \\ \hline

\end{tabularx}

As shown in  figure \ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants. Participants with higher interaction levels generally performed better in the tasks, and the lowest level was worse (one-way ANOVA since we have three samples, p = 0.0368). The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated. As discussed in user studies 1 and 2, it is likely related to the nature of these tasks. Tasks 4 and 6 are shorter tasks in the study. The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code). These may be useful because following the natural reading order will allow participants to skip lines on the bottom, and thus the advantage of \chameleon{} is less important. On the other hand, task 9 is the longest task of all. It also involves harder problems such as mutually recursive type definitions.

\begin{figure}[h]
    \centering
    \Description{
    Time to complete by task in user study 2 with confidence interval}
    \includegraphics[width=\linewidth]{images/r4-analysis.pdf}
    \caption{Time to complete by task in user study 4 with confidence interval}
    \label{fig:r4-analysis}
\end{figure}

\begin{figure}[h]
    \centering
    \Description{a screenshot from user study for version 2}
    \includegraphics[width=\linewidth]{images/r4-task9.png}
    \caption{
High interaction users were able to locate the exact line for a potential fix after navigating to deduction step 6 or 7, which directly revealed the true cause of the type error. After this, 12 out of 15 high interaction users provided the most ideal fix in the first try. Low interaction group  takes longer time to examine the problem, provided wrong fixes in the initial trials. One participant fixed a minor run-time issue (a problematic pattern matching of `[attributes]` on line 18) however failed to identify the culprit of the type error.
    }
    \label{fig:r4-task9}
\end{figure}

Another observation is when using the mode switching feature of \chameleon{}, programmers generally switching form a less informative display to a more informative one (Fig. \ref{fig:r4-mode-switching}).

\begin{figure}[h]
    \centering
    \includegraphics[width=\linewidth]{images/r4-mode-switching.png}
    \caption{
        \textbf{Row Number of tasks }

    }
    \label{fig:r4-mode-switching}
\end{figure}

### 2022-11-22 sections/introduction.tex
Initial Overleaf Import

Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer2012static} and codebase maintainability~\cite{kleinschmager2012static}. In recent years, the programming community has seen a trend of migrating away from dynamically typed codebases due to the lack of maintainability and refactoring safety when systems reach large scale and complexity~\cite{chatley2019next700}. These migration methods include the introduction of gradual type annotations (e.g., JavaScript to TypeScript or Python with mypy) or transition to a modern strongly-typed language (Scala, Rust, Haskell, PureScript, etc.). However, with dynamically typed languages (e.g., JavaScript, PHP, and Python) still being most peoples' formative experience with programming and with such languages now firmly entrenched in education~\cite{so2022survey}, this transition may pose difficulties for programmers who lack experience writing type-safe programs.

Programmers find challenges adopting typed languages in their practice. They tend to be verbose, hard to learn, and generate bad error messages. 
One obstacle programmers face when using a statically typed language is understanding and resolving type errors~\cite{marceau2011measuring, tirronen2015understanding}, especially when dealing with modern type features such as polymorphic types and implicit typing. Studies show most statically typed languages produce unhelpful and even misleading type errors~\cite{yang2000improved, some more proof}. Symptoms of bad error messages include cryptic language exposing internal constructs of the compiler and incomplete or wrong error locations. These usability issues poses challenges to learners and teachers of the languages. impede the wide adoption of statically typed languages, especially for programmers who are now accustomed to the mindset of writing un-typed programs.

This paper introduces \chameleon{}, an interactive debugging tool to visualize critical context of a type error (where it happened or could have happened, and which chunks of code caused it) and allow programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve that ambiguity. Most noticeably, features (Section \ref{sub:type-compare}) provided by \chameleon{} are the type compare tool, the deduction step feature , and the candidate expression card feature.  In addition, we integrated all the features into a unified debugging interface and provided functionality for programmers to enable and disable the features based on their preferences and debugging needs. \chameleon{} is open-source and is available in both web~\cite{chameleon} and desktop versions. The current implementation of \chameleon{} targets the Haskell 2010 language standard. However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques. Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages. Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak2007history}, and now it can be found in most popular languages such as C\#~\cite{csharpgenerics}, Java~\cite{javagenerics}, and TypeScript~\cite{tsgenerics}.

% \chameleon{} uses underlying type inference algorithms adapted from the command-line tool, Chameleon, developed in 2005 \cite{chameleon}. As described in Section \ref{sec:typeinferenceengine}, Chameleon computes the Minimum Unsatisfiable Subset (MUS) \tf{Explain what mus is} of type constraints in order to identify a set of code locations where multiple types can be inferred. Compared to standard compiler messages, which arbitrarily report the first place a type conflict arises, Chameleon error messages give a lot more context to the programmer to help them correctly resolve the conflict to match the program's intent. However, the Chameleon system was never tested with users.  Therefore, the first contribution of this paper is to address the research question: \textit{Is a minimally interactive version of Chameleon's multipart type-conflict display more effective in supporting type error debugging than traditional error messages? } (Sections~\ref{sub:us1} and \ref{sub:us2}).

% After affirming that the information Chameleon provides is beneficial for more complex type errors, we ask \textit{Do programmers benefit from the interactive exploration of type error locations?} (Sections~\ref{sub:us3} and \ref{sub:us4}). We then explore \textit{How do programmers use the interactive exploration features to solve type errors?}  (Section~\ref{sub:us4}).

This paper makes the following contributions:
\begin{itemize}
\item { We provide the design and implementation of the \chameleon{}, to visualize crucial context of a type error and allow programmers to explore and verify the error locations in small chunks interactively.  }
\item {
    We report the result of four experiments designed to evaluate \chameleon{}.}
\end{itemize}

Our experiments showed that programmers using \chameleon{} fix type errors faster than with traditional text-based error messages. This difference is more significant when solving harder tasks. Further, programmers who actively use \chameleon{} interactive features fix type errors faster than simply reading the type error output.

### 2022-11-22 sections/related-work.tex
Initial Overleaf Import

\section{Related Work}

Researchers have long realized that incomprehensible type errors result as a consequence of standard approaches to the type inference process. Haack and Wells~\cite{haack2004type} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors. To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' Researchers have proposed solutions to improve the type inference process. Type error slicing~\cite{haack2004type} is a technique that finds locations that are complete and minimal for the type error. Internally labeled constraints and MUS manipulation are used to generate these slices. The language supported in Haack's work was a subset of standard ML. The original Chameleon~\cite{stuckey2003interactive} used a similar technique but extended it to support advanced type-level features (type classes and functionally dependent types). The project also introduced the ability to query type information through a command line-style interface (Fig.~\ref{fig:original-chameleon}). Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies.

\begin{figure}[h]
    \centering
    \Description{Chameleon interactive debugging and error message}
    \includegraphics[width=\linewidth]{images/original-chameleon.pdf}
    \caption{
Originally, Chameleon was a command-line tool that can list all the potential causes of a type error and display the two branches in two forms of highlight (right), achieved with ASCII colors. In addition, it allows a few commands (debug and explain) in the debugging shell to help debug type errors (left).}
    \label{fig:original-chameleon}
\end{figure}

% The ChameleonIDE error reporting The error message of ChameleonIDE closely
% follows the argumentation structure outlined by Titus. Basic mode represent a
% simple structure argument, where the claim "The expression e can have two
% conflicting types" and the possible type 1 and two serves as grounds. When
% inspecting each possible types, the type judgements are treated as claim, and it
% is supported by the grounds: "Inferred from the orange/blue highlights on the
% left side". On the other hand, the similar type error message in GHC "Couldn't
% match expect type T1 with actual type T2" is "a ground masquerading as a clam",
% Barik commented. The balanced mode and advanced mode serve as a way to elaborate
% arguments from simple arguments into extended arguments.

Another related topic is type-explanation~\cite{yang1999explaining, jun2002explaining}. Yang~\cite{jun2002explaining} showed an alternative type inference system capable of producing a human-like text explanation for why expressions are assigned certain types. A good explanation is drawn from surveying how human experts explain types. The resulting algorithm $\mathcal{H}$, generates a succinct explanation of the type inference steps to avoid using internal constructs (such as type variables). The explanation system has the advantage of acting like a human expert. However, when presented as text-based output, explanation systems have the potential to become verbose when types are complex or variable names are long. In \chameleon{}, we attempt to address this problem by showing one step of explanation at a time and referring to variables instead of spelling out the full name.

% Another method to improve type error message is search based type
% checking~\cite{lerner2007searchingtypeerror}. This method incrementally modifies
% an ever smaller part of the abstract syntax tree and replacing it with a
% wildcard expression and querying the type checker to verify if the program still
% type-checks. Lerner's study showed how two implementations could provide
% high-quality error messages and to suggest accurate fixes. While this system is
% great for suggesting solutions, is it limited in its ability to explain the type
% errors due to the fact that the system is unaware of any inner workings of the
% type systems.

% Some studies \cite{seidel2017learningtoblame} proposed a heuristic approach to
% improve type error messages. These studies use part of type information in the
% program to query external databases. Based on the external databases'
% capability, some projects can interactively fix the type error on the fly with
% reasonable accuracy. \pjs{Add jusdgment/comparison sentence!}

Debugging using a GUI Interactive Development Environment has been the standard practice for a very long time. Compared to a command line-based interface, a graphic user interface provides programming tools with the ability to show information hierarchy, provide immediate feedback to changes, and display complex visualizations. RxFiddle~\cite{banken2018debugging} is a tool to visualize data flow and value changes in a functional reactive program. Whyline~\cite{ko2009finding} is a Java debugging system that allows a user to ask questions like "why does variable X have value Y". It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error. Both  systems target programming runtime systems instead of types systems and static analysis. In fact, support for type systems has been lacking in both academic and industry software engineering tool design. The most common approach for supporting type systems we see in current programming editors and IDEs is to display type annotations when hovering on an expression and show a squiggly line to indicate the type error location. In \chameleon{}, colours and geometric shapes are used to create visual clarity of different components of a type error; interactivity is used to show conditional information and to change the granularity of details.

### 2022-11-22 sections/walkthrough.tex
Initial Overleaf Import

\section{Walkthrough} \label{sec:walkthrough}
In this section, we introduce \chameleon{} by walking through examples of its
use. The examples are given from the perspective of a hypothetical Haskell 
programmer Maxine.

\subsection{Basic mode} \label{sub:basic}
Maxine writes a function to calculate the sum of a list of
numbers, but \chameleon{} shows there is a type error (Fig. \ref{fig:basic-mode-1}). 
After reading error reports, Maxine realized that the error possibly revolves 
around the expression \texttt{xs}.

\begin{figure}[h]
        \centering
        \Description{Maxine using basic mode}
        \includegraphics[width=\linewidth]{images/basic-mode-1.pdf}
        \caption{
            Maxine's code to calculate the sum of a list of integers;
            \chameleon{} reports an error on the expression \texttt{xs}.
            }
            \label{fig:basic-mode-1}
\end{figure}

\begin{figure}[h]
        \centering
        \Description{A screenshot of using basic mode}
        \includegraphics[width=\linewidth]{images/basic-mode-2.pdf}
        \caption{
       Hovering on possible type 1 will limit the highlights 
       in the editor to only the ones (in orange) that support \texttt{x::[a]}.
        }
        \label{fig:basic-mode-2}
\end{figure}

\begin{figure}
        \centering
        \Description{A screenshot of using basic mode}
        \includegraphics[width=\linewidth]{images/basic-mode-3.pdf}
        \caption{
            Hovering on possible type 2 (Fig. \ref{fig:basic-mode-2}-3) will limit the highlights 
            in the editor to only the ones (in blue) that support \texttt{x::Int}.
        }
        \label{fig:basic-mode-3}
\end{figure}

After scanning the type error report, Maxine realizes that \texttt{xs} can be
either \texttt{[a]} type (Fig. \ref{fig:basic-mode-2}) or \texttt{Int} type (Fig. \ref{fig:basic-mode-2}). By matching the color in the
conflicting type panel (Fig. \ref{fig:anatomy}-G) and the highlighted error locations 
Maxine knows that the \texttt{[a]} type results from the pattern matching of the
\texttt{:} operator, and the \texttt{Int} type results from using \texttt{+} to
 add two expressions.

At this point, Maxine knows the possible type 1 aligns with her understanding of
the program, and therefore, the error locations with blue highlights must be
erroneous. After examining the program, it is obvious that Maxine missed
applying the \texttt{sum} function recursively at the right-hand side of the
addition.

\subsection{Balanced mode} \label{sub:balanced}

Maxine writes additional code to add only the even numbers in a list of integers.
In this implementation, Maxine reuses the \texttt{sum} function she wrote
earlier. After saving the file, \chameleon{} shows that there is a type error in the
expression \texttt{sum} (Fig. \ref{fig:balance-mode-1}). However, this is not 
helpful because Maxine knows that the implementation of  \texttt{sum} is correct 
because it was verified in the previous task. Switching to
balanced mode, Maxine noticed that \chameleon{} shows two cards: \texttt{sum}
and \texttt{evens}.

\begin{figure}
        \centering
        \Description{A screenshot of using balanced mode}
        \includegraphics[width=\linewidth]images/{balanced-mode-1.pdf}
        \caption{
            Maxine's code to calculate only the sum 
            of even numbers. \chameleon{} reports 
            an error with two candidate expressions.
        }
        \label{fig:balance-mode-1}
\end{figure}

\begin{figure}
   \centering
        \Description{A screenshot of using balanced mode}
        \includegraphics[width=\linewidth]{images/balanced-mode-2.pdf}
        \caption{
            Clicking on the \texttt{evens} card (5) results in the changes in the
            conflicting types panel to show the possible types for \texttt{evens},
            and the changes highlights color to reflect the assumption that the
            definition of \texttt{evens} is the cause of the error.  
        }
        \label{fig:balance-mode-2}
\end{figure}

Maxine therefore clicks on the \texttt{evens} card and \chameleon{} shows two
possible types for the expression \texttt{[Int]} and \texttt{[Int] -> [Int]}
(Fig. \ref{fig:balance-mode-2}). Knowing the expression \texttt{evens} holds
a temporary list of even integers (hence it is of \texttt{[Int]} types), Maxine
knows the Possible type 2 is unintended, and blue in-text highlights must
contain the cause. It does not take long until Maxine realizes that the list 
\texttt{evens} is not supplied to the \texttt{filter} function.

\subsection{Advanced mode}  \label{sub:advanced}

For the task shown in section \ref{sub:balanced}, if Maxine is not satisfied by
the options provided by \chameleon{}, by switching to advanced
mode, Maxine has access to the debugging steps and the explanation layer.

\begin{figure}
        \centering
        \includegraphics[width=\linewidth]{images/advanced-mode-1.pdf}
        \Description{A screenshot of using advanced mode}
        \caption{
            Maxine's code to calculate only the sum 
            of even numbers in advanced mode. 
            The current step is step 5, \chameleon{} 
            explains that the two appearances of expression 
            \texttt{evens} should have the same type.
        }
        \label{fig:advanced-mode-step5}
\end{figure}

\begin{figure}
        \centering
        \includegraphics[width=\linewidth]{images/advanced-mode-2.pdf}
        \Description{A screenshot of using advanced mode}
        \caption{
            In step 6, \chameleon{} 
            explains that \texttt{evens} is defined as
            the expression \texttt{filter isEven}. The left-hand side
            and right-hand side should have the same type.
        }
        \label{fig:advanced-mode-step6}
\end{figure}

\begin{figure}
        \centering
        \includegraphics[width=\linewidth]{images/advanced-mode-3.pdf}
        \Description{A screenshot of using advanced mode}
        \caption{
            In step 7, \chameleon{} 
            explains that \texttt{filter} is applied to 
            the function \texttt{isEven}. Assisting by 
            the type of \texttt{filter} in the 
            Relevant Type Information panel on the bottom
            right, Maxine can find the type error that 
            \texttt{filter} expects two arguments but receives one.
        }
        \label{fig:advanced-mode-step7}
\end{figure}

First, Maxine clicks on step 5 (Fig. \ref{fig:advanced-mode-step5}) and verifies
that the two occurrences of \texttt{evens} are supposed to be identical, and the
second usage dictates that \texttt{evens} is a list of integers. Second, she
clicks on step 6 (Fig. \ref{fig:advanced-mode-step6}) and verifies that
\texttt{evens} should be the same type as the declaration on the right-hand
side.

Lastly, Maxine clicks on step 7 (Fig. \ref{fig:advanced-mode-step7}), and
it shows that the \texttt{filter} function is applied to one argument
\texttt{isEven}. By consulting the relevant type information, Maxine identifies
that \texttt{filter} is expecting two arguments while only one is provided.

### 2022-11-23 sections/evaluation.tex
Fix latex errors

@@
     \centering
-    \Description{A timeline of \chameleon{} evaluations}
     \includegraphics[width=\linewidth]{images/timeline.pdf}
@@
     \centering
-    \Description{Chameleon v1}
     \includegraphics[width=\linewidth]{images/chameleon-v1.pdf}
@@
     \centering
-    \Description{Time to complete by task in user study 1 with confidence interval}
     \includegraphics[width=\linewidth]{images/r1-data.pdf}
@@
     \centering
-    \Description{Time to complete by task in user study 2 confidence interval}
     \includegraphics[width=\linewidth]{images/r2-data.pdf}
@@
     \centering
-    \Description{
-    Time to complete by task in user study 2 with confidence interval}
     \includegraphics[width=\linewidth]{images/r4-analysis.pdf}
@@
     \centering
-    \Description{a screenshot from user study for version 2}
     \includegraphics[width=\linewidth]{images/r4-task9.png}
@@
     \caption{
-        \textbf{Row Number of tasks }
-
+        The rows represent the finishing mode, columns starting modes.
     }

### 2022-11-24 sections/chameleon.tex
Fix latex errors
Update images
Assets and code listing

@@
     \centering
-    \Description{Comparing Chameleon with \chameleon{}}
     \includegraphics[width=\linewidth]{images/compare-to-original.pdf}
@@
 
-\begin{lstlisting} [
-    language=Haskell, caption={
-    A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
-}, label={listing:ifelse}
-]
+% \begin{lstlisting} [
+%     language=Haskell, caption={
+%     A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.
+% }, label={listing:ifelse}
+% ]
+% if a then b else c
+% a = "True"
+% \end{lstlisting}
+
+\begin{minted}{haskell}
 if a then b else c
 a = "True"
-\end{lstlisting}
+\end{minted}
+
 For instance, the program in the listing \ref{listing:ifelse} \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition).
@@
     \centering
-    \Description{An anatomy of \chameleon{} interface}
-    \includegraphics[width=\linewidth]{images/atonomy.png}
+    \includegraphics[width=\linewidth]{images/atonomy.pdf}
     \caption{
@@
     \centering
-    \Description{Using \chameleon{} Type Compare Tool}
     \includegraphics[width=\linewidth]{images/intro-compare.pdf}
@@
     \centering
-    \begin{subfigure}[b]{\linewidth}
-        \centering
-        \Description{Using \chameleon{} Candidate Expression Cards}
-        \includegraphics[width=\linewidth]{images/intro-expression.pdf}
+    \includegraphics[width=\linewidth]{images/intro-candidate.pdf}
     \caption{
@@
     \label{fig:expression}
-    \end{subfigure}
-    \hfill
-    \begin{subfigure}[b]{\linewidth}
+\end{figure}
+
+\begin{figure}
     \centering
-        \Description{Using \chameleon{} Deduction Steps}
-        \includegraphics[width=\linewidth]{images/intro-deductionstep.pdf}
+    \includegraphics[width=\linewidth]{images/intro-deduction.pdf}
     \caption{
@@
     \label{fig:deduction}
-    \end{subfigure}
-    \hfill
-    \caption{ Candidate Expression Cards and Deduction Steps}
 \end{figure}
@@
     \centering
-    \Description{Showing an type error with a single candidate}
     \includegraphics[width=\linewidth]{images/single-error-location-example.png}
@@
     \centering
-    \Description{Chameleon Step}
     \includegraphics[width=\linewidth]{images/step-interface.pdf}

### 2022-11-28 data/v1b-analysis.rmd
Update images

---
title: "R Notebook"
output: html_notebook
---

```{r message=FALSE, warning=FALSE}

library(tidyverse)
library(lubridate)
events <- read_csv("v1b-data.csv") %>%
  mutate(time = as_datetime(time), task = as.integer(task)) %>%
  arrange(time)
```

```{r message=FALSE}
event_names <- events %>% distinct(event_name)

all_participants <- events %>%
  distinct(participant_id) %>%
  .$participant_id

completed <- events %>%
  filter(task == 8, event_name %in% c("complete", "skip")) %>%
  distinct(participant_id) %>%
  filter(! participant_id %in% c()) %>% 
  .$participant_id

starts <- events %>%
  filter(participant_id %in% completed) %>%
  filter(event_name == "start") %>%
  group_by(participant_id, task) %>%
  summarise(start_time = first(time))

ends <- events %>% 
  filter(participant_id %in% completed) %>%
  filter(event_name %in% c("complete", "skip")) %>%
  group_by(participant_id, task) %>%
  summarise(finish_time = last(time))

durations <- events %>%
  filter(participant_id %in% completed) %>%
  left_join(starts, by=c("participant_id", 'task')) %>%
  left_join(ends, by=c("participant_id", 'task')) %>%
  mutate(duration = as.numeric(finish_time - start_time))

durations |> group_by(task) |> summarise(
  avg_time=mean(duration),
  med_time= median(duration),
  min_time = min(duration),
  max_time=max(duration),
  sd_time = sd(duration))

attemps <- events %>%
  filter(participant_id %in% completed) %>% 
  filter(event_name == "attempts")

```

```{r}

err_left <- function (n, m, s) {
  return (m - qt(0.975,df=n-1)*s/sqrt(n))
}

err_right <- function (n, m, s) {
  return (m + qt(0.975,df=n-1)*s/sqrt(n))
}

errors <- durations |>
  group_by(task, treatment) %>%
  summarise(
    l = err_left(n(), mean(duration), sd(duration)),
    m = mean(duration),
    u = err_right(n(), mean(duration), sd(duration))
  )

durations |>
  ggplot() +
  geom_bar(
    durations,
    mapping = aes(x = factor(task),
                  y = duration,
                  fill = treatment),
    position = "dodge",
    stat = "summary",
    fun = "mean"
  ) +
  geom_errorbar(
    data = errors,
    mapping = aes(
      x = factor(task),
      y = m,
      ymin = l,
      ymax = u,
      group = treatment
    ),
    position = position_dodge(0.9),
    alpha=0.5,
    width=0.6
  ) +
  labs(x = "Tasks", y = "Time (Seconds)", fill="Debugging Tool:", title="User study 2") + 
  scale_x_discrete(expand = c(0,0))  +
  scale_y_continuous(expand = c(0,0)) + 
  scale_fill_manual(
    values=c("#8fbfd5", "#b14863"),
    labels=\(inputs) map_chr(inputs, ~ ifelse(. == "Chameleon", "ChameleonIDE", "GHC"))
    ) +
  theme_bw() +
  theme(
    panel.grid.major = element_blank(), 
    panel.grid.minor = element_blank(), 
    panel.border = element_blank(),
    axis.line.y = element_line(colour = "black"),
    axis.ticks.x = element_blank(),
    legend.position = "bottom",
    plot.title = element_text(size = 15),
    axis.title = element_text(size = 14),
    text = element_text(size = 14)
    )

ggsave("r2-data.pdf", device = "pdf", plot = last_plot(), units ="mm", width = 210)

```

```{r}
d = durations |>
  select(participant_id, treatment, duration) |>
  group_by(participant_id, treatment) |> 
  summarise(duration=mean(duration)) |>
  pivot_wider(names_from = treatment, values_from = duration)

r <- wilcox.test(d$Chameleon, d$Ghc, paired=TRUE)
r
```

```{r}
one.way <- aov(duration ~ treatment, durations)
summary(one.way)
```

```{r}
events |> 
  group_by(participant_id) |>
  summarise(runtime = max(time) - min(time)) |>
  summarise(
    mean_runtime = mean(runtime),
    median_runtime = median(runtime)
    )
```

```{r}
events |>
  ggplot(mapping=aes(x=time)) +
  geom_density()

```

### 2022-11-28 sections/evaluation.tex
Update images
Assets and code listing

@@
 The number of participants and their haskell experience is shown in the table below.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\linewidth]{images/participants-experience.png}
-    \caption{}
-    \label{fig:participant-experience}
-\end{figure}
 
@@
 To ensure there is no major usability issue that could lower the quality of data collection, we conducted cognitive walkthroughs and sandbox pilots before running each study.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\linewidth]{images/study-process.png}
-    \caption{}
-    \label{fig:study-process}
-\end{figure}
 
@@
 
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\linewidth]{images/procedure-1.pdf}
+    \caption{}
+    \label{fig:study-process}
+\end{figure}
+
 % The study investigates the effectiveness of \chameleon{} compared to the GHC compiler error messages.
@@
     \centering
-    \includegraphics[width=\linewidth]{images/r1-data.pdf}
-    \caption{Time to complete by task in user study 1 with confidence interval}
+    \includegraphics[width=\linewidth]{images/user-study-1a.pdf}
+    \caption{Time to complete by task in user study 1a with confidence interval}
     \label{fig:r1-analysis}
@@
     \centering
-    \includegraphics[width=\linewidth]{images/r2-data.pdf}
-    \caption{Time to complete by task in user study 2 confidence interval}
+    \includegraphics[width=\linewidth]{images/user-study-1b.pdf}
+    \caption{Time to complete by task in user study 1b confidence interval}
     \label{fig:r2-analysis}
@@
 
+\begin{figure}[h]
+    \centering
+    \includegraphics[width=\linewidth]{images/procedure-2.pdf}
+    \caption{}
+    \label{fig:study-process}
+\end{figure}
+
 \subsubsection*{\textbf{Results}}
 
-The study of version 3 \chameleon{} is exploratory, in that we were hoping to let programmers discover their own way of using the tool.
+The study of \chameleon{} 2 is exploratory, in that we were hoping to let programmers discover their own way of using the tool.
 In post-hoc analysis of the collected log data, we were able to extrapolate some interesting patterns of how the tool was used.
@@
     \centering
-    \includegraphics[width=\linewidth]{images/r4-analysis.pdf}
-    \caption{Time to complete by task in user study 4 with confidence interval}
+    \includegraphics[width=\linewidth]{images/user-study-2.pdf}
+    \caption{Time to complete by task in user study 2 with 95\% confidence interval}
     \label{fig:r4-analysis}
@@
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\linewidth]{images/r4-task9.png}
-    \caption{
-High interaction users were able to locate the exact line for a potential fix after navigating to deduction step 6 or 7, which directly revealed the true cause of the type error.
+% \begin{figure}[h]
+%     \centering
+%     \includegraphics[width=\linewidth]{images/r4-task9.png}
+%     \caption{
+% High interaction users were able to locate the exact line for a potential fix after navigating to deduction step 6 or 7, which directly revealed the true cause of the type error.
 After this, 12 out of 15 high interaction users provided the most ideal fix in the first try.
@@
 One participant fixed a minor run-time issue (a problematic pattern matching of `[attributes]` on line 18) however failed to identify the culprit of the type error.
-    }
-    \label{fig:r4-task9}
-\end{figure}
+%     }
+%     \label{fig:r4-task9}
+% \end{figure}
 
@@
     \centering
-    \includegraphics[width=\linewidth]{images/r4-mode-switching.png}
+    \includegraphics[width=\linewidth]{images/mode-switching.pdf}
     \caption{

### 2022-11-28 sections/related-work.tex
Fix latex errors
Update images

@@
 
-\begin{figure}[h]
-    \centering
-    \Description{Chameleon interactive debugging and error message}
-    \includegraphics[width=\linewidth]{images/original-chameleon.pdf}
-    \caption{
-Originally, Chameleon was a command-line tool that can list all the potential causes of a type error and display the two branches in two forms of highlight (right), achieved with ASCII colors.
+% \begin{figure}[h]
+%     \centering
+%     \includegraphics[width=\linewidth]{images/original-chameleon.pdf}
+%     \caption{
+% Originally, Chameleon was a command-line tool that can list all the potential causes of a type error and display the two branches in two forms of highlight (right), achieved with ASCII colors.
 In addition, it allows a few commands (debug and explain) in the debugging shell to help debug type errors (left).}
-    \label{fig:original-chameleon}
-\end{figure}
+%     \label{fig:original-chameleon}
+% \end{figure}
 

### 2022-11-29 sections/related-work.tex
Wide images

@@
 Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies.
+
+Another direction to approach type error debugging is suggesting fixes.
+Sheng Chen's works [1,2] on counter factual typing is able to suggesting changes to make the ill-typed program typecheck.
+While this is promising ground, the work is not evaluated with human participants.
+\cite{lerner2007searchingtypeerror} used an approach by repeatedly modifying the abstract syntax tree and querying the compiler/type checker to know whether the program still well-typed.
+This approach allows the algorithm to produce suggestions the syntax changes that could fix the type error.
+Benefiting from not needing a type system impelemntation and being language agnositic, the underlying idea was implemented and evaluated on 10 CS students.
+The system can be successfully adopted in not suitable for teaching and learning purpose, as it is not able to explain why the type error happened and why fixes are suggested.
+The lack of explanability is because the information was witheld by the type checker and the inference algorthm does not know (need to know) how types are computed.
+
+One related approach is CEME (compiler error message enhancement).
+Prather showed that ECEM casts a positive result in understanding the type error through a series of mixed methods study.
+Decafe is a tool that is able to rephase the compier error message into an enhanced version.
+A study of over 200 CS1 students showed that CEME reduce overall errors in their assignments.
+Berik proposed a framework of constructing compiler error message based on the augumentation theory.
+In this proposal,  Berik argued that error messages  follow a simple argumentation layout or an exntended argumentation layout will result in a human friendly errors.
+These research shows the significance of improving the language in a compiler error message.
+Most principles and suggestions are followed in \chameleon{} in concustructing error statements.
+However they were not targeting type errors but compiler error (some even include runtime error) in general.
+The nuances of type errors, such as alternative typings were not taken into account.
 
@@
 
-Another related topic is type-explanation~\cite{yang1999explaining, jun2002explaining}.
+% Another related topic is type-explanation~\cite{yang1999explaining, jun2002explaining}.
 Yang~\cite{jun2002explaining} showed an alternative type inference system capable of producing a human-like text explanation for why expressions are assigned certain types.
@@
 
-% Another method to improve type error message is search based type
-% checking~\cite{lerner2007searchingtypeerror}. This method incrementally modifies
-% an ever smaller part of the abstract syntax tree and replacing it with a
-% wildcard expression and querying the type checker to verify if the program still
-% type-checks. Lerner's study showed how two implementations could provide
-% high-quality error messages and to suggest accurate fixes. While this system is
-% great for suggesting solutions, is it limited in its ability to explain the type
-% errors due to the fact that the system is unaware of any inner workings of the
-% type systems.
-
-% Some studies \cite{seidel2017learningtoblame} proposed a heuristic approach to
-% improve type error messages. These studies use part of type information in the
-% program to query external databases. Based on the external databases'
-% capability, some projects can interactively fix the type error on the fly with
-% reasonable accuracy. \pjs{Add jusdgment/comparison sentence!}
-
 Debugging using a GUI Interactive Development Environment has been the standard practice for a very long time.
 Compared to a command line-based interface, a graphic user interface provides programming tools with the ability to show information hierarchy, provide immediate feedback to changes, and display complex visualizations.
-RxFiddle~\cite{banken2018debugging} is a tool to visualize data flow and value changes in a functional reactive program.
+Provide opportunities to run consecutive command to query the information about code or executing command based on the existing code.
+HazelTutor is a interactive filling type holes by suggesting template expressions (strategies as called by the authors).
+It also provide a cursor based type inspector that allows programmers to query the type of parts of the program.
 Whyline~\cite{ko2009finding} is a Java debugging system that allows a user to ask questions like "why does variable X have value Y".
 It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error.
-Both  systems target programming runtime systems instead of types systems and static analysis.
-In fact, support for type systems has been lacking in both academic and industry software engineering tool design.
-The most common approach for supporting type systems we see in current programming editors and IDEs is to display type annotations when hovering on an expression and show a squiggly line to indicate the type error location.
-In \chameleon{}, colours and geometric shapes are used to create visual clarity of different components of a type error; interactivity is used to show conditional information and to change the granularity of details.
+These debugging tools are great motivations for the development of \chameleon{}.
+However they focus on different aspect of debugging.
+Java Whyline mainly tackle the problem of unintended runtime behavior while HazelTutor specialized in interactive actions supporting type holes.

### 2022-12-03 main.tex
Fix latex errors
Assets and code listing
Wide images
Recent changes

@@
 \usepackage{tabularx}
+\usepackage{minted}
+\usepackage{graphicx}
+\usepackage[hidelinks]{hyperref}
+
+\usemintedstyle{borland}
 
@@
     T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}
+
+\newcommand{\ignore}[1]{}
+
 \begin{document}
@@
 
+\ignore{
 \author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
@@
 }
+}
 
@@
 We demonstrate \chameleon{}, a type debugging tool capable of displaying the full context of a type error and allowing programmers to search for where the type error actually occurs based on their prior knowledge and to incrementally verify their assumptions.
-\chameleon{} was developed through a user-centred, iterative process with four participatory experiments.
+\chameleon{} was developed through a user-centered, iterative process with four participatory experiments.
 Through the experiments, we showed that programmers using \chameleon{} fix type errors faster than traditional text-based error messages.
@@
 \bibliographystyle{IEEEtran}
-\bibliography{reference}
+\bibliography{icpc}
 

### 2022-12-03 sections/chameleon.tex
Wide images
Recent changes

@@
 \chameleon{} comprises two parts: a type inference engine and an original debugging interface.
-This separate architecture allows us to 1) adapt \chameleon{} into multiple front ends, such as IDE extensions, desktop applications, and web applications, and 2) reuse the debugging interface for different programming languages by replacing the back end type inference engine.
+This separate architecture allows us to 1) adapt \chameleon{} into multiple front ends, such as IDE extensions, desktop applications, and web applications, and 2) reuse the debugging interface for different programming languages by replacing the back-end type inference engine.
 The debugging interface is designed from the ground up; the type inference engine is a re-implementation of the original Chameleon with several novel improvements.
@@
 Chameleon was originally a command-line tool developed in the early 2000s to improve type error reporting and introduced a design for interactive type debugging for the Haskell programming language.
-Unlike traditional type errors produced by the Glasgow Haskell Compiler (GHC)~\cite{ghc}, which uses a Hindley–Milner type inference system, Chameleon infers types using constraint solving.
+Unlike traditional type errors produced by the Glasgow Haskell Compiler (GHC)~\cite{ben_gamari_home_2022}, which uses a Hindley–Milner type inference system, Chameleon infers types using constraint solving.
 In Chameleon, Constraints are generated from the source code based on typing rules.
 In addition, each constraint is labeled with the location (line and column numbers) where it is generated.
-This set of constraints is consistent if the program is well-typed, inconsistent if otherwise.
-When a type error occurs, an efficient algorithm is used to derive a minimal subset of the constraints that still contains inconsistencies.
+This set of constraints is consistent if the program is well-typed and inconsistent otherwise.
+When a type error occurs, an efficient algorithm is used to derive a minimal subset of the constraints that still contain inconsistencies.
 This subset is called a Minimal Unsatisfiable Subset (MUS).
 From this, Chameleon can report a list of locations, using the labels of constraints that are in the MUS.
-Stuckey et al.~\cite{stuckey2003interactive} showed that program locations linked to the constraints from a MUS are all relevant to the type error and must include the cause of the error.
+Stuckey et al.~\cite{stuckey_interactive_2003} showed that program locations linked to the constraints from a MUS are all relevant to the type error and must include the cause of the error.
 In this paper, we will refer to these locations as \textit{error locations}.
 
-Despite successfully borrowing the underlying ideas, we could not reuse the implementation of Chameleon due to the project language standard and libraries used was out of date.
+Despite successfully borrowing the underlying ideas, we could not reuse the implementation of Chameleon since the project language standard and libraries used were out of date.
 In addition to what Chameleon can do, a few new type inference features were added in the \chameleon{} implementation.
@@
 
-Using only constraints from the MUS is sufficient to locating the type error, but to recover types from type errors we need the constraints from parts of the program that irrelevant to the type errors.
+Using only constraints from the MUS is sufficient to locate the type error, but to recover types from type errors we need  constraints from parts of the program that are irrelevant to the type error.
 For instance, consider an ill-typed 2-tuple where two possible types can be assigned: \texttt{(Int, Int)} and \texttt{(Int, String)}.
 The types reconstructed from Chameleon may be \texttt{(a, Int)} and \texttt{(a, String)}.
-Although the recovered types are theoretically correct, they introduce new type variable \texttt{a} making the error message harder to understand.
+Although the recovered types are theoretically correct, they introduce a new type variable \texttt{a} making the error message harder to understand.
 To solve this issue in \chameleon{}, we first expand the MUS to a maximally satisfiable subset.
@@
 In addition, \chameleon{} provides support for type explanation.
-Similar to the type explanation system ~\cite{jun2002explaining} from Yang,  \chameleon{} is able to produce a human-readable explanation, but for type errors.
+Similar to the type explanation system ~\cite{jun_explaining_2002} from Yang,  \chameleon{} is able to produce a human-readable explanation, but for type errors.
 This is achieved by expanding the metadata fields of each constraint to contain the information needed for type explanation.
@@
 % \end{lstlisting}
-
+\begin{listing}[!ht]
 \begin{minted}{haskell}
@@
 \end{minted}
-
-For instance, the program in the listing \ref{listing:ifelse} \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition).
+\caption{A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.}
+\label{listing:1}
+\end{listing}
+For instance, the program in the listing \ref{listing:1} \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition).
 Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed.
-\chameleon{} can construct a human-readable explanation from the MUS, for example: \texttt{a} has type \texttt{Bool}, because \texttt{a} is the condition of an if statement; however \texttt{a} has type \texttt{String}, because \texttt{a} is defined as the string literal \texttt{"True"}.
+\chameleon{} can construct a human-readable explanation from the MUS, for example, \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}.
 This low-level explanation is used in a high-level user interface described in section \ref{sub:deduction-steps} to help programmers reason about the validity of error locations.
 
-\begin{figure}[h]
+\begin{figure*}
     \centering
-    \includegraphics[width=\linewidth]{images/atonomy.pdf}
+    \includegraphics[width=\textwidth]{images/atonomy.pdf}
     \caption{
         \textbf{The anatomy of \chameleon{}.}
-        \chameleon{} consists of two windows: an editor window on the left and a
-        debugging window on the right. The editor window is similar to a
+        \chameleon{} consists of two panes: an editor pane on the left and a
+        debugging pane on the right. The editor pane is similar to a
         traditional code editor. Fragments of source code may have a highlight
         color (A). Additionally, an explanation layer (B) displays in the editor
-        window, if deduction steps are enabled. The debugging window contains
-        three panels. A message panel contains an error description (C), and, optionally,
-        a list of candidate expression cards (D), a list of deduction steps (E), 
-        and a control bar (F) to increment/decrement deduction step. 
+        window, if deduction steps are enabled. The debugging window contains three blocks. 
+        An error statement and configuration block, containing
+        An error statement (D), and, optionally,
+        a list of candidate expression cards (E), a list of deduction steps (F), 
+        and a control bar (G) to increment/decrement deduction step. 
         The conflicting types
-        panel (G) shows two possible types inferred by the type inference
-        engine. A relevant type information panel to show additional information
+        block shows two alternative types (H) inferred by the type inference
+        engine. A relevant type information block to show additional information (I)
         that may help understand type errors.
@@
     \label{fig:anatomy}
-\end{figure}
+\end{figure*}
 
@@
 
-The \chameleon{} debugging interface provides three main features to visualize the relation between an error statement and the error locations in the source code, and to explore the explanation of each error location given by the type inference engine.
+The \chameleon{} debugging interface provides three main features to visualize the relation between an error statement and the error locations in the source code and to explore the explanation of each error location given by the type inference engine.
 Further, programmers  can turn on and off features to suit their preferences and debugging needs.
@@
 \paragraph{Type compare tool} \label{sub:type-compare}
+In designing \chameleon{}, we want to argue against the practice of reporting type errors as the conflicts between \texttt{Expected type} and \texttt{Actual type}.
+Technically, the two alternatives are merely the side effect of different unification orders of the internal type-checking technique and have no bearing on what the programmers think and expect.
 
@@
 If the programmers know the expression's intended type (they usually do), they will be able to eliminate half of the possible locations.
-To make the bisecting effect more pronounced, a user can hover on one of the possible types and only show the relevant locations that
-contribute to that type. This is a convenient way to put the error "under the spotlight".
+To make the bisecting effect more pronounced, a user can hover over one of the possible types and only show the relevant locations that
+contribute to that type. This is a convenient way to put the error \texttt{under the spotlight}.
 
@@
 We propose candidate expression cards to make it clear to the user that there are multiple ways to resolve a type conflict.
-By contrast,  standard compiler error messages such as those of GHC arbitrarily focus the users attention on only one candidate expression, e.g.\ ``\textit{Couldn't match expected type ‘Char’ with actual type ‘Bool’ in expression x}" here x is the candidate expression.
-In practice this candidate expression often does not match programmers' intention (Fig.
+By contrast,  standard compiler error messages such as those of GHC arbitrarily focus the users' attention on only one candidate expression, e.g.\ ``\textit{Couldn't match expected type ‘Char’ with actual type ‘Bool’ in expression x}" where x is the candidate expression.
+In practice, this candidate expression often does not match programmers' intention (Fig.
 \ref{fig:single-candidate}).
@@
 
-This feature can be effective in resolving the problem that traditional tools tend to report errors that happened in source code from third party libraries.
-With candidate expression cards, a programmer can freely change the context of the type error to the user defined variables and functions to gain a better understanding of why her own code does not match the library code instead of the other direction.
+This feature can be effective in resolving the problem that traditional tools tend to report errors that happened in source code from third-party libraries.
+With candidate expression cards, a programmer can freely change the context of the type error to the user-defined variables and functions to gain a better understanding of why her own code does not match the library code instead of the other direction.
 
@@
 
-Deduction steps are motivated by the lack of explain why the program has  type errors.
-When compiler reject a program, it dumps the internal state of type checking.
+Deduction steps are motivated by the lack of explaining why the program has type errors.
+When the compiler rejects a program, it dumps the internal state of type checking.
 This information may be a result of complex computing but this process is not reported by compilers.
 For a typical type error shown in Fig.
-\ref{fig:ghc-error-example}, clearly the evidence for type error are gathered from are generated from the previous two declarations.
-These have to be re-discovered by the programmers again, using harder and less sound methods.
+\ref{fig:ghc-error-example}, clearly, the evidence for the type error is gathered from the previous two declarations.
+These have to be rediscovered by the programmers again, using harder and less-sound methods.
 
@@
 Programmers increment or decrement the step number using the step control bar (Fig.
-\ref{fig:anatomy}-F) or by directly clicking on a step button (Fig.
-\ref{fig:anatomy}-E).
-To increment or decrement the deduction step can be intuitively thought of as to move the position of the \textit{split point} where the blue and orange highlights divide (right).
+\ref{fig:anatomy}-G) or by directly clicking on a step button (Fig.
+\ref{fig:anatomy}-F).
+To increment or decrement the deduction step can be intuitively thought of as moving the position of the \textit{split point} where the blue and orange highlights divide (right).
         }
@@
 
-Nielson pointed out \cite{nielsen1994usability}  that the two most important issues in designing for usability are understanding the users' tasks and the differences in users.
+Nielson pointed out \cite{jakob_nielsen_usability_1993}  that the two most important issues in designing for usability are understanding the users' tasks and the differences in users.
 From analyzing how users use \chameleon{}, we realized that the ideal debugging interface should adapt to the specific programmer and programming task.
@@
 Programmers can switch between modes by clicking on the mode switching toggles (Fig.
-\ref{fig:anatomy}-I).
+\ref{fig:anatomy}-C).
 The features accessible from different modes are shown in table~\ref{tab:chameleon-features}.

### 2022-12-03 sections/conclusion.tex
Recent changes

@@
 We present \chameleon{}, a type debugging tool for the Haskell programming language.
-\chameleon{} facilitates a constraint-based type inference back end that provides accurate and comprehensive type error reports.
-\chameleon{} also facilitates an interactive front end with three different display modes, enabling programmers to navigate deep type relations and triangulate conflicting statements in source code.
-Four participatory experiments were conducted to evaluate the tool's design with programmers with varied experiences.
+\chameleon{} facilitates a constraint-based type inference engine that provides accurate and comprehensive type error reports.
+\chameleon{} also features an interactive debugging interface with three different debugging modes, enabling programmers to navigate deep type relations and triangulate conflicting statements in source code.
+Four experiments with human participants were conducted to evaluate the tool's design with programmers with varied experiences.
 We found that, particularly for more complex tasks, \chameleon{} helped programmers to fix type errors more quickly than traditional text-based error messages.
-Further, programmers actively using \chameleon{} interactive features are shown to be more efficient in fixing type errors than simply reading the type error output.
+Further, programmers actively using \chameleon{} interactive features are shown to fix type errors faster than simply reading the type error output.

### 2022-12-03 sections/discussion.tex
Assets and code listing
Recent changes

@@
 We believe one factor of this effectiveness of \chameleon{} is to reduce the amount of reading time.
-Jbara \cite{jbara2017programmers} showed that reading source code is generally the initial step of solving programming problems, and is done in several passes.
+Jbara \cite{jbara2017programmers} showed that reading source code is generally the initial step of solving programming problems and is done in several passes.
 Although traditional compiler error message tools initially show fewer locations, it is unreliable, meaning that programmers have to expand the reading span without clear guidance.
@@
 
-\begin{lstlisting}[language=Haskell, float, caption={
-    In this simple program, \chameleon{} report 
-    the error happened in \texttt{f} and \texttt{z}.
-}, label={listing:fz}]
+\begin{listing}[!ht]
+\begin{minted}{haskell}
 f z     
@@
     | z == '4' = True
-\end{lstlisting}
+\end{minted}
+\caption{     In this simple program, \chameleon{} report the error happened in \texttt{f} and \texttt{z}.}
+\label{listing:2}
+\end{listing}
 
 We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information.
-With the program showing in the listing \ref{listing:fz}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}.
+With the program showing in the listing \ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}.
 Although, in theory, these two statements are to the same effect, programmers are often required to compute the latter from the former or vice versa.
@@
 Programmers have to remember all the intermediate types and their reasoning throughout such mental gymnastics.
-Assisted by candidate expression cards and deduction steps, this intermediate information is preserved on screen and can be retrieved anytime.
-This may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.
+Assisted by candidate expression cards and deduction steps, this intermediate information is externalized on screen and can be retrieved anytime.
+A recent study on working memory \cite{crichton_role_2021} suggested this approach may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.
 
@@
 
-Another area for future study is to extend on the evaluation process.
-In the user studies 2, 3, and 4, we used harder tasks compared to the ones in user study 1.
+Another area for future study is to extend the evaluation process.
+In studies 1b and 2, we used harder tasks compared to study 1a.
 However, the difficulty level is nowhere as hard as a professional Haskell programmer may face in the production Haskell codebase.
@@
 Our study focuses on the Haskell language for its popularity in the academic world.
-However, the low-quality of textual type errors is a problem not limited to Haskell.
+However, the low quality of textual type errors is a problem not limited to Haskell.
 Modern statically typed languages more or less all share the same problem.
@@
 Our design of \chameleon{}  emphasizes allowing programmers to update the type error based on their prior knowledge.
-It unfolds to answering three questions: which expression should be considered un-typeable?
+It unfolds to answer three questions: which expression should be considered un-typeable?
 Between multiple possible types, which one should be considered intended?

### 2022-12-03 sections/evaluation.tex
Wide images
Recent changes

@@
 
-Over 12 months, several studies were conducted to answer new research questions or rephrase existing ones.
-These studies are grouped in 3 phases, each tested a different version of \chameleon{} with human participants.
-The number of participants and their haskell experience is shown in the table below.
+Three studies were conducted to answer new research questions or rephrase existing ones.
+These studies are grouped into 2 phases, and each tested a different version of \chameleon{} with human participants.
+The number of participants and their Haskell experience, as well as key research questions and conclusions, were shown in figure \ref{fig:timeline}.
 
-A summary of the timeline of the user studies, their major questions and conclusions is given in Figure~\ref{fig:timeline}.
-
-\begin{figure}[h]
+\begin{figure*}
     \centering
-    \includegraphics[width=\linewidth]{images/timeline.pdf}
-    \caption{The timeline of \chameleon{}  development and 
-    evaluation.}
+    \includegraphics[width=\textwidth]{images/timeline.pdf}
+    \caption{The timeline of \chameleon{}  evaluation.}
     \label{fig:timeline}
-\end{figure}
+\end{figure*}
 
@@
 
-Participant recruiting channel was the social news aggregation website Reddit.
-Specifically, our user study advertisements were posted in the Haskell programming language community "r/haskell" and a general programming language-focused community "r/programminglanguages".
+The recruiting channel we used was the social news aggregation website Reddit.
+Specifically, our user study advertisements were posted in the Haskell programming language community \textit{r/haskell} and a general "programming language" focused community \textit{r/programminglanguages}.
 Recruiting from social media allowed us to access a more diverse demographic that better represent the true population of Haskell programmers.
@@
 
-One consequence of our recruiting approach is it is harder for previous participants to enter an later study.
-To minimise the lurking variable of previous experience, we use new code challenge every study and conduct trial runs in every study to bring new participants up to speed.
+One consequence of our recruiting approach is it is harder for previous participants to enter a later study.
+To minimize the lurking variable of previous experience, we use new code challenges in every study and conduct trial runs in every study to bring new participants up to speed.
 
 \subsubsection*{\textbf{Experiment setting}}
-The experiment took place remotely and unsupervised.
-Participants took the study online via web browser and at the physical venue of their choosing.
+The experiments took place remotely and unsupervised.
+Participants took the study online via a web browser and at the physical venue of their choosing.
 All user studies use a web-based debugging environment developed by the authors.
 Conducting the studies online helped us avoid variation when performing tasks in unfamiliar places and using different setups.
-The downside is that to intervene when a user encounters bugs or usability issues mid-study is impossible.
+The downside is that intervening when users encounter usability issues mid-study is impossible.
 To ensure there is no major usability issue that could lower the quality of data collection, we conducted cognitive walkthroughs and sandbox pilots before running each study.
@@
 \subsubsection*{\textbf{Training and group assignment}}
-After consent, participants received interactive training on every function of the tool interface and interactive features.
-Participants were also shown the important functionalities of the tool interface in a summarized cheat sheet.
+After consent, participants received interactive training on the tool interface and interactive features.
+Participants were also shown a cheat sheet summarizing the key functionality of the interface.
 Participants had access to the cheat sheet at all times during the study.
-Participants were given 4 trial runs (2 for each setting) before the data collection starts.
+Participants were given 4 trial runs (2 for each setting) before the data collection started.
 
 All the user studies used a within-subject design to evaluate the effectiveness of different tools or feature sets while counterbalancing the difference in programming proficiency between participants.
-In each user study, participants were required to complete a series of programming tasks (8 for user studies 1 - 3, 9 for user study 4).
+In each study, participants were required to complete a series of programming tasks (8 for studies 1a and 1b, 9 for study 2).
 At each task, a participant receives a single Haskell file that contains one or more type errors.
-The participant was asked to make the code type check with the help of the tool given for this task.
-
-% During each study, a participant received two (user studies 1 - 3) or three (user study 4) different debugging tools.
-The debugging tools were different systems (user studies 1 - 3) or different feature sets of the same system (user study 4).
-For each participant, the debugging tools alternative through consecutive tasks to ensure the fatigue level does not impair the rigorousness of data collection.
+The participant was asked to make the code type check with the help of the given tool.
 
@@
 The data is automatically recorded by the online debugging environment.
-To not introduce a barrier to completing the study, every task can be skipped if the participant made three attempts or spent over 1 minute on the task.
+To not introduce a barrier to completing the study, every task can be skipped if the participant made three failed attempts or is stuck for over 1 minute on the task.
 
-After completing all the tasks (8 in user studies 1-3, 9 in user study 4), users are prompted to fill in a debriefing survey.
-The survey questions include their Haskell programming experience and feedback on the tools and feature sets participants used during the study.
+After completing all the tasks, users are prompted to complete a debriefing survey.
+The survey questions include their Haskell experience and feedback on the tools and feature set participants used during the study.
 
-We used an browser session recording tool~\cite{openreplay} to record the user study sessions.
-These recordings are used to find usability issues in the study and recognize general patterns.
-However, the recording technology does not provide high enough sampling rate for us to be confidently used for rigorous analysis.
+We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions.
+This provides us with a "poor man's video recording"  to find usability issues in the study and recognize general patterns.
+The only reason this format is inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.
 
-\subsection{\chameleon{} User Studies}
+One other use of the session recording is for outlier trimming.
+identify participants who left the computer for a long period of time.
+We first ran a statistical analysis to identify the long gap (2sd) between program logs for each participant.
+Then, we manually checked with the session recording to verify there was actually no user input (mouse movement, scrolling).
 
-\subsubsection{\textbf{Version 1}}  
-\chameleon{} version 1 provides the rewrite of the Chameleon type inference engine and the type compare tool to show the alternative types and possible error locations in matching colors.
+\subsection{\chameleon{} Human Studies}
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\linewidth]{images/chameleon-v1.pdf}
-    \caption{
-        \chameleon{} Version 1
-        }
-    \label{fig:chameleon-v1}
-\end{figure}
+\subsubsection{\textbf{\chameleon{} 1}}  
+\chameleon{} 1 is an earlier iteration of \chameleon{} development.
+This version features the type inference engine that recovers most concrete types after type errors occur and a minimal set of debugging features.
+Key features in \chameleon{} 1 include showing two alternative types for the ill-typed expression,  providing all possible locations that contribute to the type error, and dividing possible locations into two groups based on which alternative type can be inferred.
+In terms of features, \chameleon{} 1 is equivalent to \chameleon{} 2 in basic mode.
 
-Two user studies were conducted to test \chameleon{} version 1, to compare the effectiveness of solving type errors using \chameleon{} and GHC compiler error messages.
-We choose GHC compiler error messages as the baseline because it is the canonical tool for working with type error in Haskell.
+% \begin{figure}[h]
+%     \centering
+%     \includegraphics[width=\linewidth]{images/chameleon-v1.pdf}
+%     \caption{
+%         \chameleon{} 1
+%         }
+%     \label{fig:chameleon-v1}
+% \end{figure}
+
+Two user studies (1a & 1b) were conducted to compare the effectiveness of solving type errors using \chameleon{} 1 and GHC compiler error messages.
+We choose GHC compiler error messages as the baseline because it is the canonical tool for working with type errors in Haskell.
 Although high-level tools like Haskell Language Server exist, they generally relay the GHC error messages verbatim.
+The procedure of these studies is shown in figure \ref{fig:procedure-1}.
 
 Eight tasks were given in both studies.
-In the first study, the tasks were taken from a exercises from the exercises of the Haskell programming units.
+In study 1a, the tasks were taken from exercises from the Haskell programming units.
 These tasks cover a variety of type errors.
-\tf{break down the type error classes.} However these tasks, as shown from the result and feedback of user study 1, were too simple to fully evaluate the two options.
-In the second study, the tasks are sourced from the top 20 Haskell topics on GitHub~\cite{githubHaskell}.
+However, these tasks, as shown from the result and feedback of user study 1, were too simple to fully evaluate the two options.
+In the second study, the tasks are sourced from the top 20 Haskell topics on GitHub~\cite{github_github_2022}.
 The authors manually added type errors.
@@
     \caption{}
-    \label{fig:study-process}
+    \label{fig:procedure-1}
 \end{figure}
@@
     \includegraphics[width=\linewidth]{images/user-study-1a.pdf}
-    \caption{Time to complete by task in user study 1a with confidence interval}
-    \label{fig:r1-analysis}
+    \caption{Time to complete each task in user study 1a with 95\% confidence interval}
+    \label{fig:analysis-1a}
 \end{figure}
@@
     \includegraphics[width=\linewidth]{images/user-study-1b.pdf}
-    \caption{Time to complete by task in user study 1b confidence interval}
-    \label{fig:r2-analysis}
+    \caption{Time to complete each task in user study 1b with 95\% confidence interval}
+    \label{fig:analysis-1b}
 \end{figure}
@@
 
-From the data collected during user study 1 (\ref{fig:r1-analysis}), we could not identify which group performed better (p-value =0.2041, Wilcoxin signed test).
+From the data collected during study 1a (\ref{fig:analysis-1a}), we were not able to identify which group performed better.
 For the trivial challenges we set for users, the individual differences are generally more significant than differences between treatments.
-The most common feedback (7 out of 35) from the user study 1 was that the tasks were too trivial to invite meaningful evaluation.
-One participant put it, "Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice".
+The most common feedback (7 out of 35) the study 1a was that the tasks were too trivial to invite meaningful evaluation.
+One participant said, "Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."
 
-One interesting exception is task 8, where the \chameleon{} group outperformed the GHC group more consistently and with a larger effect size (p-value = 0.002456, Wilcoxin signed test).
-The discrepancy is thought to be related to the nature of Task 8:
-\begin{itemize}
-    \item {It has a longer source file than other tasks (only shorter than task 6, but task
-    6 contains two independent type errors while task 8 is one connected task
-    error);}
-    \item {It is more complex than others (involves abstract data types and function application); and
-    }
-    \item {
+One interesting observation is task 8, where the \chameleon{} group outperformed the GHC group more consistently and with a larger effect size (p-value = 0.002456, Wilcoxin signed test).
+The discrepancy is thought to be related to the nature of Task 8.
+It has a lengthier source file and involves more language features (abstract data types and high-level functions).
 GHC struggles to produce a relevant error message for this type of error.
-    }
-  \end{itemize}
-It can be further shown by examining the process some participants took to solve the problem.
+
+In study 1b, the \chameleon{} group solved the type error faster (p =0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1.
+The authors are unclear about the cause of low performance from \chameleon{} in task 1.
+It is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity and ineffective training.
+For all the other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program.
+
+In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}.
+"It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant.
+"I think \chameleon{}  does a much better job than GHC's error messages.
+I like that it shows the sources for the type judgments.
+This makes it quite easy to figure out how to rectify errors." reported another participant.
+
+% \begin{itemize}
+%     \item {It has a longer source file than other tasks (only shorter than task 6, but task
+%     6 contains two independent type errors while task 8 is one connected task
+%     error);}
+%     \item {It is more complex than others (involves abstract data types and function application); and
+%     }
+%     \item {
+%         GHC struggles to produce a relevant error message for this type of error.
+%     }
+%   \end{itemize}
 
@@
 
-In comparison, the \chameleon{} group solved the type error faster than the GHC group in almost all tasks (p =0.00742, Wilcoxin signed test) in study 2 (figure \ref{fig:r2-analysis}), barring task 1.
-The authors are unclear about the cause of low performance from \chameleon{} in task 1.
-It is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity.
-From the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program.
+\subsubsection{\textbf{\chameleon{} 2}}  \label{sub:us4}
 
-In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}.
-"Its most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant.
-"I think \chameleon{}  does a much better job than GHC's error messages.
-I like that it shows the sources for the type judgements.
-This makes it quite easy to figure out how to rectify errors." reported another participant.
+\chameleon{} 2 is latest version of \chameleon{}.
+Advanced debugging features such as deduction steps and candidate expressions were available in this iteration.
+In addition, mode switching is enabled by default as well.
+A few other user interfaces were designed and prototyped between the development of \chameleon{} 1 and \chameleon{} 2.
+Study 2 was designed to evaluate \chameleon{} 2.
+We were trying to answer the research question: \textit{How do programmers use the interactive features in \chameleon{} 2?}.
+More specifically, we were trying to inquiry the following three sub-questions:
+\begin{itemize}
+    \item Do programmers prefer switching modes during debugging type errors?
+    \item What are programmers' preferences among the three modes provided by \chameleon{} 2?
+    \item How do programmers use the advanced features provided by \chameleon{} 2?
 
-\subsubsection{\textbf{Version 2}}  \label{sub:us4}
+\end{itemize}
 
-Version 2 is latest version of \chameleon{}.
-Not only the added features the deduction steps and candidate expressions are We designed an exploratory study to get better understanding of how users learn and use the multi-mode design of \chameleon{} version 3.
-During the study the initial mode of each task alternative through the three different modes, and repeat three cycles in nine tasks.
+During the study, the initial mode of each task alternated through the three different modes and repeated three cycles in nine tasks.
 The order of the three modes in each cycle is counterbalanced among all participants.
-However, participants are free to switch to other modes at any time.
+However, participants can switch to other modes at any time (figure \ref{fig:procedure-2}).
 
@@
     \caption{}
-    \label{fig:study-process}
+    \label{fig:procedure-2}
 \end{figure}
@@
 
-The study of \chameleon{} 2 is exploratory, in that we were hoping to let programmers discover their own way of using the tool.
-In post-hoc analysis of the collected log data, we were able to extrapolate some interesting patterns of how the tool was used.
+The study of \chameleon{} 2 is an exploratory study.
+We were expecting to let programmers discover their own way of using the tool.
+In post hoc analysis of the collected log data, we were able to extrapolate some interesting patterns of how the tool was used.
 We share a few of our observations in this section.
 
-The most striking feature of the data is users' tendency of using tools differ wildly.
-That is to say, some users used the feature extensively however some users completed the tasks without actively explore the given information.
-Based on this difference we divided the users into three groups.
+The most striking feature of the data is users' tendencies to use the tool differ wildly.
+Some users used the feature extensively however some users completed the tasks without actively exploring the given information.
+Based on this discrepancy, we divided the users into three groups.
 
@@
         Low Interaction & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
-        High Interaction & Users did everything from the Common feature group but used features specific to Balanced and Advanced mode, such as activating steps and expression cards. \\ \hline
-
+        High Interaction & Users did everything from the Common feature group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards.
+\\ \hline
 \end{tabularx}
@@
 As discussed in user studies 1 and 2, it is likely related to the nature of these tasks.
-Tasks 4 and 6 are shorter tasks in the study.
+Tasks 4 and 6 are shorter than other tasks.
 The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code).
-These may be useful because following the natural reading order will allow participants to skip lines on the bottom, and thus the advantage of \chameleon{} is less important.
-On the other hand, task 9 is the longest task of all.
-It also involves harder problems such as mutually recursive type definitions.
+These may be useful because following the natural reading order will allow participants to skip lines on the bottom, and thus the advantage of \chameleon{} is less critical.
+On the other hand, task 9 is the lengthiest task of all.
+It also involves more complex features, such as mutually recursive type definitions.
 
@@
     \includegraphics[width=\linewidth]{images/user-study-2.pdf}
-    \caption{Time to complete by task in user study 2 with 95\% confidence interval}
+    \caption{Time to complete each task in study 2 with 95\% confidence interval}
     \label{fig:r4-analysis}
@@
 One participant fixed a minor run-time issue (a problematic pattern matching of `[attributes]` on line 18) however failed to identify the culprit of the type error.
+This location is not pinpointed by \chameleon{} and therefore it is understandable that programmers who used the deduction steps would be able to ignore this part of the source code.
 %     }
@@
 
-Another observation is when using the mode switching feature of \chameleon{}, programmers generally switching form a less informative display to a more informative one (Fig.
+Another observation is when using the mode switching feature of \chameleon{}, we find presenting the starting mode and finishing mode of each task and each participant in a confusion matrix is very telling (Fig.
 \ref{fig:r4-mode-switching}).
-
+This observation suggests two characteristics of using multi-mode debugging tools.
+First, programmers are generally reluctant to switch modes (146/277 changing modes vs.
+131/277 staying in the same mode).
+Second, when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
 \begin{figure}[h]
@@
     \caption{
-        The rows represent the finishing mode, columns starting modes.
+        % The rows represent the finishing mode, columns starting modes.
     }

### 2022-12-03 sections/introduction.tex
Wide images
Recent changes

@@
-Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer2012static} and codebase maintainability~\cite{kleinschmager2012static}.
-In recent years, the programming community has seen a trend of migrating away from dynamically typed codebases due to the lack of maintainability and refactoring safety when systems reach large scale and complexity~\cite{chatley2019next700}.
+Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}.
+In recent years, the programming community has seen a trend of migrating away from dynamically typed codebases due to the lack of maintainability and refactoring safety when systems reach large scale and complexity~\cite{chatley_next_2019}.
 These migration methods include the introduction of gradual type annotations (e.g., JavaScript to TypeScript or Python with mypy) or transition to a modern strongly-typed language (Scala, Rust, Haskell, PureScript, etc.).
-However, with dynamically typed languages (e.g., JavaScript, PHP, and Python) still being most peoples' formative experience with programming and with such languages now firmly entrenched in education~\cite{so2022survey}, this transition may pose difficulties for programmers who lack experience writing type-safe programs.
+However, with dynamically typed languages (e.g., JavaScript, PHP, and Python) still being most peoples' formative experience with programming and with such languages now firmly entrenched in education~\cite{stackoverflow_stack_2022}, this transition may pose difficulties for programmers who lack experience writing type-safe programs.
 
-Programmers find challenges adopting typed languages in their practice. They tend to be verbose, hard to learn, and generate bad error messages. 
-One obstacle programmers face when using a statically typed language is understanding and resolving type errors~\cite{marceau2011measuring, tirronen2015understanding}, especially when dealing with modern type features such as polymorphic types and implicit typing.
-Studies show most statically typed languages produce unhelpful and even misleading type errors~\cite{yang2000improved, some more proof}.
+Programmers find challenges in adopting typed languages in their practice. They tend to be verbose, hard to learn and generate bad error messages. 
+One obstacle programmers face when using a statically typed language is understanding and resolving type errors~\cite{marceau_measuring_2011, tirronen_understanding_2015}, especially when dealing with modern type features such as polymorphic types and implicit typing.
+Studies show most statically typed languages produce unhelpful and even misleading type errors~\cite{}.
 Symptoms of bad error messages include cryptic language exposing internal constructs of the compiler and incomplete or wrong error locations.
-These usability issues poses challenges to learners and teachers of the languages.
+These usability issues pose challenges to learners and teachers of the languages.
 impede the wide adoption of statically typed languages, especially for programmers who are now accustomed to the mindset of writing un-typed programs.
 
-This paper introduces \chameleon{}, an interactive debugging tool to visualize critical context of a type error (where it happened or could have happened, and which chunks of code caused it) and allow programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve that ambiguity.
-Most noticeably, features (Section \ref{sub:type-compare}) provided by \chameleon{} are the type compare tool, the deduction step feature , and the candidate expression card feature.
+This paper introduces \chameleon{}, an interactive debugging tool to visualize the relevant context of a type error (where it happens or could have happened, and which parts of code cause it) and allow programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve that ambiguity.
+Most noticeably, features (Section \ref{sub:type-compare}) provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature.
 In addition, we integrated all the features into a unified debugging interface and provided functionality for programmers to enable and disable the features based on their preferences and debugging needs.
-\chameleon{} is open-source and is available in both web~\cite{chameleon} and desktop versions.
+\chameleon{} is open-source and is available in both web~\cite{shuai_fu_chameleon_2022} and desktop versions.
 The current implementation of \chameleon{} targets the Haskell 2010 language standard.
@@
 Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages.
-Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak2007history}, and now it can be found in most popular languages such as C\#~\cite{csharpgenerics}, Java~\cite{javagenerics}, and TypeScript~\cite{tsgenerics}.
+Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak_history_2007}, and now can be found in most popular languages such as C\#~\cite{bill_wagner_constraints_2022}, Java~\cite{oracle_generic_2022}, and TypeScript~\cite{microsoft_documentation_2022}.
 
@@
 \begin{itemize}
-\item { We provide the design and implementation of the \chameleon{}, to visualize crucial context of a type error and allow programmers to explore and verify the error locations in small chunks interactively.
+\item { We provide the design and implementation of the \chameleon{}, to visualize the relevant context of a type error and allow programmers to explore and verify the error locations in small chunks interactively.
 }

### 2022-12-03 sections/related-work.tex
Recent changes

@@
 \section{Related Work}
-
+%%%  Todo: Discussing "Practical SMT-based Type Error Localization"
 Researchers have long realized that incomprehensible type errors result as a consequence of standard approaches to the type inference process.
-Haack and Wells~\cite{haack2004type} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors.
-To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' Researchers have proposed solutions to improve the type inference process.
-Type error slicing~\cite{haack2004type} is a technique that finds locations that are complete and minimal for the type error.
-Internally labeled constraints and MUS manipulation are used to generate these slices.
-The language supported in Haack's work was a subset of standard ML.
-The original Chameleon~\cite{stuckey2003interactive} used a similar technique but extended it to support advanced type-level features (type classes and functionally dependent types).
-The project also introduced the ability to query type information through a command line-style interface (Fig.~\ref{fig:original-chameleon}).
+Many studied [Citation here] the approach of finding all locations that contribute to a type error.
+Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors.
+To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}''  Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error.
+Internally labeled constraints and Minimal Unsatisfiable Set (MUS) manipulation are used to generate these slices.
+The language supported in Haack's work was a subset of Standard ML.
+The original Chameleon~\cite{stuckey_interactive_2003} used a similar technique but generate constraints in Constraint handling rules (CHR).
+As a result,  Chameleon supported advanced type-level features (type classes and functionally dependent types).
+The project also introduced the ability to query type information through a command line-style interface.
 Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies.
 
-Another direction to approach type error debugging is suggesting fixes.
-Sheng Chen's works [1,2] on counter factual typing is able to suggesting changes to make the ill-typed program typecheck.
-While this is promising ground, the work is not evaluated with human participants.
-\cite{lerner2007searchingtypeerror} used an approach by repeatedly modifying the abstract syntax tree and querying the compiler/type checker to know whether the program still well-typed.
-This approach allows the algorithm to produce suggestions the syntax changes that could fix the type error.
-Benefiting from not needing a type system impelemntation and being language agnositic, the underlying idea was implemented and evaluated on 10 CS students.
-The system can be successfully adopted in not suitable for teaching and learning purpose, as it is not able to explain why the type error happened and why fixes are suggested.
-The lack of explanability is because the information was witheld by the type checker and the inference algorthm does not know (need to know) how types are computed.
+One direction to approach type error debugging is by suggesting solutions to fix the type error.
+Sheng Chen's works \cite{chen_counter-factual_2014, chen_efficient_2020, chen_improving_2022} on counterfactual typing are able to suggest changes to make the ill-typed program type-check.
+While this is a promising ground, the work is not evaluated with human participants.
+\cite{lerner_searching_2007} studied an approach of repeatedly modifying the AST (abstract syntax tree) and querying the compiler/type checker to know whether the new program is well-typed.
+This approach allows the algorithm to suggest syntax changes that could fix the type error.
+Benefiting from not needing a type system implementation and being language agnostic, the underlying idea was implemented and evaluated in two different languages with ten university students.
+While suggesting syntax changes is very intuitive in practice, this approach has limitations for teaching and learning purposes, as it is not able to explain why the type errors occur and why fixes are suggested.
+The lack of explanation is due to the inquiry into the underlying type-check being a binary state.
+Deep type-level analysis (typing rules and language features) was not communicated to the inference algorithm.
 
-One related approach is CEME (compiler error message enhancement).
-Prather showed that ECEM casts a positive result in understanding the type error through a series of mixed methods study.
-Decafe is a tool that is able to rephase the compier error message into an enhanced version.
-A study of over 200 CS1 students showed that CEME reduce overall errors in their assignments.
-Berik proposed a framework of constructing compiler error message based on the augumentation theory.
-In this proposal,  Berik argued that error messages  follow a simple argumentation layout or an exntended argumentation layout will result in a human friendly errors.
-These research shows the significance of improving the language in a compiler error message.
-Most principles and suggestions are followed in \chameleon{} in concustructing error statements.
-However they were not targeting type errors but compiler error (some even include runtime error) in general.
-The nuances of type errors, such as alternative typings were not taken into account.
+Another related approach is through ECEM (Enhanced compiler error message).
+Through a series of mixed-method studies, Prather showed \cite{prather_novices_2017} that ECEM has a positive result in understanding compiler errors.
+Decafe \cite{becker_effective_2016} is a tool that can rephrase the Java compiler error messages into their enhanced version.
+In a study of over 200 CS1 students, Decaf was shown to reduce overall errors in their coding practices.
+Berik proposed a framework \cite{barik_how_2018} for constructing compiler error messages based on argumentation theory.
+In this proposal,  Berik argued that error messages following a simple argumentation layout or an extended argumentation layout are more human-friendly.
+These works show the significance of improving the language in the compiler error messages.
+Most principles and suggestions are followed in \chameleon{} in constructing error statements.
+However, they were not targeting type errors alone but compiler errors (some even include run time errors) in general.
+The nuances of type errors, such as alternative typing, were not considered.
 
@@
 
-Debugging using a GUI Interactive Development Environment has been the standard practice for a very long time.
-Compared to a command line-based interface, a graphic user interface provides programming tools with the ability to show information hierarchy, provide immediate feedback to changes, and display complex visualizations.
-Provide opportunities to run consecutive command to query the information about code or executing command based on the existing code.
-HazelTutor is a interactive filling type holes by suggesting template expressions (strategies as called by the authors).
-It also provide a cursor based type inspector that allows programmers to query the type of parts of the program.
-Whyline~\cite{ko2009finding} is a Java debugging system that allows a user to ask questions like "why does variable X have value Y".
-It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error.
-These debugging tools are great motivations for the development of \chameleon{}.
-However they focus on different aspect of debugging.
-Java Whyline mainly tackle the problem of unintended runtime behavior while HazelTutor specialized in interactive actions supporting type holes.
+Debugging using a GUI Interactive Development Environment has been the standard practice for a long time.
+Compared to a command line-based interface, a graphic user interface provides programming tools that can show information hierarchy, provide immediate feedback to changes, and display complex visualizations.
+Hazel Tutor \cite{potter_hazel_2020} is an interactive type-driven environment for the OCaml language.
+It can automatically fill type holes by suggesting template expressions (strategies as called by the authors) through a popup window.
+It also provides a cursor-based type inspector that allows programmers to query the types of different parts of the program.
+Whyline~\cite{ko_finding_2009} is a Java debugging system that allows a user to ask questions like "why does variable X have value Y." It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error.
+These debugging tools are important motivations for developing \chameleon{}.
+However, they focus on different contexts of the debugging process.
+Java Whyline mainly tackles the problem of unintended run-time behavior, while Hazel Tutor specializes in interactive actions supporting type holes.

### 2022-12-05 main.tex
John's comments

@@
 \newcommand{\tf}[1]{\textcolor{magenta}{\textsc{tf}: #1}}
+
+% Comment out second line to disable.
+\newcommand{\todo}[1]{}
+\renewcommand{\todo}[1]{{\color{red} TODO: {#1}}}
 
@@
 \begin{abstract}
+
+\todo{I would start with saying untyped langs popular e.g. Python and the need to address problems this brings...}
+
 In recent years, many programming communities have been undergoing a transition to writing strongly typed programs.
@@
 \bibliographystyle{IEEEtran}
-\bibliography{icpc}
+\bibliography{ICPC}
 

### 2022-12-05 sections/discussion.tex
John's comments

@@
 \section{Discussion}
+
+\todo{Suggest try and strucuture this a bit more e.g. each para bold the key lesson @ start of para. Can you group into implications for research/implications for practice to make more actionable??}
+
 This paper presents the interactive type debugging tool \chameleon{} and charts the evolution of its design across several iterations in response to user evaluation and feedback, as well as examining the effectiveness of the general approach compared to traditional static type error messages.
@@
 We believe one factor of this effectiveness of \chameleon{} is to reduce the amount of reading time.
-Jbara \cite{jbara2017programmers} showed that reading source code is generally the initial step of solving programming problems and is done in several passes.
+Jbara \cite{jbara_how_2015} showed that reading source code is generally the initial step of solving programming problems and is done in several passes.
 Although traditional compiler error message tools initially show fewer locations, it is unreliable, meaning that programmers have to expand the reading span without clear guidance.
@@
 It may produce effective results by guessing the possible un-typable expression based on the position it appears in the syntax tree and project structure or guessing the intended type signature by comparing the number of supporting slices in the source code.
+
+\subsection{Threats to validity}
+
+\todo{Put threats BEFORE Discussion and at end Eval section?}
+
+\todo{These are too terse/unclear - need expanding. Suggest classify as internal / external threats etc .  What about background and number of participants as threats??}
+
+One consequence of our recruiting approach is it is harder for previous participants to enter a later study.
+To minimize the lurking variable of previous experience, we use new code challenges in every study and conduct trial runs in every study to bring new participants up to speed.
+
+Conducting studies remotely and unsupervised is that intervening when users encounter usability issues mid-study is impossible.
+To ensure there is no major usability issue that could lower the quality of data collection, we conducted cognitive walkthroughs and sandbox pilots before running each study.

### 2022-12-05 sections/evaluation.tex
John's comments

@@
 The number of participants and their Haskell experience, as well as key research questions and conclusions, were shown in figure \ref{fig:timeline}.
+\todo{Explain was a user study of the tool with practitioners?}
 
@@
 
-One consequence of our recruiting approach is it is harder for previous participants to enter a later study.
-To minimize the lurking variable of previous experience, we use new code challenges in every study and conduct trial runs in every study to bring new participants up to speed.
+\todo{Need more detail on participants - years experience, male vs female, etc}
 
@@
 Conducting the studies online helped us avoid variation when performing tasks in unfamiliar places and using different setups.
-The downside is that intervening when users encounter usability issues mid-study is impossible.
-To ensure there is no major usability issue that could lower the quality of data collection, we conducted cognitive walkthroughs and sandbox pilots before running each study.
 
@@
 
-After completing all the tasks, users are prompted to complete a debriefing survey.
+After completing all the tasks, participants are prompted to complete a debriefing survey.
 The survey questions include their Haskell experience and feedback on the tools and feature set participants used during the study.
@@
 We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions.
-This provides us with a "poor man's video recording"  to find usability issues in the study and recognize general patterns.
-The only reason this format is inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.
+This provides us with a "poor man's video recording" to identify usability issues in the study and to recognize general patterns.
+The reason we found this format inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.
 
-One other use of the session recording is for outlier trimming.
-identify participants who left the computer for a long period of time.
-We first ran a statistical analysis to identify the long gap (2sd) between program logs for each participant.
+One use of the session recording is for outlier trimming.
+We identify participants who left the computer for an extended period of time.
+We first ran a statistical analysis to identify the long gap (2 standard deviations) between actions using program logs for each participant.
 Then, we manually checked with the session recording to verify there was actually no user input (mouse movement, scrolling).
@@
 This version features the type inference engine that recovers most concrete types after type errors occur and a minimal set of debugging features.
-Key features in \chameleon{} 1 include showing two alternative types for the ill-typed expression,  providing all possible locations that contribute to the type error, and dividing possible locations into two groups based on which alternative type can be inferred.
-In terms of features, \chameleon{} 1 is equivalent to \chameleon{} 2 in basic mode.
+Key features in \chameleon{} 1 include showing two (or more) alternative types, showing all possible error locations, dividing possible error locations into groups based on alternative types, and concrete type restoration.
+In short, \chameleon{} 1 is equivalent to \chameleon{} 2 set to basic mode.
 
@@
 
-Two user studies (1a & 1b) were conducted to compare the effectiveness of solving type errors using \chameleon{} 1 and GHC compiler error messages.
+Two  studies (1a \& 1b) were conducted to compare the effectiveness of solving type errors using \chameleon{} 1 and GHC compiler error messages.
 We choose GHC compiler error messages as the baseline because it is the canonical tool for working with type errors in Haskell.
@@
 Eight tasks were given in both studies.
-In study 1a, the tasks were taken from exercises from the Haskell programming units.
+In study 1a, the tasks were taken from the exercises of the Haskell programming class in the authors' institute.
 These tasks cover a variety of type errors.
-However, these tasks, as shown from the result and feedback of user study 1, were too simple to fully evaluate the two options.
 In the second study, the tasks are sourced from the top 20 Haskell topics on GitHub~\cite{github_github_2022}.
-The authors manually added type errors.
+The authors then manually added type errors into the program.
+In both studies, the type errors include simple mismatch, confusing syntax, missing instance, precedence and fixation, infinite types, and confusing list versus element.
+The categories of type errors to include in the tasks followed the common type errors in Tirronen's study \cite{tirronen_understanding_2015}.
+
+The  studies (1a \& 1b) are designed to answer the research question:
+\begin{itemize}
+    \item \textbf{RQ1.} Do programmers solve type errors faster with \chameleon{} than GHC compiler error messages?
+\end{itemize}
 
@@
     \includegraphics[width=\linewidth]{images/procedure-1.pdf}
-    \caption{}
+    \caption{\todo{Does this figure really help at all?}}
     \label{fig:procedure-1}
@@
 For the trivial challenges we set for users, the individual differences are generally more significant than differences between treatments.
-The most common feedback (7 out of 35) the study 1a was that the tasks were too trivial to invite meaningful evaluation.
-One participant said, "Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."
-
-One interesting observation is task 8, where the \chameleon{} group outperformed the GHC group more consistently and with a larger effect size (p-value = 0.002456, Wilcoxin signed test).
+One interesting observation is task 8, where the \chameleon{} group outperformed the GHC group more consistently (p-value = 0.002456, Wilcoxin signed test).
 The discrepancy is thought to be related to the nature of Task 8.
@@
 GHC struggles to produce a relevant error message for this type of error.
+We hypothesized that we may observe a more significant difference using tasks with lengthier and more realistic source code.
+This hypothesis is also supported by the most common feedback (7 out of 35) the study 1a was that the tasks were too trivial to invite meaningful evaluation.
+One participant said, "Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."
 
 In study 1b, the \chameleon{} group solved the type error faster (p =0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1.
-The authors are unclear about the cause of low performance from \chameleon{} in task 1.
-It is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity and ineffective training.
+For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity and ineffective training.
 For all the other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program.
-
 In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}.
@@
     \includegraphics[width=\linewidth]{images/procedure-2.pdf}
-    \caption{}
+    \caption{\todo{Does this figure really help at all?}}
     \label{fig:procedure-2}

### 2022-12-05 sections/introduction.tex
John's comments

@@
+\todo{As above, would start with motivation of dynamically typed langs e.g. JavaScript, Pythjon popularity but issues this brings...
+}
+
+Dynamically
+
 Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}.
@@
 
-Programmers find challenges in adopting typed languages in their practice. They tend to be verbose, hard to learn and generate bad error messages. 
+\todo{There are a lot of minor grammer fix-ups needed throughout...}
+Programmers find challenges in adopting typed languages in their practice. \todo{I think this statement is plain wrong myself... 
+ Need strong references to justify if going to say this to SE audience!:}They tend to be verbose, hard to learn and generate bad error messages. 
 One obstacle programmers face when using a statically typed language is understanding and resolving type errors~\cite{marceau_measuring_2011, tirronen_understanding_2015}, especially when dealing with modern type features such as polymorphic types and implicit typing.
@@
 
+\todo{Way too long sentence ; confusing to reader - rephrase:}
 This paper introduces \chameleon{}, an interactive debugging tool to visualize the relevant context of a type error (where it happens or could have happened, and which parts of code cause it) and allow programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve that ambiguity.
-Most noticeably, features (Section \ref{sub:type-compare}) provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature.
+Most noticeably, features provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature.
 In addition, we integrated all the features into a unified debugging interface and provided functionality for programmers to enable and disable the features based on their preferences and debugging needs.
 \chameleon{} is open-source and is available in both web~\cite{shuai_fu_chameleon_2022} and desktop versions.
-The current implementation of \chameleon{} targets the Haskell 2010 language standard.
+\todo{Introduce Haskell earlier?
+Motivate WHY Haskell?} The current implementation of \chameleon{} targets the Haskell 2010 language standard.
 However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques.

### 2022-12-05 sections/related-work.tex
John's comments

@@
 \section{Related Work}
-%%%  Todo: Discussing "Practical SMT-based Type Error Localization"
-Researchers have long realized that incomprehensible type errors result as a consequence of standard approaches to the type inference process.
-Many studied [Citation here] the approach of finding all locations that contribute to a type error.
+
+\todo{I would do a short Motivating Example section straight after intro showing example of untyped lang, problems it brings, and outline requirments for the proposed IDE to address it. Make sense??}
+
+\todo{Do yoou need Related Work here - or put @ end of paper just before Summary - I tend to do the later now UNLESS its REALLY needed by reader - IMO could put @ end and get to your work/contributions much quicker}
+
+\subsection{Finding all type error locations}
 Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors.
-To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}''  Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error.
+To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' Many studied \cite{stuckey_interactive_2003, haack_type_2004, pavlinovic_practical_2015, schilling_constraint-free_2012} the approach of finding all locations that contribute to a type error.
+Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error.
 Internally labeled constraints and Minimal Unsatisfiable Set (MUS) manipulation are used to generate these slices.
@@
 Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies.
+While finding all error locations is useful in comprehending type errors, it is only 1 of the 7 properties listed in the proposed manifesto of good type error reporting~\cite{yang_improved_2000}.
 
+\subsection{Producing high-quality error explanation}
+One weakness of compiler error messages, in general, is that they fail to explain the error in human language.
+As put in~\cite{barik_developers_2017}, "Error messages appear to take the form of natural language, yet are as difficult to read as source code."  A well-studied approach to producing better error explanations is through ECEM (Enhanced compiler error message).
+Through a series of mixed-method studies, Prather showed~\cite{prather_novices_2017} that ECEM has a positive result in understanding compiler errors.
+Decafe~\cite{becker_effective_2016} is a tool that can rephrase the Java compiler error messages into their enhanced version.
+In a study of over 200 CS1 students, Decaf was shown to reduce overall errors in their coding practices.
+Berik proposed a framework~\cite{barik_how_2018} for constructing compiler error messages based on argumentation theory.
+In this proposal,  Berik argued that error messages following a simple argumentation layout or an extended argumentation layout are more human-friendly.
+These works show the significance of improving the language in the compiler error messages.
+Most principles and suggestions are followed in \chameleon{} in constructing error statements.
+However, they were not targeting type errors alone but compiler errors (some even include run time errors) in general.
+The nuances of type errors, such as alternative typing, were not considered.
+Moreover, explanation systems are designed specifically for novice users.
+
+\subsection{Suggesting resolution}
 One direction to approach type error debugging is by suggesting solutions to fix the type error.
-Sheng Chen's works \cite{chen_counter-factual_2014, chen_efficient_2020, chen_improving_2022} on counterfactual typing are able to suggest changes to make the ill-typed program type-check.
+Sheng Chen's works~\cite{chen_counter-factual_2014, chen_efficient_2020, chen_improving_2022} on counterfactual typing are able to suggest changes to make the ill-typed program type-check.
 While this is a promising ground, the work is not evaluated with human participants.
-\cite{lerner_searching_2007} studied an approach of repeatedly modifying the AST (abstract syntax tree) and querying the compiler/type checker to know whether the new program is well-typed.
+Lerner  studied an approach~\cite{lerner_searching_2007} of repeatedly modifying the AST (abstract syntax tree) and querying the compiler/type checker to know whether the new program is well-typed.
 This approach allows the algorithm to suggest syntax changes that could fix the type error.
@@
 
-Another related approach is through ECEM (Enhanced compiler error message).
-Through a series of mixed-method studies, Prather showed \cite{prather_novices_2017} that ECEM has a positive result in understanding compiler errors.
-Decafe \cite{becker_effective_2016} is a tool that can rephrase the Java compiler error messages into their enhanced version.
-In a study of over 200 CS1 students, Decaf was shown to reduce overall errors in their coding practices.
-Berik proposed a framework \cite{barik_how_2018} for constructing compiler error messages based on argumentation theory.
-In this proposal,  Berik argued that error messages following a simple argumentation layout or an extended argumentation layout are more human-friendly.
-These works show the significance of improving the language in the compiler error messages.
-Most principles and suggestions are followed in \chameleon{} in constructing error statements.
-However, they were not targeting type errors alone but compiler errors (some even include run time errors) in general.
-The nuances of type errors, such as alternative typing, were not considered.
+\subsection{Interactive Debugging}
+Interactive debugging tools have been studied using the command line interface.
+Chameleon provides a certain level of interactive debugging~\cite{stuckey_interactive_2003}.
+The interaction is used to query different parts of the program.
+A recent study \cite{tsushima_type_2021} provided an interactive type debugging system that restores type information based on the output from CFT-based type inference.
+Both tools allow programmers to change their assumptions about which expression is ill-typed.
+While both tools show a strong influence on theoretical development, they were not evaluated in realistic studies with human participants.
+In addition, command line-based debugging tools suffer the same limitation as other line-based tools, namely, 1 2 and 3.
 
-% \begin{figure}[h]
-%     \centering
-%     \includegraphics[width=\linewidth]{images/original-chameleon.pdf}
-%     \caption{
-% Originally, Chameleon was a command-line tool that can list all the potential causes of a type error and display the two branches in two forms of highlight (right), achieved with ASCII colors.
-In addition, it allows a few commands (debug and explain) in the debugging shell to help debug type errors (left).}
-%     \label{fig:original-chameleon}
-% \end{figure}
+A graphic user interface (GUI) provides programming tools with immediate feedback on changes and the ability to show complex visualizations.
+Many tools dedicate to improving the debugging experience.
+We list two.
+Hazel Tutor \cite{potter_hazel_2020} is an interactive type-driven environment for the OCaml language.
+It can automatically fill type holes by suggesting template expressions (strategies as called by the authors) through a popup window.
+It also provides a cursor-based type inspector that allows programmers to query the types of different parts of the program.
+Whyline~\cite{ko_finding_2009} is a Java debugging system that allows a user to ask questions like "why does variable X have value Y." It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error.
+These debugging tools are important motivations for developing \chameleon{}.
+However, they focus on different aspects of the debugging process.
+Java Whyline mainly tackles the problem of unintended run-time behavior, while Hazel Tutor specializes in development assistance supported by type holes.
 
-% The ChameleonIDE error reporting The error message of ChameleonIDE closely
-% follows the argumentation structure outlined by Titus. Basic mode represent a
-% simple structure argument, where the claim "The expression e can have two
-% conflicting types" and the possible type 1 and two serves as grounds. When
-% inspecting each possible types, the type judgements are treated as claim, and it
-% is supported by the grounds: "Inferred from the orange/blue highlights on the
-% left side". On the other hand, the similar type error message in GHC "Couldn't
-% match expect type T1 with actual type T2" is "a ground masquerading as a clam",
-% Barik commented. The balanced mode and advanced mode serve as a way to elaborate
-% arguments from simple arguments into extended arguments.
-
-% Another related topic is type-explanation~\cite{yang1999explaining, jun2002explaining}.
+% A few studies \cite{yang1999explaining, jun2002explaining} focus on explain type error in particular.
 Yang~\cite{jun2002explaining} showed an alternative type inference system capable of producing a human-like text explanation for why expressions are assigned certain types.
@@
 However, when presented as text-based output, explanation systems have the potential to become verbose when types are complex or variable names are long.
-In \chameleon{}, we attempt to address this problem by showing one step of explanation at a time and referring to variables instead of spelling out the full name.
-
-Debugging using a GUI Interactive Development Environment has been the standard practice for a long time.
-Compared to a command line-based interface, a graphic user interface provides programming tools that can show information hierarchy, provide immediate feedback to changes, and display complex visualizations.
-Hazel Tutor \cite{potter_hazel_2020} is an interactive type-driven environment for the OCaml language.
-It can automatically fill type holes by suggesting template expressions (strategies as called by the authors) through a popup window.
-It also provides a cursor-based type inspector that allows programmers to query the types of different parts of the program.
-Whyline~\cite{ko_finding_2009} is a Java debugging system that allows a user to ask questions like "why does variable X have value Y." It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error.
-These debugging tools are important motivations for developing \chameleon{}.
-However, they focus on different contexts of the debugging process.
-Java Whyline mainly tackles the problem of unintended run-time behavior, while Hazel Tutor specializes in interactive actions supporting type holes.

### 2022-12-05 sections/walkthrough.tex
Fix latex errors
Recent changes
John's comments

@@
 programmer Maxine.
+
+\todo{A bit of redundant/repeated here c.f. prev section - if this is Usage example, again use motivating example in it?}
 
@@
         \centering
-        \Description{Maxine using basic mode}
         \includegraphics[width=\linewidth]{images/basic-mode-1.pdf}
@@
         \centering
-        \Description{A screenshot of using basic mode}
         \includegraphics[width=\linewidth]{images/basic-mode-2.pdf}
@@
         \centering
-        \Description{A screenshot of using basic mode}
         \includegraphics[width=\linewidth]{images/basic-mode-3.pdf}
@@
 earlier. After saving the file, \chameleon{} shows that there is a type error in the
-expression \texttt{sum} (Fig. \ref{fig:balance-mode-1}). However, this is not 
+expression \texttt{sum} (Fig. \ref{fig:balance-mode-1}). However, this is not   \todo{Add annotation to figure like prev ones i.e. (1) so can refer to in text without reader searching for it?}
 helpful because Maxine knows that the implementation of  \texttt{sum} is correct 
@@
         \centering
-        \Description{A screenshot of using balanced mode}
-        \includegraphics[width=\linewidth]images/{balanced-mode-1.pdf}
+        \includegraphics[width=\linewidth]{images/balanced-mode-1.pdf}
         \caption{
@@
    \centering
-        \Description{A screenshot of using balanced mode}
         \includegraphics[width=\linewidth]{images/balanced-mode-2.pdf}
@@
 contain the cause. It does not take long until Maxine realizes that the list 
-\texttt{evens} is not supplied to the \texttt{filter} function.
+\texttt{l} is not supplied to the \texttt{filter} function.
 
@@
         \includegraphics[width=\linewidth]{images/advanced-mode-1.pdf}
-        \Description{A screenshot of using advanced mode}
         \caption{
@@
         \includegraphics[width=\linewidth]{images/advanced-mode-2.pdf}
-        \Description{A screenshot of using advanced mode}
         \caption{
@@
         \includegraphics[width=\linewidth]{images/advanced-mode-3.pdf}
-        \Description{A screenshot of using advanced mode}
         \caption{

### 2022-12-05 sections/chameleon.tex
John's comments
A temporary save

@@
 The debugging interface is designed from the ground up; the type inference engine is a re-implementation of the original Chameleon with several novel improvements.
+
+\todo{I wonder about a high level overview diagram of tool / usage of tool, then the details of each component?}
+
+\todo{Can you use this example in Motivating Example sectoon (see my comment above) then use it to explain tool in this section?  Make sense?
+
+Is this example too simple...?}
 
@@
     \centering
-    \includegraphics[width=\linewidth]{images/single-error-location-example.png}
+    \includegraphics[width=\linewidth]{images/add-pair-example.pdf}
     \caption{
 With traditional type error messages, one expression is blamed for causing the type error. This expression may not always match the programmers' intention. 
+% \todo{this figure unreadable...}
     }

### 2022-12-05 sections/discussion.tex
A temporary save

@@
 It may produce effective results by guessing the possible un-typable expression based on the position it appears in the syntax tree and project structure or guessing the intended type signature by comparing the number of supporting slices in the source code.
-
-\subsection{Threats to validity}
-
-\todo{Put threats BEFORE Discussion and at end Eval section?}
-
-\todo{These are too terse/unclear - need expanding. Suggest classify as internal / external threats etc .  What about background and number of participants as threats??}
-
-One consequence of our recruiting approach is it is harder for previous participants to enter a later study.
-To minimize the lurking variable of previous experience, we use new code challenges in every study and conduct trial runs in every study to bring new participants up to speed.
-
-Conducting studies remotely and unsupervised is that intervening when users encounter usability issues mid-study is impossible.
-To ensure there is no major usability issue that could lower the quality of data collection, we conducted cognitive walkthroughs and sandbox pilots before running each study.

### 2022-12-05 sections/evaluation.tex
A temporary save

@@
 \end{figure}
+
+\subsection{Threats to validity}
+
+% \todo{Put threats BEFORE Discussion and at end Eval section?}
+
+\todo{These are too terse/unclear - need expanding. Suggest classify as internal / external threats etc .  What about background and number of participants as threats??}
+
+One consequence of our recruiting approach is it is harder for previous participants to enter a later study.
+To minimize the lurking variable of previous experience, we use new code challenges in every study and conduct trial runs in every study to bring new participants up to speed.
+
+Conducting studies remotely and unsupervised is that intervening when users encounter usability issues mid-study is impossible.
+To ensure there is no major usability issue that could lower the quality of data collection, we conducted cognitive walkthroughs and sandbox pilots before running each study.

### 2022-12-05 sections/introduction.tex
update from overleaf
A temporary save

--- 2022-12-03 sections/introduction.tex
@@
-Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}.
+% \todo{As above, would start with motivation of dynamically typed langs e.g. JavaScript, Pythjon popularity but issues this brings...
+% }
+
+Dynamically typed programming languages such as JavaScript and Python have risen in popularity in recent decades \cite{chatley_next_2019}.
+These languages have the advantage of enormous flexibility: they require no type declaration, variable types or object structures can be modified dynamically, and functions can deal with dynamic input using ad-hoc polymorphism and runtime reflection.
+However, studies show that dynamically typed languages show challenges in development productivity \cite{kleinschmager_static_2012}, code usability \cite{mayer_empirical_2012}, and code quality \cite{gao_type_2017, ray_large-scale_2017, meyerovich_empirical_2013}.
+They are often found to produce error-prone code \cite{chen_empirical_2020, wang_empirical_2015,xu_python_2016} and reuqire strong discipline to avoid pitfalls \cite{chen_empirical_2020}.
+For these reasons, many modern dynamically typed languages have introduced statically typing as part of the core language features in recent years.
+
+Functional programming languages have long enjoyed rigorous type systems and expressive type-level features.
+Techniques such as type inference and algebraic types have long been standard practice in functional languages such as ML and Haskell, and now fresh blood in imperative languages, such as Rust and typescript.
+Haskell is considered a "type-system laboratory" \cite{hudak_history_2007}.
+Various type system advancements were experimented with in Haskell and ended up in mainstream languages years or even decades after.
+Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak_history_2007}, and now can be found in most popular languages such as C\#~\cite{bill_wagner_constraints_2022}, Java~\cite{oracle_generic_2022}, and TypeScript~\cite{microsoft_documentation_2022}.
+
+The downside of programming in functional, statically typed language is its complexity, especially when encountering type errors.
+Understanding and fixing type errors are shown to be hard\cite{tirronen_understanding_2015}.
+In particular, they may point to locations that are distant from real error causes, expose errors in hard to comprehend, or provide misleading fixing suggestions.
+In addition, the tools equipped are generally ineffective in showing and explaining type errors\cite{wu_how_2017}.
+% Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}.
 In recent years, the programming community has seen a trend of migrating away from dynamically typed codebases due to the lack of maintainability and refactoring safety when systems reach large scale and complexity~\cite{chatley_next_2019}.
@@
 
-Programmers find challenges in adopting typed languages in their practice. They tend to be verbose, hard to learn and generate bad error messages. 
-One obstacle programmers face when using a statically typed language is understanding and resolving type errors~\cite{marceau_measuring_2011, tirronen_understanding_2015}, especially when dealing with modern type features such as polymorphic types and implicit typing.
+% \todo{There are a lot of minor grammer fix-ups needed throughout...}
+% Programmers find challenges in adopting typed languages in their practice. \todo{I think this statement is plain wrong myself... 
+%  Need strong references to justify if going to say this to SE audience!:}They tend to be verbose, hard to learn and generate bad error messages. 
+% One obstacle programmers face when using a statically typed language is understanding and resolving type errors~\cite{marceau_measuring_2011, tirronen_understanding_2015}, especially when dealing with modern type features such as polymorphic types and implicit typing.
 Studies show most statically typed languages produce unhelpful and even misleading type errors~\cite{}.
@@
 
-This paper introduces \chameleon{}, an interactive debugging tool to visualize the relevant context of a type error (where it happens or could have happened, and which parts of code cause it) and allow programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve that ambiguity.
-Most noticeably, features (Section \ref{sub:type-compare}) provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature.
-In addition, we integrated all the features into a unified debugging interface and provided functionality for programmers to enable and disable the features based on their preferences and debugging needs.
+This paper introduces \chameleon{}, an interactive debugging tool for Haskell.
+It can visualize the relevant context of a type error (where it happens or could have happened and which parts of code cause it).
+In addition, \chameleon{} allows programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve that ambiguity.
+The most noticeable features provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature.
+These features are integrated into a unified debugging environment and can be enabled or disabled separately based on the programmers' preferences and debugging needs.
 \chameleon{} is open-source and is available in both web~\cite{shuai_fu_chameleon_2022} and desktop versions.
-The current implementation of \chameleon{} targets the Haskell 2010 language standard.
+% \todo{Introduce Haskell earlier?
+Motivate WHY Haskell?} The current implementation of \chameleon{} targets the Haskell 2010 language standard.
 However, it is planned to extend it to support other programming languages using the same underlying ideas and techniques.
 Historically, Haskell has been a test bed for advanced language features, and it is common for features established in Haskell to be transferred to mainstream programming languages.
-Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak_history_2007}, and now can be found in most popular languages such as C\#~\cite{bill_wagner_constraints_2022}, Java~\cite{oracle_generic_2022}, and TypeScript~\cite{microsoft_documentation_2022}.
 
@@
 \begin{itemize}
-\item { We provide the design and implementation of the \chameleon{}, to visualize the relevant context of a type error and allow programmers to explore and verify the error locations in small chunks interactively.
-}
+\item We provide the design and implementation of the \chameleon{} to visualize the relevant context of a type error and allow programmers to explore and verify the error locations in small chunks interactively.
 \item {

### 2022-12-08 main.tex
A temporary save
More Tim's edits

--- 2022-12-03 main.tex
@@
 \usepackage{minted}
+\makeatletter
+\let\@float@c@listing\@caption
+\makeatother
 \usepackage{graphicx}
@@
 \newcommand{\tf}[1]{\textcolor{magenta}{\textsc{tf}: #1}}
+
+% Comment out second line to disable.
+\newcommand{\todo}[1]{}
+% \renewcommand{\todo}[1]{{\color{red} TODO: {#1}}}
 
@@
 \begin{abstract}
-In recent years, many programming communities have been undergoing a transition to writing strongly typed programs.
-To benefit from type-safe codebases, programmers must first overcome the challenge of resolving type errors where text messages can often be unhelpful and misleading.
-Modern programming tools do not concentrate on making type errors easy to understand.
-We demonstrate \chameleon{}, a type debugging tool capable of displaying the full context of a type error and allowing programmers to search for where the type error actually occurs based on their prior knowledge and to incrementally verify their assumptions.
-\chameleon{} was developed through a user-centered, iterative process with four participatory experiments.
-Through the experiments, we showed that programmers using \chameleon{} fix type errors faster than traditional text-based error messages.
+
+% \todo{I would start with saying untyped langs popular e.g. Python and the need to address problems this brings...}
+
+Dynamically-typed programming languages are popular in education and the software industry.
+While presenting a low barrier to entry, they suffer from run-time type errors and longer-term problems in code quality and maintainability.
+Statically typed languages, while showing strength in these aspects, lack in learnability and ease of use.
+In particular, fixing type errors poses challenges to both novice users and experts.
+Further, compiler type error messages are presented in a static way that is biased towards the first occurence of the error in the program code.
+To help users resolve such type errors we introduce \chameleon{}, a type debugging tool that presents type errors to the user in an unbiased way, allowing them to explore the full context of where the errors could occur.
+Programmers can interactively verify the steps of reasoning against their intention.
+Through three studies involving real programmers, we showed that  \chameleon{} is more effective in fixing type errors than traditional text-based error messages.
 This difference is more significant in harder tasks.
-Further, programmers actively using \chameleon{} interactive features are shown to be more efficient in fixing type errors than simply reading the type error output.
+Further, programmers actively using \chameleon{}'s interactive features are shown to be more efficient in fixing type errors than passively reading the type error output.
 \end{abstract}
@@
 \begin{IEEEkeywords}
-component, formatting, style, styling, insert
+types, type errors, debugging, visualization, exploration
 \end{IEEEkeywords}
@@
 \input{sections/introduction.tex}
-\input{sections/related-work.tex}
 \input{sections/chameleon.tex}
@@
 \input{sections/discussion.tex}
+\input{sections/related-work.tex}
 \input{sections/conclusion.tex}

### 2022-12-08 sections/chameleon.tex
More Tim's edits

@@
+\section{Motivation}
+The design requirements of \chameleon{} are motivated by limitations of traditional type errors, as documented in a number of studies (e.g.~\cite{yang_improved_2000}), but which we illustrate here with a few motivating examples.
+%These limitations are sourced from the authors' frustration in teaching and writing Haskell.
+This list of shortcomings of traditional type errors is also strengthened by multiple studies pursuing improvement\cite{yang_improved_2000}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth,trim=0mm 35mm 0mm 0mm]{images/add-pair-example.pdf}
+    \caption{
+With traditional type error messages, one expression is blamed for causing the type error. This expression may not always match the programmers' intention. 
+    }
+    \label{fig:motivation-example}
+\end{figure}
+\subsubsection{\textbf{Traditional type errors show only limited location}}
+Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors.
+To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}''  In the example of Fig \ref{fig:motivation-example}, the type errors can be fixed in multiple locations.
+For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or  replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}.
+In the type error message, only the \texttt{addPair} expression on line 4 was blamed.
+In this small example the whole context is visible, but the problem is worse for the programmer in large programs where the lines of code contributing to the type error are far apart in the source code.
+
+\subsubsection{\textbf{Traditional type errors are biased}}
+In the example of Fig \ref{fig:motivation-example}, the error message arbitrarily focuses the users' attention on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect.
+%Technically, this results from the side effect of different unification orders of the internal type-checking technique and has no bearing on what the programmers think and expect.
+Another related form of bias is that traditional type errors are often framed as conflicts between \texttt{Expected type} and \texttt{Actual type}.
+This framing is standard practice in most typed languages.
+However, what is the \texttt{expected} and what is \texttt{actual} are a side effect of different unification orders rather than the intention of the programmer.
+In both of these cases, the error message may lead programmers to falsely believe the validity of parts of code and wrongly accuse others.
+
+\subsubsection{\textbf{Traditional type errors give poor explanations}}
+When the compiler rejects a program, the internal state of type checking is the result of complex computation.
+But the details of this process are hard to explain to users and usually not reported by compilers.
+For the typical type error shown in Fig.~\ref{fig:motivation-example}, the evidence for the type error is gathered from the previous two declarations.
+These have to be rediscovered by the programmer, using less rigourous methods.
+
+\subsection{Design Goals of \chameleon{}}
+Based on the limitations of traditional type errors, we give the following design requirements for \chameleon{}:
+
+\noindent\textbf{Show} all the possible locations where the type error happened or could have happened.
+
+\noindent\textbf{Explain} type errors avoiding jargon and internal constructs of the type checker.
+
+\noindent\textbf{Do not presume} which expression is to blame for the type error based on the order of computation or which possible type for an expression is 'actual' or 'expected'
+
+\end{itemize}
+
 \section{Chameleon IDE} \label{chameleon}
+\begin{figure*}
+    \centering
+    \includegraphics[width=\textwidth, trim=0mm 10mm 0mm 0mm]{images/atonomy.pdf}
+    \caption{
+        \textbf{The anatomy of \chameleon{}.}
+        %\chameleon{} consists of two panes: an editor pane on the left and a
+        %debugging pane on the right. 
+        The editor pane (left) is similar to a
+        traditional code editor. Fragments of source code may have a highlight
+        color (A). Additionally, an explanation layer (B) displays in the editor
+        window, if deduction steps are enabled. The debugging window contains three blocks. 
+        An error statement and configuration block, containing
+        An error statement (D), and, optionally,
+        a list of candidate expression cards (E), a list of deduction steps (F), 
+        and a control bar (G) to increment/decrement deduction step. 
+        The conflicting types
+        block shows two alternative types (H) inferred by the type inference
+        engine. A relevant type information block to show additional information (I)
+        that may help understand type errors.
+    }
+    \label{fig:anatomy}
+\end{figure*}
 
-\chameleon{} comprises two parts: a type inference engine and an original debugging interface.
-This separate architecture allows us to 1) adapt \chameleon{} into multiple front ends, such as IDE extensions, desktop applications, and web applications, and 2) reuse the debugging interface for different programming languages by replacing the back-end type inference engine.
-The debugging interface is designed from the ground up; the type inference engine is a re-implementation of the original Chameleon with several novel improvements.
+\chameleon{} comprises two parts: a type inference engine and a novel interactive debugging interface.
+This separation of concerns allows us to (1) adapt \chameleon{} into multiple front ends, such as IDE extensions, desktop applications, and web applications; and (2) reuse the debugging interface for different programming languages by replacing the back-end type inference engine.
+The debugging interface is designed from the ground up; the type inference engine is a re-implementation of the original Chameleon with several novel improvements, as described in Section \ref{sec:typeinferenceengine}.
 
@@
 
+\subsection{The Debugging Interface}
+
+The \chameleon{} debugging interface provides three main features to visualize the relation between error statement and locations in code and to explore the explanation of each error location given by the type inference engine.
+%Further, programmers  can turn on and off features to suit their preferences and debugging needs.
+
+\paragraph{Type compare tool} \label{sub:type-compare}
+% In designing \chameleon{}, we want to argue against the practice of reporting type errors as the conflicts between \texttt{Expected type} and \texttt{Actual type}.
+Technically, the two alternatives are merely the side effect of different unification orders of the internal type-checking technique and have no bearing on what the programmers think and expect.
+
+The type compare tool shows conflicting types in different colors, each associated with one or more error locations highlighted in a matching colour (Fig.~\ref{fig:compare}).  
+%The type error must be fixed by modifying at least one of these locations. 
+%Error locations are highlighted with the matching color of the conflicting types. The type compare tool is useful to quickly bisect the type error. 
+If the programmers know the expression's intended type (they usually do), they will be able to eliminate half of the possible locations. 
+%To make this bisecting effect more pronounced, a user can hover over one of the possible types and only show the relevant locations that contribute to that type. 
+A hover interaction over one of the possible types facilitates such bisection, causing only the relevant locations that contribute to that type to be highlighted. 
+%This is a convenient way to put the error \emph{under the spotlight}.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/intro-compare.pdf}
+    \caption{
+        \textbf{\chameleon{} minimal output (with candidate expression cards and deduction step disabled)}.
+\chameleon{} identified the conflicting types for the function \texttt{a} and associated the relevant locations with each type.
+In addition, \chameleon{} leaves out the right-hand side of the declaration '\texttt{0}' and '\texttt{1}' because they are unrelated.
+}
+    \label{fig:compare}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/intro-candidate.pdf}
+    \caption{
+        \textbf{\chameleon{} with only candidate expression cards enabled.}
+        Indicates the type error can
+        occur in the definition of \texttt{x} or \texttt{y}.
+    }
+    \label{fig:expression}
+\end{figure}
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/intro-deduction.pdf}
+    \caption{
+        \textbf{\chameleon{} with both candidate expression cards and deduction steps
+        enabled.}
+        \chameleon{} explains the type error in four steps.
+In the screenshot, the active step is step 2, where \chameleon{} shows that the expression \texttt{x} and \texttt{y} should have the same type.
+    }
+    \label{fig:deduction}
+\end{figure}
+
+\paragraph{Candidate Expression Cards}  \label{sub:candidate-expression}
+
+% We propose candidate expression cards to make it clear to the user that there are multiple ways to resolve a type conflict.
+By contrast,  standard compiler error messages such as those of GHC arbitrarily focus the users' attention on only one candidate expression, e.g.\ ``\textit{Couldn't match expected type ‘Char’ with actual type ‘Bool’ in expression x}" where x is the candidate expression.
+In practice, this candidate expression often does not match programmers' intention (Fig.
+\ref{fig:single-candidate}).
+A candidate expression is an expression that can be inferred to have two conflicting types. 
+When a type error is detected, \chameleon{} provides a list of all candidate expressions and programmers are free to choose the problem to resolve by clicking on one candidate expression card.
+In the example shown in Fig.
+\ref{fig:expression}, \texttt{x} and \texttt{y} are both candidate expressions.
+Fixing either type error will make both expressions well-typed.
+
+Selecting a candidate expression card is a way for programmers to communicate to the type checker the knowledge that the other candidate expressions are in fact well-typed.
+Once a card is selected, the information in the conflicting types panel changes according to the change of candidate expression.
+In the editor window, some error locations change highlight colors to reflect the updated knowledge of the program.
+
+This feature can be effective in resolving the problem that traditional tools tend to report errors that happened in source code from third-party libraries.
+With candidate expression cards, a programmer can freely change the context of the type error to the user-defined variables and functions to gain a better understanding of why her own code does not match the library code instead of the other direction.
+
+Programmers interact with the candidate expression cards by clicking on them to commit the change of meaning or hovering on them to peek at the effect of the change; the effect is reverted once the cursor moves away.
+
+\paragraph{Deduction steps}  \label{sub:deduction-steps}
+
+% Deduction steps are motivated by the lack of explaining why the program has type errors.
+When the compiler rejects a program, it dumps the internal state of type checking.
+This information may be a result of complex computing but this process is not reported by compilers.
+For a typical type error shown in Fig.
+\ref{fig:ghc-error-example}, clearly, the evidence for the type error is gathered from the previous two declarations.
+These have to be rediscovered by the programmers again, using harder and less-sound methods.
+
+% \begin{figure}
+%     \centering
+%     \includegraphics[width=\linewidth]{images/ghc-error-example.png}
+%     \caption{
+%         }
+%     \label{fig:ghc-error-example}
+% \end{figure}
+
+Deduction steps allow programmers to explore all the error locations one at a time (Fig.
+\ref{fig:deduction}).
+Steps are shown as a list of sequentially  numbered circular buttons (step buttons) and an explanation layer in the editor window.
+In the explanation layer, two highlighted error locations are outlined, and a line is drawn to connect these two locations to a human-readable text explanation of
+their semantic connection.
+Programmers are free to activate any step.
+The active step is shown as the green button.
+When activating a step, some highlights switch color.
+The message in the explanation layer changes as well.
+A program in Fig.
+\ref{fig:deduction} generates a list of steps shown in Fig.
+\ref{fig:step-interface} left.
+
+Programmers can use mouse and keyboard shortcuts to increment or decrement the step number or jump to any step.
+Typically, a programmer resolves  type errors by navigating through all the deduction steps and verifying whether each explanation aligns with the intended meaning of the program.
+Eventually, she will find a step that does not match her intention, and the type error can be fixed by modifying one of the two outlined error locations.
+
+\begin{figure}
+    \centering
+    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/step-interface.pdf}
+    \caption{
+Deduction steps if they are shown all at once (left).
+In practice, steps are shown one at a time.
+Programmers increment or decrement the step number using the step control bar (Fig.
+\ref{fig:anatomy}-G) or by directly clicking on a step button (Fig.
+\ref{fig:anatomy}-F).
+To increment or decrement the deduction step can be intuitively thought of as moving the position of the \textit{split point} where the blue and orange highlights divide (right).
+        }
+    \label{fig:step-interface}
+\end{figure}
+
+Internally, deduction steps are different ways to group the error locations into two different highlighted colors. 
+If all the highlights are placed in a list-like data structure, each increment of the step numbers moves the end of one color and the start of the other forward (Fig.~\ref{fig:step-interface} right).
+
+\paragraph{Multiple Modes}
+
+Nielson pointed out  that the two most important issues in designing for usability are understanding the users' tasks and the differences in users \cite{jakob_nielsen_usability_1993}.
+From analyzing how users use \chameleon{}, we realized that the ideal debugging interface should adapt to the specific programmer and programming task.
+It is common that a programmer wants the debugger to just ``show the answer", and it is also possible a programmer wants to dive deeper into the problem domain and search for the optimal solution.
+To accommodate the need to customize the level of information density and granularity of control, \chameleon{} provides three modes: basic, balanced, and advanced.
+Programmers can switch between modes by clicking on the mode switching toggles (Fig.
+\ref{fig:anatomy}-C).
+The features accessible from different modes are summarised in the following table:%shown in table~\ref{tab:chameleon-features}.
+
+% \begin{table}
+%     \centering
+\begin{scriptsize}
+\begin{center}
+    \begin{tabular}{ l l  }
+     \textit{Mode} & \textit{Features} \\ \hline
+     Basic Mode & Type Compare Tool \\ \hline
+     Balanced mode & Type Compare Tool \\
+     & Candidate Expression Card \\  \hline
+     Advanced mode & Type Compare Tool \\
+     & Candidate Expression Card \\
+     & Deduction Steps \\
+    \end{tabular}
+    \end{center}
+\end{scriptsize}
+%     \caption{\chameleon{} modes and features}
+%     \label {tab:chameleon-features}
+% \end{table}
+
 \subsection{The Type Inference Engine}
@@
 
-Chameleon was originally a command-line tool developed in the early 2000s to improve type error reporting and introduced a design for interactive type debugging for the Haskell programming language.
+Chameleon was originally a command-line tool developed in the early 2000s to improve type error reporting %and introduced a design for interactive type debugging 
+for the Haskell programming language.
 Unlike traditional type errors produced by the Glasgow Haskell Compiler (GHC)~\cite{ben_gamari_home_2022}, which uses a Hindley–Milner type inference system, Chameleon infers types using constraint solving.
@@
 From this, Chameleon can report a list of locations, using the labels of constraints that are in the MUS.
-Stuckey et al.~\cite{stuckey_interactive_2003} showed that program locations linked to the constraints from a MUS are all relevant to the type error and must include the cause of the error.
+Stuckey et al.~\cite{stuckey_interactive_2003} showed that program locations linked to the constraints from an MUS are all relevant to the type error and must include the cause of the error.
 In this paper, we will refer to these locations as \textit{error locations}.
 
-Despite successfully borrowing the underlying ideas, we could not reuse the implementation of Chameleon since the project language standard and libraries used were out of date.
-In addition to what Chameleon can do, a few new type inference features were added in the \chameleon{} implementation.
+Despite successfully borrowing the underlying ideas, we could not reuse the original implementation of Chameleon since the project language standard and libraries used were out of date. 
+Our \chameleon{} implementation extends the original Chameleon approach in a number of ways.
+%In addition to what Chameleon can do, a few new type inference features were added in the \chameleon{} implementation.
 
@@
 The types reconstructed from Chameleon may be \texttt{(a, Int)} and \texttt{(a, String)}.
-Although the recovered types are theoretically correct, they introduce a new type variable \texttt{a} making the error message harder to understand.
+Although the recovered types are consistent, they introduce a new type variable \texttt{a} making the error message harder to understand.
 To solve this issue in \chameleon{}, we first expand the MUS to a maximally satisfiable subset.
@@
 This process stops when no more constraints can be added.
-The resulting subset, while not helpful to inform error locations, will produce the most concrete conflicting types.
-This difference can be seen in figure~\ref{fig:compare-to-original}.
+The resulting subset, while not helpful to inform error locations, will produce the most concrete conflicting types, see Fig.~\ref{fig:compare-to-original}.
 
-\begin{figure}[h]
+\begin{figure}
     \centering
-    \includegraphics[width=\linewidth]{images/compare-to-original.pdf}
+    \includegraphics[width=\linewidth, trim=0mm 6mm 0mm 0mm]{images/compare-to-original.pdf}
     \caption{
@@
 In addition, \chameleon{} provides support for type explanation.
-Similar to the type explanation system ~\cite{jun_explaining_2002} from Yang,  \chameleon{} is able to produce a human-readable explanation, but for type errors.
-This is achieved by expanding the metadata fields of each constraint to contain the information needed for type explanation.
+Similar to the type explanation system  in ~\cite{jun_explaining_2002},  \chameleon{} is able to produce a human-readable explanation, but for type errors.
+This is achieved by annotating nodes in the abstract syntax tree with constraints and the type of rules applied.
+We generate an inference history from constraints and accompanying annotations.
 
@@
 \end{minted}
-\caption{A simple program that is ill-typed. This program generates two constraints from line 1 and one constraint from line 2.}
+\vspace{-3mm}
+\caption{A simple program that is ill-typed. It generates two constraints from line 1 and one constraint from line 2. }
 \label{listing:1}
 \end{listing}
-For instance, the program in the listing \ref{listing:1} \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition).
+For instance, for the program in the listing \ref{listing:1}, \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition).
 Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed.
@@
 This low-level explanation is used in a high-level user interface described in section \ref{sub:deduction-steps} to help programmers reason about the validity of error locations.
-
-\begin{figure*}
-    \centering
-    \includegraphics[width=\textwidth]{images/atonomy.pdf}
-    \caption{
-        \textbf{The anatomy of \chameleon{}.}
-        \chameleon{} consists of two panes: an editor pane on the left and a
-        debugging pane on the right. The editor pane is similar to a
-        traditional code editor. Fragments of source code may have a highlight
-        color (A). Additionally, an explanation layer (B) displays in the editor
-        window, if deduction steps are enabled. The debugging window contains three blocks. 
-        An error statement and configuration block, containing
-        An error statement (D), and, optionally,
-        a list of candidate expression cards (E), a list of deduction steps (F), 
-        and a control bar (G) to increment/decrement deduction step. 
-        The conflicting types
-        block shows two alternative types (H) inferred by the type inference
-        engine. A relevant type information block to show additional information (I)
-        that may help understand type errors.
-    }
-    \label{fig:anatomy}
-\end{figure*}
-
-\subsection{The Debugging Interface}
-
-The \chameleon{} debugging interface provides three main features to visualize the relation between an error statement and the error locations in the source code and to explore the explanation of each error location given by the type inference engine.
-Further, programmers  can turn on and off features to suit their preferences and debugging needs.
-
-\paragraph{Type compare tool} \label{sub:type-compare}
-In designing \chameleon{}, we want to argue against the practice of reporting type errors as the conflicts between \texttt{Expected type} and \texttt{Actual type}.
-Technically, the two alternatives are merely the side effect of different unification orders of the internal type-checking technique and have no bearing on what the programmers think and expect.
-
-The type compare tool shows the two conflicting types in different colors, each associated with one or more error locations (Fig.
-\ref{fig:compare}).
-The type error must be fixed by modifying at least one of these locations.
-The error locations are highlighted with the matching color of the conflicting types.
-The type compare tool is useful to quickly bisect the type error.
-If the programmers know the expression's intended type (they usually do), they will be able to eliminate half of the possible locations.
-To make the bisecting effect more pronounced, a user can hover over one of the possible types and only show the relevant locations that
-contribute to that type. This is a convenient way to put the error \texttt{under the spotlight}.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\linewidth]{images/intro-compare.pdf}
-    \caption{
-        \textbf{\chameleon{} minimal output (with candidate expression cards and deduction step disabled)}.
-\chameleon{} identified the conflicting types for the function \texttt{a} and associated the relevant locations with each type.
-In addition, \chameleon{} leaves out the right-hand side of the declaration '\texttt{0}' and '\texttt{1}' because they are unrelated.
-}
-    \label{fig:compare}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{images/intro-candidate.pdf}
-    \caption{
-        \textbf{\chameleon{} with only candidate expression cards enabled.}
-        \chameleon{} shows that the type error can
-        occur in the definition of '\texttt{x}' or '\texttt{y}'.
-    }
-    \label{fig:expression}
-\end{figure}
-
-\begin{figure}
-    \centering
-    \includegraphics[width=\linewidth]{images/intro-deduction.pdf}
-    \caption{
-        \textbf{\chameleon{} with both candidate expression cards and deduction steps
-        enabled.}
-        \chameleon{} explains the type error in four steps.
-In the screenshot, the active step is step 2, where \chameleon{} shows that the expression \texttt{x} and \texttt{y} should have the same type.
-    }
-    \label{fig:deduction}
-\end{figure}
-
-\paragraph{Candidate Expression Cards}  \label{sub:candidate-expression}
-
-A candidate expression is an expression that can be inferred to have two conflicting types. 
-We propose candidate expression cards to make it clear to the user that there are multiple ways to resolve a type conflict.
-By contrast,  standard compiler error messages such as those of GHC arbitrarily focus the users' attention on only one candidate expression, e.g.\ ``\textit{Couldn't match expected type ‘Char’ with actual type ‘Bool’ in expression x}" where x is the candidate expression.
-In practice, this candidate expression often does not match programmers' intention (Fig.
-\ref{fig:single-candidate}).
-
-When a type error is detected, \chameleon{} provides a list of all candidate expressions and programmers are free to choose the problem to resolve by clicking on one candidate expression card.
-In the example shown in Fig.
-\ref{fig:expression}, \texttt{x} and \texttt{y} are both candidate expressions.
-Indeed, fixing one type error will make both expressions well-typed.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\linewidth]{images/add-pair-example.pdf}
-    \caption{
-With traditional type error messages, one expression is blamed for causing the type error. This expression may not always match the programmers' intention. 
-% \todo{this figure unreadable...}
-    }
-    \label{fig:single-candidate}
-\end{figure}
-
-Selecting a candidate expression card is a way for programmers to communicate to the type checker the knowledge that the other candidate expressions are in fact well-typed.
-Once a card is selected, the information in the conflicting types panel changes according to the change of candidate expression.
-In the editor window, some error locations change highlight colors to reflect the updated knowledge of the program.
-
-This feature can be effective in resolving the problem that traditional tools tend to report errors that happened in source code from third-party libraries.
-With candidate expression cards, a programmer can freely change the context of the type error to the user-defined variables and functions to gain a better understanding of why her own code does not match the library code instead of the other direction.
-
-Programmers interact with the candidate expression cards by clicking on them to commit the change of meaning or hovering on them to peek at the effect of the change; the effect is reverted once the cursor moves away.
-
-\paragraph{Deduction steps}  \label{sub:deduction-steps}
-
-Deduction steps are motivated by the lack of explaining why the program has type errors.
-When the compiler rejects a program, it dumps the internal state of type checking.
-This information may be a result of complex computing but this process is not reported by compilers.
-For a typical type error shown in Fig.
-\ref{fig:ghc-error-example}, clearly, the evidence for the type error is gathered from the previous two declarations.
-These have to be rediscovered by the programmers again, using harder and less-sound methods.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\linewidth]{images/ghc-error-example.png}
-    \caption{
-        }
-    \label{fig:ghc-error-example}
-\end{figure}
-
-Deduction steps allow programmers to explore all the error locations one at a time (Fig.
-\ref{fig:deduction}).
-Steps are shown as a list of sequentially  numbered circular buttons (step buttons) and an explanation layer in the editor window.
-In the explanation layer, two highlighted error locations are outlined, and a line is drawn to connect these two locations to a human-readable text explanation of
-their semantic connection.
-Programmers are free to activate any step.
-The active step is shown as the green button.
-When activating a step, some highlights switch color.
-The message in the explanation layer changes as well.
-A program in Fig.
-\ref{fig:deduction} generates a list of steps shown in Fig.
-\ref{fig:step-interface} left.
-
-Programmers use their mouse and keyboard shortcuts to increment or decrement the step number or jump to any step.
-Typically, a programmer resolves  type errors by navigating through all the deduction steps and verifying whether each explanation aligns with the intended meaning of the program.
-Eventually, she will find a step that does not match her intention, and the type error can be fixed by modifying one of the two outlined error locations.
-
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\linewidth]{images/step-interface.pdf}
-    \caption{
-Deduction steps if they are shown all at once (left).
-In practice, steps are shown one at a time.
-Programmers increment or decrement the step number using the step control bar (Fig.
-\ref{fig:anatomy}-G) or by directly clicking on a step button (Fig.
-\ref{fig:anatomy}-F).
-To increment or decrement the deduction step can be intuitively thought of as moving the position of the \textit{split point} where the blue and orange highlights divide (right).
-        }
-    \label{fig:step-interface}
-\end{figure}
-
-Internally, deduction steps are different ways to group the error locations into two different highlighted colors.
-If all the highlights are placed in a list-like data structure, each increment of the step numbers moves the end of one color and the start of the other forward (Fig.
-\ref{fig:step-interface} right).
-
-\paragraph{Multiple Modes}
-
-Nielson pointed out \cite{jakob_nielsen_usability_1993}  that the two most important issues in designing for usability are understanding the users' tasks and the differences in users.
-From analyzing how users use \chameleon{}, we realized that the ideal debugging interface should adapt to the specific programmer and programming task.
-It is common that a programmer wants the debugger to just "show the answer", and it is also possible a programmer wants to dive deeper into the problem domain and search for the optimal solution.
-To accommodate the need to customize the level of information density and granularity of control, \chameleon{} provides three modes: basic, balanced, and advanced.
-Programmers can switch between modes by clicking on the mode switching toggles (Fig.
-\ref{fig:anatomy}-C).
-The features accessible from different modes are shown in table~\ref{tab:chameleon-features}.
-
-\begin{table}
-    \centering
-    \begin{tabular}{ l l  }
-     Mode & Features \\ \hline
-     Basic Mode & Type Compare Tool \\ \hline
-     Balanced mode & Type Compare Tool \\
-     & Candidate Expression Card \\  \hline
-     Advanced mode & Type Compare Tool \\
-     & Candidate Expression Card \\
-     & Deduction Steps \\
-    \end{tabular}
-    \caption{\chameleon{} modes and features}
-    \label {tab:chameleon-features}
-\end{table}

### 2022-12-08 sections/conclusion.tex
More Tim's edits

@@
 We present \chameleon{}, a type debugging tool for the Haskell programming language.
-\chameleon{} facilitates a constraint-based type inference engine that provides accurate and comprehensive type error reports.
+Its constraint-based type inference engine provides accurate and comprehensive type error reports.
 \chameleon{} also features an interactive debugging interface with three different debugging modes, enabling programmers to navigate deep type relations and triangulate conflicting statements in source code.
-Four experiments with human participants were conducted to evaluate the tool's design with programmers with varied experiences.
+Our studies evaluated the tool's design with programmers.
 We found that, particularly for more complex tasks, \chameleon{} helped programmers to fix type errors more quickly than traditional text-based error messages.

### 2022-12-08 sections/discussion.tex
More Tim's edits

@@
 
-\todo{Suggest try and strucuture this a bit more e.g. each para bold the key lesson @ start of para. Can you group into implications for research/implications for practice to make more actionable??}
+% \todo{Suggest try and strucuture this a bit more e.g. each para bold the key lesson @ start of para. Can you group into implications for research/implications for practice to make more actionable??}
 
@@
 This effect is shown more clearly when the task is not trivial.
-We found that programmers who actively use ChameleonIDE's interactive features (candidate expression cards and deduction steps) are more efficient in fixing type errors than simply reading the type error output.
+We found that programmers who actively use \chameleon{}'s interactive features (candidate expression cards and deduction steps) are more efficient in fixing type errors than passively reading the type error output.
 In this section, we will discuss a few interpretations of the results.
 
-From the results of user studies 1 and 2, we observed that the choice of debugging tool had little effect on how fast programmers solve type errors.
-Conversely, when facing more realistic problems (longer source code, error locations more scattered), programmers are more effective using \chameleon{}.
-We believe one factor of this effectiveness of \chameleon{} is to reduce the amount of reading time.
-Jbara \cite{jbara_how_2015} showed that reading source code is generally the initial step of solving programming problems and is done in several passes.
-Although traditional compiler error message tools initially show fewer locations, it is unreliable, meaning that programmers have to expand the reading span without clear guidance.
+\subsection{Effect on Reading Source Code}
+From the results of Study 1a, we observed that the choice of debugging tool had little effect on how fast programmers solve simple type errors.
+Conversely, when facing more realistic problems (longer source code, error locations more scattered) in study 1b, programmers are more effective using \chameleon{}.
+One explanation is that \chameleon{} reduces the amount of reading time by taking programmers more directly to the problem.
+Earlier studies \cite{jbara_how_2015, peitek_what_2020} showed that reading source code is generally the initial step of solving programming problems and is done in several passes.
+Although traditional compiler error message tools initially show fewer locations, these may be incomplete, meaning that programmers have to expand the reading span without clear guidance.
 In contrast, Chameleon shows more error locations initially.
@@
 
-From the results of user studies 3 and 4, we found that programmers who use interactive tool fix type errors faster than the ones who passively read the error output.
-This difference is stronger in harder tasks.
+\subsection{Forming Debugging Plans}
+From the results of Study 2, we found that programmers who use the interactive tool fix type errors faster than the ones who passively read the error output.
+This effect is stronger in harder tasks.
 We speculate that one factor of this result is that  \chameleon{} helps to develop debugging plans.
 We observed that when working with \chameleon{}, programmers form different debugging tactics to attack the problem.
-Among the high interactivity participants in user study 4, some programmers cycle through deduction steps as a guide to reading source code; some programmers navigate to both ends of the deduction steps where types are normally grounded and concrete.
-In contrast, minimal interactive participants generally form more similar plans, including carefully reading the program text and manually annotating expressions based on their understanding of the program.
+Among the \textit{high} interactivity participants in user study 4, some programmers cycle through deduction steps as a guide to reading source code; some programmers navigate to both ends of the deduction steps where types are normally grounded and concrete.
+In contrast, \textit{minimal} interactive participants generally form similar plans, including carefully reading the program text and manually annotating expressions based on their understanding of the program.
 
-\begin{listing}[!ht]
-\begin{minted}{haskell}
-f z     
-    | z == 3 = False
-    | z == '4' = True
-\end{minted}
-\caption{     In this simple program, \chameleon{} report the error happened in \texttt{f} and \texttt{z}.}
-\label{listing:2}
-\end{listing}
-
+\subsection{Externalise Intermediate Typing Inforamtion}
 We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information.
@@
 
-In our studies, \chameleon{} shows a few different designs for type error visualization and interaction.
+\begin{listing}
+\begin{minted}{haskell}
+f z     
+    | z == 3 = False
+    | z == '4' = True
+\end{minted}
+\caption{In this simple program, \chameleon{} reports the error happened in \texttt{f} and \texttt{z}.}
+\label{listing:2}
+\end{listing}
+% Implication for research
+% In our studies, \chameleon{} shows a few different designs for type error visualization and interaction.
 However, we are far from exhausting the search space.
@@
 
-On a similar note, the current implementation of \chameleon{} requires non-trivial adaptation for editors such as VS Code and IntelliJ due to the overlay explanation layer.
+% Integration with existing tools
+% The current implementation of \chameleon{} requires non-trivial adaptation for editors such as VS Code and IntelliJ due to the overlay explanation layer.
 This type of error visualization is non-standard in mainstream editors.
@@
 
-Another area for future study is to extend the evaluation process.
-In studies 1b and 2, we used harder tasks compared to study 1a.
-However, the difficulty level is nowhere as hard as a professional Haskell programmer may face in the production Haskell codebase.
-It would be interesting to see how \chameleon{}  faces type errors that span multiple files and packages.
-Another dimension of difficulty is higher-level extractions.
-Haskell programs get more tricky to the debugger when more abstractions are added to the codebase, such as Monads, Monad transformers, and Lenses.
-These topics were not included in our study to accommodate novice users.
-However, observing how expert Haskell programmers use \chameleon{} features to solve advanced problems will gain valuable insight into the true usability of the tool.
-
-Another direction to expand the work is in data collection.
-The online study limited the amount of qualitative data we could gather from the users.
-Although we handed out debriefing surveys after each study, it is no substitute for more in-depth qualitative study methods.
-Inviting users for speak-aloud walkthroughs and interviews are viable methods to unveil deeper usability issues about \chameleon{}.
-
-Our study focuses on the Haskell language for its popularity in the academic world.
+% Extend to other languages
+% Our study focuses on the Haskell language for its popularity in the academic world.
 However, the low quality of textual type errors is a problem not limited to Haskell.
@@
 
-Our design of \chameleon{}  emphasizes allowing programmers to update the type error based on their prior knowledge.
-It unfolds to answer three questions: which expression should be considered un-typeable?
+% Ranking candidate expressions and alternative types
+% Our design of \chameleon{}  emphasizes allowing programmers to update the type error based on their prior knowledge.
+It unfolds to answer three questions: which expression should be considered ill-typed?
 Between multiple possible types, which one should be considered intended?
 Among all the possible locations, which one should be considered the cause?
-\chameleon{} answers all three questions by leaving it to the programmers to decide.
-However, it is possible and likely that some programmers enjoy conclusive error messages instead of flexible ones provided by \chameleon{}.
+\chameleon{} answers all three questions by leaving them to the hands of the programmers.
+However, it is possible in situations, prescriptive error messages are preferred.
 Using heuristic methods may achieve the best of both worlds.

### 2022-12-08 sections/evaluation.tex
More Tim's edits

--- 2022-12-05 sections/evaluation.tex
@@
 
-Three studies were conducted to answer new research questions or rephrase existing ones.
-These studies are grouped into 2 phases, and each tested a different version of \chameleon{} with human participants.
-The number of participants and their Haskell experience, as well as key research questions and conclusions, were shown in figure \ref{fig:timeline}.
+We conducted three user studies, iteratively refining the \chameleon{} UI and evaluating several research questions as per Fig.~\ref{fig:timeline}. 
+%The number of participants and their Haskell experience, as well as key research questions and conclusions, were shown in figure \ref{fig:timeline}.
 \todo{Explain was a user study of the tool with practitioners?}
 
-\begin{figure*}
+\begin{figure}
     \centering
-    \includegraphics[width=\textwidth]{images/timeline.pdf}
+    \includegraphics[width=\columnwidth,trim=30mm 35mm 35mm 5mm]{images/timeline.pdf}
     \caption{The timeline of \chameleon{}  evaluation.}
     \label{fig:timeline}
-\end{figure*}
+\end{figure}
 
@@
 
-The recruiting channel we used was the social news aggregation website Reddit.
-Specifically, our user study advertisements were posted in the Haskell programming language community \textit{r/haskell} and a general "programming language" focused community \textit{r/programminglanguages}.
-Recruiting from social media allowed us to access a more diverse demographic that better represent the true population of Haskell programmers.
-The participation is fully anonymized.
-The detailed ethical implications of these experiments are reviewed and approved by the IRB of the institution of the authors.
+Participants were recruited via the Reddit \textit{r/haskell} and \textit{r/programminglanguages} communities. 
+%Recruiting from social media allowed us to access a more diverse demographic that better represent the true population of Haskell programmers. 
+Participation is fully anonymized; detailed ethical implications of these experiments are reviewed and approved by the IRB of the institution of the authors.
 
@@
 \subsubsection*{\textbf{Experiment setting}}
-The experiments took place remotely and unsupervised.
-Participants took the study online via a web browser and at the physical venue of their choosing.
+Experiments were conducted online and unsupervised. 
+%Participants took the study online via a web browser and at the physical venue of their choosing. 
 All user studies use a web-based debugging environment developed by the authors. 
-Conducting the studies online helped us avoid variation when performing tasks in unfamiliar places and using different setups.
+%Conducting the studies online helped us avoid variation when performing tasks in unfamiliar places and using different setups.
 
@@
 After consent, participants received interactive training on the tool interface and interactive features.
-Participants were also shown a cheat sheet summarizing the key functionality of the interface.
-Participants had access to the cheat sheet at all times during the study.
+Participants were also shown a cheat sheet summarizing the key functionality of the interface, and had access to the cheat sheet at all times during the study.
 Participants were given 4 trial runs (2 for each setting) before the data collection started.
-
-All the user studies used a within-subject design to evaluate the effectiveness of different tools or feature sets while counterbalancing the difference in programming proficiency between participants.
+All the studies used a within-subject design to evaluate the effectiveness of different tools or feature sets while counterbalancing the difference in programming proficiency between participants.
 In each study, participants were required to complete a series of programming tasks (8 for studies 1a and 1b, 9 for study 2).
 At each task, a participant receives a single Haskell file that contains one or more type errors.
-The participant was asked to make the code type check with the help of the given tool.
+The participant was asked to correct the code with the help of the given tool.
 
@@
 To not introduce a barrier to completing the study, every task can be skipped if the participant made three failed attempts or is stuck for over 1 minute on the task.
-
-After completing all the tasks, participants are prompted to complete a debriefing survey.
-The survey questions include their Haskell experience and feedback on the tools and feature set participants used during the study.
+After completing all tasks, participants are prompted to complete a debriefing survey. The survey questions include their Haskell experience and feedback on the tools.
 
@@
 This provides us with a "poor man's video recording" to identify usability issues in the study and to recognize general patterns.
-The reason we found this format inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.
-
-One use of the session recording is for outlier trimming.
+%The reason we found this format inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.
+%One use of the session recording is for outlier trimming.
 We identify participants who left the computer for an extended period of time.
@@
 \subsubsection{\textbf{\chameleon{} 1}}  
-\chameleon{} 1 is an earlier iteration of \chameleon{} development.
-This version features the type inference engine that recovers most concrete types after type errors occur and a minimal set of debugging features.
+An earlier version of the UI than that depicted in Figs.~(2-13), it featured the type inference engine that recovers most concrete types after type errors occur and a minimal set of debugging features.
 Key features in \chameleon{} 1 include showing two (or more) alternative types, showing all possible error locations, dividing possible error locations into groups based on alternative types, and concrete type restoration.
@@
 We choose GHC compiler error messages as the baseline because it is the canonical tool for working with type errors in Haskell.
-Although high-level tools like Haskell Language Server exist, they generally relay the GHC error messages verbatim.
-The procedure of these studies is shown in figure \ref{fig:procedure-1}.
+%Although high-level tools like Haskell Language Server exist, they generally relay the GHC error messages verbatim. 
+% The procedure of these studies is shown in figure \ref{fig:procedure-1}.
 
@@
 
-The  studies (1a \& 1b) are designed to answer the research question:
-\begin{itemize}
-    \item \textbf{RQ1.} Do programmers solve type errors faster with \chameleon{} than GHC compiler error messages?
-\end{itemize}
+Studies (1a \& 1b) address the research question:
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\linewidth]{images/procedure-1.pdf}
-    \caption{\todo{Does this figure really help at all?}}
-    \label{fig:procedure-1}
-\end{figure}
+\noindent\textbf{RQ1.} \textit{Do programmers solve type errors faster with \chameleon{} than GHC compiler error messages?}
+
+% \begin{figure}[h]
+%     \centering
+%     \includegraphics[width=\linewidth]{images/procedure-1.pdf}
+%     \caption{\todo{Does this figure really help at all?}}
+%     \label{fig:procedure-1}
+% \end{figure}
 
@@
 
-\begin{figure}[h]
+\begin{figure}
     \centering
-    \includegraphics[width=\linewidth]{images/user-study-1a.pdf}
-    \caption{Time to complete each task in user study 1a with 95\% confidence interval}
+    \includegraphics[width=0.8\linewidth,trim=0mm 12mm 0mm 35mm,clip]{images/user-study-1a.pdf}
+    \caption{Study 1a task completion time (secs.) with 95\% confidence interval.}
     \label{fig:analysis-1a}
@@
 
-\begin{figure}[h]
+\begin{figure}
     \centering
-    \includegraphics[width=\linewidth]{images/user-study-1b.pdf}
-    \caption{Time to complete each task in user study 1b with 95\% confidence interval}
+    \includegraphics[width=0.8\linewidth,trim=0mm 15mm 0mm 35mm,clip]{images/user-study-1b.pdf}
+    \caption{Study 1b task completion time (secs.) with 95\% confidence interval.}
     \label{fig:analysis-1b}
@@
 
-From the data collected during study 1a (\ref{fig:analysis-1a}), we were not able to identify which group performed better.
-For the trivial challenges we set for users, the individual differences are generally more significant than differences between treatments.
-One interesting observation is task 8, where the \chameleon{} group outperformed the GHC group more consistently (p-value = 0.002456, Wilcoxin signed test).
-The discrepancy is thought to be related to the nature of Task 8.
-It has a lengthier source file and involves more language features (abstract data types and high-level functions).
+The data collected during study 1a, Fig.~\ref{fig:analysis-1a} does not show significant differences across Tasks 1-7.
+In hindsight, these tasks were trivial challenges for most users and the individual differences of participants are generally more significant than differences between treatments.
+However, one interesting observation is task 8, where the \chameleon{} group outperformed the GHC group (p-value = 0.002456, Wilcoxin signed test).
+We attribute this significant difference to the hardness of Task 8.
+The source file is longer and involves more language features (abstract data types and high-level functions).
 GHC struggles to produce a relevant error message for this type of error.
-We hypothesized that we may observe a more significant difference using tasks with lengthier and more realistic source code.
-This hypothesis is also supported by the most common feedback (7 out of 35) the study 1a was that the tasks were too trivial to invite meaningful evaluation.
-One participant said, "Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."
+From this result, we hypothesized that we may observe a more significant difference using tasks with lengthier and more realistic source code. 
+This hypothesis is also supported by the most common feedback (7 out of 35) that the tasks were too trivial to invite meaningful evaluation.
+One participant said, ``Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."
 
-In study 1b, the \chameleon{} group solved the type error faster (p =0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1.
-For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity and ineffective training.
-For all the other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program.
+Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group were faster (p =0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1.
+For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity.
+For all other tasks, from the video recordings, we saw many \chameleon{} users confidently skip reading unrelated chunks of code, while GHC users generally read through the whole program.
 In harder problems and messier code, we notice programmers start to report the benefits of \chameleon{}.
-"It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant.
-"I think \chameleon{}  does a much better job than GHC's error messages.
+``It's most useful feature that I noticed was that it points out the locations of both conflicting uses; GHC often makes it difficult to figure out how it's coming to a conclusion about a type." reported one participant.
+``I think \chameleon{}  does a much better job than GHC's error messages.
 I like that it shows the sources for the type judgments.
@@
 \subsubsection{\textbf{\chameleon{} 2}}  \label{sub:us4}
+Based on observations of Study 1 we introduced several new features to \chameleon{}, eventually resulting in the UI depicted in Figs.~(2-13).
+Advanced debugging features such as deduction steps and candidate expressions were available in this iteration.
+In addition, mode switching is enabled by default.
+A few other user interfaces were designed and prototyped between the development of \chameleon{} 1 and \chameleon{} 2.
+Study 2 therefore addresses the research question:
 
-\chameleon{} 2 is latest version of \chameleon{}.
-Advanced debugging features such as deduction steps and candidate expressions were available in this iteration.
-In addition, mode switching is enabled by default as well.
-A few other user interfaces were designed and prototyped between the development of \chameleon{} 1 and \chameleon{} 2.
-Study 2 was designed to evaluate \chameleon{} 2.
-We were trying to answer the research question: \textit{How do programmers use the interactive features in \chameleon{} 2?}.
-More specifically, we were trying to inquiry the following three sub-questions:
+\noindent\textbf{RQ2:} \textit{How do programmers use the interactive features in \chameleon{} 2?}.
+
+More specifically:
 \begin{itemize}
@@
 The order of the three modes in each cycle is counterbalanced among all participants.
-However, participants can switch to other modes at any time (figure \ref{fig:procedure-2}).
+However, participants can switch to other modes at any time.
 
-\begin{figure}[h]
-    \centering
-    \includegraphics[width=\linewidth]{images/procedure-2.pdf}
-    \caption{\todo{Does this figure really help at all?}}
-    \label{fig:procedure-2}
-\end{figure}
+% \begin{figure}[h]
+%     \centering
+%     \includegraphics[width=\linewidth]{images/procedure-2.pdf}
+%     \caption{\todo{Does this figure really help at all?}}
+%     \label{fig:procedure-2}
+% \end{figure}
 
@@
 
-The study of \chameleon{} 2 is an exploratory study.
+Study 2 is more exploratory in methodology than Study 1. 
 We were expecting to let programmers discover their own way of using the tool.
 In post hoc analysis of the collected log data, we were able to extrapolate some interesting patterns of how the tool was used.
-We share a few of our observations in this section.
+%We share a few of our observations in this section.
 
-The most striking feature of the data is users' tendencies to use the tool differ wildly.
-Some users used the feature extensively however some users completed the tasks without actively exploring the given information.
-Based on this discrepancy, we divided the users into three groups.
+The most striking feature of the data is that users tend to vary wildly in their use of the tool.
+Some users used the features extensively while others completed the tasks without actively exploring the given information.
+Based on this discrepancy, we divided the users into three groups:
 
-\begin{tabularx}{\linewidth}{ 
-  | >{\raggedright\arraybackslash}X 
-  | >{\raggedright\arraybackslash}X  | }
+\begin{small}
+\noindent\begin{tabularx}{\linewidth}{ 
+  | >{\hsize=.3\hsize}X 
+  | >{\hsize=.7\hsize \raggedright\arraybackslash}X  | }
 
@@
         Interaction level & Description \\ \hline
-        Minimal Interaction & Users completed the tasks by making changes in source code, type checking, and reading error messages. \\ \hline
-        Low Interaction & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
-        High Interaction & Users did everything from the Common feature group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards.
+        \textit{Minimal}  & Users completed the tasks by making changes in source code, type checking, and reading error messages. \\ \hline
+        \textit{Low}  & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
+        \textit{High}  & Users did everything from the Common feature group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards.
 \\ \hline
 \end{tabularx}
+\end{small}
 
-As shown in  figure \ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants.
-Participants with higher interaction levels generally performed better in the tasks, and the lowest level was worse (one-way ANOVA since we have three samples, p = 0.0368).
+As shown in  Fig.~\ref{fig:r4-analysis}, time to complete each task roughly relates to the interaction level of participants.
+Participants with higher interaction levels generally performed better, and the lowest interaction level was worse (one-way ANOVA since we have three samples, p = 0.0368).
 The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated.
-As discussed in user studies 1 and 2, it is likely related to the nature of these tasks.
+As with Study 1a and 1b, this difference is likely related to task difficulty.
 Tasks 4 and 6 are shorter than other tasks.
@@
 
-\begin{figure}[h]
+\begin{figure}
     \centering
-    \includegraphics[width=\linewidth]{images/user-study-2.pdf}
-    \caption{Time to complete each task in study 2 with 95\% confidence interval}
+    \includegraphics[width=\linewidth,trim=0mm 15mm 0mm 35mm,clip]{images/user-study-2.pdf}
+    \caption{Study 2 task completion time (secs.) with 95\% confidence intervals.}
     \label{fig:r4-analysis}
@@
 Second, when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
-\begin{figure}[h]
+\begin{figure}
     \centering
-    \includegraphics[width=\linewidth]{images/mode-switching.pdf}
-    \caption{
+    \includegraphics[width=0.7\linewidth,trim=0mm 5mm 0mm 5mm,clip]{images/mode-switching.pdf}
+    \caption{Study 2 mode switches by starting mode.  Users overwhelmingly switched to the more sophisticated interface mode.
         % The rows represent the finishing mode, columns starting modes.
@@
 \end{figure}
+
+\subsection{Limitations}
+
+One threat to the validity of the evaluation is the number of participants.
+Although for each study we received hundreds of online participants, the studies suffer from a high abandonment rate (especially study 1b).
+This was expected: the programming challenges are difficult and our volunteer participants are unremunerated.
+%To encourage completion, we included a cheat sheet containing a summary of the key functionality of the tools and necessary type signatures.
+A skip button is available for when participants are stuck.
+Because we recruited participants online and anonymized all the participants, it is possible for participants of a previous study to enter a later one.
+This creates variation in familiarity.
+We offset this by using new code challenges in every study and conducting trial runs before data collection to bring new participants up to speed.
+Conducting studies remotely and unsupervised there is no way to intervene when users encounter usability issues.
+To mitigate this we conducted cognitive walkthroughs and sandbox pilots before running each study.
+
+Future evaluation would benefit from using more realistic tasks.
+In studies 1b and 2, we used more challenging tasks compared to study 1a.
+However, the complexity is just short of what professional Haskell programmers may face in the production Haskell codebase.
+It would be interesting to see how \chameleon{} is used against type errors that span multiple files and packages and include confusing abstractions, like Monads, Monad transformers, and Lenses.
+
+% The online study limited the amount of qualitative data we could gather from the users.
+Although we handed out debriefing surveys after each study, it is no substitute for more in-depth qualitative study methods.
+Inviting users for speak-aloud walkthroughs and interviews are viable methods to unveil deeper usability issues about \chameleon{}.

### 2022-12-08 sections/introduction.tex
More Tim's edits

@@
 
-Dynamically typed programming languages such as JavaScript and Python have risen in popularity in recent decades \cite{chatley_next_2019}.
-These languages have the advantage of enormous flexibility: they require no type declaration, variable types or object structures can be modified dynamically, and functions can deal with dynamic input using ad-hoc polymorphism and runtime reflection.
-However, studies show that dynamically typed languages show challenges in development productivity \cite{kleinschmager_static_2012}, code usability \cite{mayer_empirical_2012}, and code quality \cite{gao_type_2017, ray_large-scale_2017, meyerovich_empirical_2013}.
-They are often found to produce error-prone code \cite{chen_empirical_2020, wang_empirical_2015,xu_python_2016} and reuqire strong discipline to avoid pitfalls \cite{chen_empirical_2020}.
-For these reasons, many modern dynamically typed languages have introduced statically typing as part of the core language features in recent years.
+Dynamically-typed programming languages such as JavaScript and Python have risen in popularity in recent decades \cite{chatley_next_2019}.
+These languages present a low barrier of entry, especially to beginner programmers: they require no type declaration, variable types or object structures can be modified dynamically, and functions can deal with dynamic input using ad-hoc polymorphism and runtime reflection.
+However, studies show that dynamically-typed languages negatively affect development productivity \cite{kleinschmager_static_2012}, code usability \cite{mayer_empirical_2012}, and code quality \cite{gao_type_2017, ray_large-scale_2017, meyerovich_empirical_2013}.
+They are often found to produce error-prone code \cite{chen_empirical_2020, wang_empirical_2015,xu_python_2016} and require strong programmer discipline to avoid pitfalls \cite{chen_empirical_2020}.
+For these reasons, many modern dynamically-typed languages have introduced static typing annotations as part of the core language features in recent years (e.g.\ \textit{TypeScript} and \textit{mypy}).
 
 Functional programming languages have long enjoyed rigorous type systems and expressive type-level features.
-Techniques such as type inference and algebraic types have long been standard practice in functional languages such as ML and Haskell, and now fresh blood in imperative languages, such as Rust and typescript.
-Haskell is considered a "type-system laboratory" \cite{hudak_history_2007}.
-Various type system advancements were experimented with in Haskell and ended up in mainstream languages years or even decades after.
+Techniques such as type inference and algebraic types have been standard practice for decades in functional languages such as ML and Haskell, and more recently in multi-paradigm languages, such as Rust and TypeScript.
+Haskell is considered a ``type-system laboratory" \cite{hudak_history_2007}.
+Various type system advances were introduced in Haskell and ended up in mainstream languages years or even decades after.
 Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak_history_2007}, and now can be found in most popular languages such as C\#~\cite{bill_wagner_constraints_2022}, Java~\cite{oracle_generic_2022}, and TypeScript~\cite{microsoft_documentation_2022}.
 
-The downside of programming in functional, statically typed language is its complexity, especially when encountering type errors.
-Understanding and fixing type errors are shown to be hard\cite{tirronen_understanding_2015}.
-In particular, they may point to locations that are distant from real error causes, expose errors in hard to comprehend, or provide misleading fixing suggestions.
-In addition, the tools equipped are generally ineffective in showing and explaining type errors\cite{wu_how_2017}.
+The challenge of programming in statically-typed languages is that type errors can sometimes be difficult to resolve~\cite{tirronen_understanding_2015}.
+In particular, they may point to locations that are distant from real error causes, expose errors in hard to comprehend language, or provide misleading fixing suggestions~\cite{wu_how_2017}.
 % Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}.
@@
 
-This paper introduces \chameleon{}, an interactive debugging tool for Haskell.
+This paper introduces \chameleon{}, an interactive type debugging tool for Haskell.
 It can visualize the relevant context of a type error (where it happens or could have happened and which parts of code cause it).
-In addition, \chameleon{} allows programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve that ambiguity.
+In addition, \chameleon{} allows programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve ambiguity.
 The most noticeable features provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature.
@@
 \item {
-    We report the result of four experiments designed to evaluate \chameleon{}.}
+    We report the results of four experiments designed to evaluate \chameleon{}.}
 \end{itemize}

### 2022-12-08 sections/related-work.tex
More Tim's edits

@@
 
-\todo{I would do a short Motivating Example section straight after intro showing example of untyped lang, problems it brings, and outline requirments for the proposed IDE to address it. Make sense??}
+% \todo{I would do a short Motivating Example section straight after intro showing example of untyped lang, problems it brings, and outline requirments for the proposed IDE to address it.
+Make sense??}
 
-\todo{Do yoou need Related Work here - or put @ end of paper just before Summary - I tend to do the later now UNLESS its REALLY needed by reader - IMO could put @ end and get to your work/contributions much quicker}
+% \todo{Do yoou need Related Work here - or put @ end of paper just before Summary - I tend to do the later now UNLESS its REALLY needed by reader - IMO could put @ end and get to your work/contributions much quicker}
 
 \subsection{Finding all type error locations}
-Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors.
-To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' Many studied \cite{stuckey_interactive_2003, haack_type_2004, pavlinovic_practical_2015, schilling_constraint-free_2012} the approach of finding all locations that contribute to a type error.
+Many have studied the approach of finding all locations that contribute to a type error (\cite{stuckey_interactive_2003, haack_type_2004, pavlinovic_practical_2015, schilling_constraint-free_2012}).
 Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error.
@@
 One weakness of compiler error messages, in general, is that they fail to explain the error in human language.
-As put in~\cite{barik_developers_2017}, "Error messages appear to take the form of natural language, yet are as difficult to read as source code."  A well-studied approach to producing better error explanations is through ECEM (Enhanced compiler error message).
+As put in~\cite{barik_developers_2017}, ``Error messages appear to take the form of natural language, yet are as difficult to read as source code."  A well-studied approach to producing better error explanations is through ECEM (Enhanced compiler error message).
 Through a series of mixed-method studies, Prather showed~\cite{prather_novices_2017} that ECEM has a positive result in understanding compiler errors.
-Decafe~\cite{becker_effective_2016} is a tool that can rephrase the Java compiler error messages into their enhanced version.
+Decafe~\cite{becker_effective_2016} is a tool that can rephrase Java compiler error messages into an enhanced version.
 In a study of over 200 CS1 students, Decaf was shown to reduce overall errors in their coding practices.
@@
 Most principles and suggestions are followed in \chameleon{} in constructing error statements.
-However, they were not targeting type errors alone but compiler errors (some even include run time errors) in general.
+However, these earlier studies were not targeting type errors alone but general compiler errors (some even include run time errors).
 The nuances of type errors, such as alternative typing, were not considered.
-Moreover, explanation systems are designed specifically for novice users.
+Moreover, these explanation systems were designed specifically for novice users.
 
-\subsection{Suggesting resolution}
-One direction to approach type error debugging is by suggesting solutions to fix the type error.
+% \subsection{Suggesting resolution}
+% One direction to approach type error debugging is by suggesting solutions to fix the type error.
 Sheng Chen's works~\cite{chen_counter-factual_2014, chen_efficient_2020, chen_improving_2022} on counterfactual typing are able to suggest changes to make the ill-typed program type-check.
-While this is a promising ground, the work is not evaluated with human participants.
 Lerner  studied an approach~\cite{lerner_searching_2007} of repeatedly modifying the AST (abstract syntax tree) and querying the compiler/type checker to know whether the new program is well-typed.
@@
 \subsection{Interactive Debugging}
-Interactive debugging tools have been studied using the command line interface.
-Chameleon provides a certain level of interactive debugging~\cite{stuckey_interactive_2003}.
-The interaction is used to query different parts of the program.
-A recent study \cite{tsushima_type_2021} provided an interactive type debugging system that restores type information based on the output from CFT-based type inference.
+The original Chameleon provides a certain level of ``interactive'' debugging through a command-line interface~\cite{stuckey_interactive_2003}, allowing the programmer to query different parts of the program.
+A recent study \cite{tsushima_type_2021} provided an interactive type debugging system that restores type information based on the output from CFT(Counter-factual typing)~\cite{chen_counter-factual_2014, chen_efficient_2020, chen_improving_2022} inference.
 Both tools allow programmers to change their assumptions about which expression is ill-typed.
 While both tools show a strong influence on theoretical development, they were not evaluated in realistic studies with human participants.
-In addition, command line-based debugging tools suffer the same limitation as other line-based tools, namely, 1 2 and 3.
+In addition, command line-based debugging tools suffer the same limitation as other line-based tools, namely, limited user interaction, limited graphic capability (ANSI text), and the requirement of remembering the command language.
 
-A graphic user interface (GUI) provides programming tools with immediate feedback on changes and the ability to show complex visualizations.
-Many tools dedicate to improving the debugging experience.
+A GUI provides programming tools with immediate feedback on changes and the ability to show complex visualizations.
+Many tools aim to improve the debugging experience.
 We list two.
 Hazel Tutor \cite{potter_hazel_2020} is an interactive type-driven environment for the OCaml language.
-It can automatically fill type holes by suggesting template expressions (strategies as called by the authors) through a popup window.
+It can automatically fill type holes by suggesting template expressions (called ``strategies'' by the authors) through a popup window.
 It also provides a cursor-based type inspector that allows programmers to query the types of different parts of the program.
-Whyline~\cite{ko_finding_2009} is a Java debugging system that allows a user to ask questions like "why does variable X have value Y." It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error.
+Whyline~\cite{ko_finding_2009} is a Java debugging system that allows a user to ask questions like ``why does variable X have value Y." It also allows users to interactively ask follow-up questions to gain further knowledge of the nature of an error.
 These debugging tools are important motivations for developing \chameleon{}.

### 2022-12-08 sections/walkthrough.tex
More Tim's edits

@@
 \section{Walkthrough} \label{sec:walkthrough}
-In this section, we introduce \chameleon{} by walking through examples of its
+In this section, we showcase \chameleon{} by walking through examples of its
 use. The examples are given from the perspective of a hypothetical Haskell 
@@
 
-\todo{A bit of redundant/repeated here c.f. prev section - if this is Usage example, again use motivating example in it?}
+% \todo{A bit of redundant/repeated here c.f. prev section - if this is Usage example, again use motivating example in it?}
 
@@
 Maxine writes a function to calculate the sum of a list of
-numbers, but \chameleon{} shows there is a type error (Fig. \ref{fig:basic-mode-1}). 
-After reading error reports, Maxine realized that the error possibly revolves 
+numbers, but \chameleon{} shows there is a type error (Fig.~\ref{fig:basic-mode-1}). 
+After reading the error reports, Maxine realizes that the error possibly revolves 
 around the expression \texttt{xs}.
 
-\begin{figure}[h]
+\begin{figure}
         \centering
-        \includegraphics[width=\linewidth]{images/basic-mode-1.pdf}
+        \includegraphics[width=\linewidth,trim=0mm 8mm 0mm 0mm]{images/basic-mode-1.pdf}
         \caption{
@@
 
-\begin{figure}[h]
-        \centering
-        \includegraphics[width=\linewidth]{images/basic-mode-2.pdf}
-        \caption{
-       Hovering on possible type 1 will limit the highlights 
-       in the editor to only the ones (in orange) that support \texttt{x::[a]}.
-        }
-        \label{fig:basic-mode-2}
-\end{figure}
+% \begin{figure}[h]
+%         \centering
+%         \includegraphics[width=\linewidth]{images/basic-mode-2.pdf}
+%         \caption{
+%        Hovering on possible type 1 will limit the highlights 
+%        in the editor to only the ones (in orange) that support \texttt{x::[a]}.
+%         }
+%         \label{fig:basic-mode-2}
+% \end{figure}
 
-\begin{figure}
-        \centering
-        \includegraphics[width=\linewidth]{images/basic-mode-3.pdf}
-        \caption{
-            Hovering on possible type 2 (Fig. \ref{fig:basic-mode-2}-3) will limit the highlights 
-            in the editor to only the ones (in blue) that support \texttt{x::Int}.
-        }
-        \label{fig:basic-mode-3}
-\end{figure}
+% \begin{figure}
+%         \centering
+%         \includegraphics[width=\linewidth]{images/basic-mode-3.pdf}
+%         \caption{
+%             Hovering on possible type 2 (Fig. \ref{fig:basic-mode-2}-3) will limit the highlights 
+%             in the editor to only the ones (in blue) that support \texttt{x::Int}.
+%         }
+%         \label{fig:basic-mode-3}
+% \end{figure}
 
 After scanning the type error report, Maxine realizes that \texttt{xs} can be
-either \texttt{[a]} type (Fig. \ref{fig:basic-mode-2}) or \texttt{Int} type (Fig. \ref{fig:basic-mode-2}). By matching the color in the
+either \texttt{[a]} or \texttt{Int}. By matching the color in the
 conflicting type panel (Fig. \ref{fig:anatomy}-G) and the highlighted error locations 
-Maxine knows that the \texttt{[a]} type results from the pattern matching of the
-\texttt{:} operator, and the \texttt{Int} type results from using \texttt{+} to
+Maxine knows that the \texttt{[a]} results from the pattern matching of the
+\texttt{:} operator, while \texttt{Int} results from using \texttt{+} to
  add two expressions. 
-
 At this point, Maxine knows the possible type 1 aligns with her understanding of
@@
 
-Maxine writes additional code to add only the even numbers in a list of integers.
-In this implementation, Maxine reuses the \texttt{sum} function she wrote
+Maxine writes additional code to add only even numbers in a list of integers, reusing the \texttt{sum} function she wrote
 earlier. After saving the file, \chameleon{} shows that there is a type error in the
-expression \texttt{sum} (Fig. \ref{fig:balance-mode-1}). However, this is not   \todo{Add annotation to figure like prev ones i.e. (1) so can refer to in text without reader searching for it?}
+expression \texttt{sum} (Fig.~\ref{fig:balance-mode-1}). However, this is not   \todo{Add annotation to figure like prev ones i.e. (1) so can refer to in text without reader searching for it?}
 helpful because Maxine knows that the implementation of  \texttt{sum} is correct 
@@
         \centering
-        \includegraphics[width=\linewidth]{images/balanced-mode-1.pdf}
+        \includegraphics[width=\linewidth,trim=0mm 8mm 0mm 0mm]{images/balanced-mode-1.pdf}
         \caption{
@@
    \centering
-        \includegraphics[width=\linewidth]{images/balanced-mode-2.pdf}
+        \includegraphics[width=\linewidth,trim=0mm 8mm 0mm 0mm]{images/balanced-mode-2.pdf}
         \caption{
@@
 
-For the task shown in section \ref{sub:balanced}, if Maxine is not satisfied by
-the options provided by \chameleon{}, by switching to advanced
-mode, Maxine has access to the debugging steps and the explanation layer.
-
 \begin{figure}
         \centering
-        \includegraphics[width=\linewidth]{images/advanced-mode-1.pdf}
+        \includegraphics[width=\linewidth,trim=0mm 8mm 0mm 0mm]{images/advanced-mode-1.pdf}
         \caption{
@@
         \centering
-        \includegraphics[width=\linewidth]{images/advanced-mode-2.pdf}
+        \includegraphics[width=\linewidth,trim=0mm 8mm 0mm 0mm]{images/advanced-mode-2.pdf}
         \caption{
@@
             the expression \texttt{filter isEven}. The left-hand side
-            and right-hand side should have the same type.
+            and the right-hand side should have the same type.
         }
@@
         \centering
-        \includegraphics[width=\linewidth]{images/advanced-mode-3.pdf}
+        \includegraphics[width=\linewidth,trim=0mm 8mm 0mm 0mm]{images/advanced-mode-3.pdf}
         \caption{
@@
 
+For the task shown in section \ref{sub:balanced}, if Maxine is not satisfied by
+the options provided by \chameleon{}, by switching to advanced
+mode, she has access to the debugging steps and the explanation layer.
+
 First, Maxine clicks on step 5 (Fig. \ref{fig:advanced-mode-step5}) and verifies
 that the two occurrences of \texttt{evens} are supposed to be identical, and the
-second usage dictates that \texttt{evens} is a list of integers. Second, she
+second use means \texttt{evens} is a list of integers. Second, she
 clicks on step 6 (Fig. \ref{fig:advanced-mode-step6}) and verifies that

### 2022-12-09 sections/chameleon.tex
Update bib

@@
 \section{Motivation}
-The design requirements of \chameleon{} are motivated by limitations of traditional type errors, as documented in a number of studies (e.g.~\cite{yang_improved_2000}), but which we illustrate here with a few motivating examples.
+The design requirements of \chameleon{} are motivated by limitations of traditional type errors, as documented in a number of studies (e.g.~\cite{yang_improved_2000, hage_solved_2020}), but which we illustrate here with a few motivating examples.
 %These limitations are sourced from the authors' frustration in teaching and writing Haskell.

### 2022-12-11 sections/chameleon.tex
Minor changes

@@
 This feature can be effective in resolving the problem that traditional tools tend to report errors that happened in source code from third-party libraries.
-With candidate expression cards, a programmer can freely change the context of the type error to the user-defined variables and functions to gain a better understanding of why her own code does not match the library code instead of the other direction.
+With candidate expression cards, a programmer can freely change the context of the type error to the user-defined variables and functions to gain a better understanding of why their own code does not match the library code instead of the other direction.
 
@@
 Typically, a programmer resolves  type errors by navigating through all the deduction steps and verifying whether each explanation aligns with the intended meaning of the program.
-Eventually, she will find a step that does not match her intention, and the type error can be fixed by modifying one of the two outlined error locations.
+Eventually, they will find a step that does not match her intention, and the type error can be fixed by modifying one of the two outlined error locations.
 

### 2022-12-11 sections/chameleon.tex
Updates

@@
 \noindent\textbf{Do not presume} which expression is to blame for the type error based on the order of computation or which possible type for an expression is 'actual' or 'expected'
-
-\end{itemize}
 
@@
 Selecting a candidate expression card is a way for programmers to communicate to the type checker the knowledge that the other candidate expressions are in fact well-typed.
-Once a card is selected, the information in the conflicting types panel changes according to the change of candidate expression.
+Once a card is selected, the information in the conflicting types block changes according to the change of candidate expression.
 In the editor window, some error locations change highlight colors to reflect the updated knowledge of the program.
@@
 Steps are shown as a list of sequentially  numbered circular buttons (step buttons) and an explanation layer in the editor window.
-In the explanation layer, two highlighted error locations are outlined, and a line is drawn to connect these two locations to a human-readable text explanation of
-their semantic connection.
+In the explanation layer, two highlighted error locations are outlined, and a line is drawn to connect these two locations.
+This line is accompanied by a human-readable text explanation of their semantic connection.
 Programmers are free to activate any step.
@@
 Programmers can use mouse and keyboard shortcuts to increment or decrement the step number or jump to any step.
-Typically, a programmer resolves  type errors by navigating through all the deduction steps and verifying whether each explanation aligns with the intended meaning of the program.
-Eventually, they will find a step that does not match her intention, and the type error can be fixed by modifying one of the two outlined error locations.
+Typically, programmers resolve  type errors by navigating through all the deduction steps and verifying whether each explanation aligns with the intended meaning of the program.
+Eventually, they will find a step that does not match their intention, and the type error can be fixed by modifying one of the two outlined error locations.
 
@@
 
-Internally, deduction steps are different ways to group the error locations into two different highlighted colors. 
+Internally, deduction steps are different ways to group the error locations into two highlighted colors.
 If all the highlights are placed in a list-like data structure, each increment of the step numbers moves the end of one color and the start of the other forward (Fig.~\ref{fig:step-interface} right).
@@
 Stuckey et al.~\cite{stuckey_interactive_2003} showed that program locations linked to the constraints from an MUS are all relevant to the type error and must include the cause of the error.
-In this paper, we will refer to these locations as \textit{error locations}.
 
@@
 Second, with concrete types, the debugging front end has more flexibility to change the representation of types.
-If the front end
-understands the type language in Haskell, it is possible to recognize and replace ground facts that are overlapping in both possible types and have the same effect. The other direction is impossible.
+If the front end understands the type language in Haskell, it is possible to recognize and replace ground facts that are overlapping in both possible types and have the same effect.
+The other direction is impossible.
+\tf{This sentence is not very clear}
 

### 2022-12-11 sections/conclusion.tex
Updates

@@
 We present \chameleon{}, a type debugging tool for the Haskell programming language.
-Its constraint-based type inference engine provides accurate and comprehensive type error reports.
-\chameleon{} also features an interactive debugging interface with three different debugging modes, enabling programmers to navigate deep type relations and triangulate conflicting statements in source code.
+Its constraint-based type inference engine provides unbiased and comprehensive type error reports.
+\chameleon{} also features an interactive debugging interface with three different debugging modes, enabling programmers to interactively explore the stepwise explanations.
 Our studies evaluated the tool's design with programmers.

### 2022-12-11 sections/evaluation.tex
Updates

@@
 At each task, a participant receives a single Haskell file that contains one or more type errors.
-The participant was asked to correct the code with the help of the given tool.
+They was then asked to correct the code with the help of the given tool.
 
-\subsubsection*{\textbf{Measurement}}
+\subsubsection*{\textbf{Data Collection}}
 Time is measured from the start of each task to the first time the program is successfully type-checked and also passes all the functional tests.
@@
 
-Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group were faster (p =0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1.
+Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group were faster (p = 0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1.
 For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity.
@@
         \textit{Low}  & Users only actively used universal features in all modes, for example, hovering on "Possible type 1" and "Possible type 2" to narrow down error space. \\ \hline
-        \textit{High}  & Users did everything from the Common feature group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards.
+        \textit{High}  & Users did everything from the low interaction group but used features specific to the Balanced mode and the Advanced mode, such as activating steps and expression cards.
 \\ \hline

### 2022-12-11 sections/chameleon.tex
Some changes

@@
 \noindent\textbf{Do not presume} which expression is to blame for the type error based on the order of computation or which possible type for an expression is 'actual' or 'expected'
+
+\end{itemize}
 
@@
 Selecting a candidate expression card is a way for programmers to communicate to the type checker the knowledge that the other candidate expressions are in fact well-typed.
-Once a card is selected, the information in the conflicting types block changes according to the change of candidate expression.
+Once a card is selected, the information in the conflicting types panel changes according to the change of candidate expression.
 In the editor window, some error locations change highlight colors to reflect the updated knowledge of the program.
@@
 Stuckey et al.~\cite{stuckey_interactive_2003} showed that program locations linked to the constraints from an MUS are all relevant to the type error and must include the cause of the error.
+In this paper, we will refer to these locations as \textit{error locations}.
 
@@
 Second, with concrete types, the debugging front end has more flexibility to change the representation of types.
-If the front end understands the type language in Haskell, it is possible to recognize and replace ground facts that are overlapping in both possible types and have the same effect.
-The other direction is impossible.
-\tf{This sentence is not very clear}
+If the front end
+understands the type language in Haskell, it is possible to recognize and replace ground facts that are overlapping in both possible types and have the same effect. The other direction is impossible.
 

### 2022-12-11 sections/related-work.tex
Some changes

@@
 The original Chameleon provides a certain level of ``interactive'' debugging through a command-line interface~\cite{stuckey_interactive_2003}, allowing the programmer to query different parts of the program.
-A recent study \cite{tsushima_type_2021} provided an interactive type debugging system that restores type information based on the output from CFT(Counter-factual typing)~\cite{chen_counter-factual_2014, chen_efficient_2020, chen_improving_2022} inference.
+A recent study \cite{tsushima_type_2021} provided an interactive type debugging system that restores type information based on the output from CFT (Counter-factual typing)~\cite{chen_counter-factual_2014, chen_efficient_2020, chen_improving_2022} inference.
 Both tools allow programmers to change their assumptions about which expression is ill-typed.

### 2022-12-12 main.tex
Change figure heads

@@
 
-Dynamically-typed programming languages are popular in education and the software industry.
+Dynamically typed programming languages are popular in education and the software industry.
 While presenting a low barrier to entry, they suffer from run-time type errors and longer-term problems in code quality and maintainability.
@@
 In particular, fixing type errors poses challenges to both novice users and experts.
-Further, compiler type error messages are presented in a static way that is biased towards the first occurence of the error in the program code.
+Further, compiler type error messages are presented in a static way that is biased toward the first occurrence of the error in the program code.
 To help users resolve such type errors we introduce \chameleon{}, a type debugging tool that presents type errors to the user in an unbiased way, allowing them to explore the full context of where the errors could occur.

### 2022-12-12 sections/chameleon.tex
Change figure heads

--- 2022-12-11 sections/chameleon.tex
@@
     \caption{
-With traditional type error messages, one expression is blamed for causing the type error. This expression may not always match the programmers' intention. 
+    A type error displayed in Visual Studio Code\cite{microsoft_visual_nodate} and the Haskell Vscode extension\cite{haskell_haskell_nodate}.
+The expression \texttt{addPair} is blamed for causing the type error. This may not match the programmers' intention. 
     }
@@
 In the type error message, only the \texttt{addPair} expression on line 4 was blamed.
-In this small example the whole context is visible, but the problem is worse for the programmer in large programs where the lines of code contributing to the type error are far apart in the source code.
+In this small example, the whole context is visible, but the problem is worse for the programmer in large programs where the lines of code contributing to the type error are far apart in the source code.
 
@@
 When the compiler rejects a program, the internal state of type checking is the result of complex computation.
-But the details of this process are hard to explain to users and usually not reported by compilers.
+But the details of this process are hard to explain to users and are usually not reported by compilers.
 For the typical type error shown in Fig.~\ref{fig:motivation-example}, the evidence for the type error is gathered from the previous two declarations.
-These have to be rediscovered by the programmer, using less rigourous methods.
+These have to be rediscovered by programmers using less rigorous methods.
 
@@
         \textbf{The anatomy of \chameleon{}.}
-        %\chameleon{} consists of two panes: an editor pane on the left and a
-        %debugging pane on the right. 
-        The editor pane (left) is similar to a
-        traditional code editor. Fragments of source code may have a highlight
-        color (A). Additionally, an explanation layer (B) displays in the editor
-        window, if deduction steps are enabled. The debugging window contains three blocks. 
-        An error statement and configuration block, containing
-        An error statement (D), and, optionally,
-        a list of candidate expression cards (E), a list of deduction steps (F), 
-        and a control bar (G) to increment/decrement deduction step. 
-        The conflicting types
-        block shows two alternative types (H) inferred by the type inference
-        engine. A relevant type information block to show additional information (I)
-        that may help understand type errors.
+        The editor pane (left) is similar to a traditional code editor. Fragments of source code may have a highlight
+        color (A).
+Additionally, an explanation layer (B) displays if deduction steps are enabled.
+The debugging pane contains three blocks.
+First, the error statement block contains an error statement (D), optionally, a list of candidate expression cards (E), a list of deduction steps (F), and a control bar (G) to increment/decrement deduction step.
+Second, the conflicting types block shows two alternative types (H).
+Third, the relevant type information block shows additional information (I) that may help understand type errors.
     }
@@
 
-The \chameleon{} debugging interface provides three main features to visualize the relation between error statement and locations in code and to explore the explanation of each error location given by the type inference engine.
+The \chameleon{} debugging interface provides three main features to visualize the relation between the error statement and locations in code and to explore the explanation of each error location given by the type inference engine.
 %Further, programmers  can turn on and off features to suit their preferences and debugging needs.
@@
 
-The type compare tool shows conflicting types in different colors, each associated with one or more error locations highlighted in a matching colour (Fig.~\ref{fig:compare}).  
+The type compare tool shows conflicting types in different colors, each associated with one or more error locations highlighted in a matching color (Fig.~\ref{fig:compare}).  
 %The type error must be fixed by modifying at least one of these locations. 
@@
     \caption{
-        \textbf{\chameleon{} with only candidate expression cards enabled.}
-        Indicates the type error can
-        occur in the definition of \texttt{x} or \texttt{y}.
+        \textbf{\chameleon{} with candidate expression cards enabled.}
+        Indicates the type error can occur in the definition of \texttt{x} or \texttt{y}.
     }
@@
     \caption{
-        \textbf{\chameleon{} with both candidate expression cards and deduction steps
-        enabled.}
+        \textbf{\chameleon{} with deduction steps enabled.}
         \chameleon{} explains the type error in four steps.
@@
 A candidate expression is an expression that can be inferred to have two conflicting types. 
-When a type error is detected, \chameleon{} provides a list of all candidate expressions and programmers are free to choose the problem to resolve by clicking on one candidate expression card.
+When a type error is detected, \chameleon{} provides a list of all candidate expressions, and programmers are free to choose the problem to resolve by clicking on one candidate expression card.
 In the example shown in Fig.
@@
 Internally, deduction steps are different ways to group the error locations into two highlighted colors.
-If all the highlights are placed in a list-like data structure, each increment of the step numbers moves the end of one color and the start of the other forward (Fig.~\ref{fig:step-interface} right).
+All the relevant constraints (highlights) are divided into two groups, denoted by the two colors.
+Each color infers a different type of the candidate expression.
+Each increment of the step changes the place where the two colors are divided by moving one constraint from one color to the other.
+(Fig.~\ref{fig:step-interface} right).
 
@@
 From analyzing how users use \chameleon{}, we realized that the ideal debugging interface should adapt to the specific programmer and programming task.
-It is common that a programmer wants the debugger to just ``show the answer", and it is also possible a programmer wants to dive deeper into the problem domain and search for the optimal solution.
+There are cases where a programmer wants the debugger to simply ``show the answer", and others to dive deeper into the problem domain and search for the optimal solution.
 To accommodate the need to customize the level of information density and granularity of control, \chameleon{} provides three modes: basic, balanced, and advanced.
@@
 The types reconstructed from Chameleon may be \texttt{(a, Int)} and \texttt{(a, String)}.
-Although the recovered types are consistent, they introduce a new type variable \texttt{a} making the error message harder to understand.
-To solve this issue in \chameleon{}, we first expand the MUS to a maximally satisfiable subset.
-We do so by removing one constraint from the MUS to get a consistent set, and then repeatedly adding constraints from the global constraint store while maintaining consistency.
-This process stops when no more constraints can be added.
-The resulting subset, while not helpful to inform error locations, will produce the most concrete conflicting types, see Fig.~\ref{fig:compare-to-original}.
+Although the recovered types are theoretically correct, they introduce a new type variable \texttt{a}, making the error message harder to understand.
+To solve this issue in \chameleon{}, for each constraint \texttt{c} in the MUS, we find a maximally satisfiable subset (MSS) from all the constraints that contain every other element of MUS but not \texttt{c}.
+These maximally satisfiable subsets, while not helpful to inform error locations, will produce the most concrete conflicting types, see Fig.~\ref{fig:compare-to-original}.
 
@@
 
-This has a few benefits.
-First, concrete types provide extra information to programmers.
+Using concrete types provides extra information to programmers.
 With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space.
 However, this information is not preserved in \texttt{(a, Float)}.
-Second, with concrete types, the debugging front end has more flexibility to change the representation of types.
-If the front end understands the type language in Haskell, it is possible to recognize and replace ground facts that are overlapping in both possible types and have the same effect.
-The other direction is impossible.
-\tf{This sentence is not very clear}
+If the types get too verbose for programmers to identify the conflict between the two signatures, it is always possible to mute the overlapping information and highlight the differences from the debugging front end.
 
@@
 Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed.
-\chameleon{} can construct a human-readable explanation from the MUS, for example, \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}.
-This low-level explanation is used in a high-level user interface described in section \ref{sub:deduction-steps} to help programmers reason about the validity of error locations.
+\chameleon{} can construct a human-readable explanation from the MUS.
+An example output for listing \ref{listing:1} can be: \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}.
+This explanation facilitates the deduction steps (Section \ref{sub:deduction-steps}).

### 2022-12-12 sections/evaluation.tex
Change figure heads

@@
 The data collected during study 1a, Fig.~\ref{fig:analysis-1a} does not show significant differences across Tasks 1-7.
-In hindsight, these tasks were trivial challenges for most users and the individual differences of participants are generally more significant than differences between treatments.
+In hindsight, these tasks were trivial challenges for most users and the individual differences among participants are generally more significant than the differences between treatments.
 However, one interesting observation is task 8, where the \chameleon{} group outperformed the GHC group (p-value = 0.002456, Wilcoxin signed test).
@@
 
-Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group were faster (p = 0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1.
+Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group was faster (p = 0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1.
 For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity.
@@
 In addition, mode switching is enabled by default.
-A few other user interfaces were designed and prototyped between the development of \chameleon{} 1 and \chameleon{} 2.
+A few other user interfaces \cite{anonymous_interactive_2021} were designed and prototyped between the development of \chameleon{} 1 and \chameleon{} 2.
 Study 2 therefore addresses the research question:
@@
 
-As shown in  Fig.~\ref{fig:r4-analysis}, time to complete each task roughly relates to the interaction level of participants.
+As shown in  Fig.~\ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants.
 Participants with higher interaction levels generally performed better, and the lowest interaction level was worse (one-way ANOVA since we have three samples, p = 0.0368).
@@
 
-Another observation is when using the mode switching feature of \chameleon{}, we find presenting the starting mode and finishing mode of each task and each participant in a confusion matrix is very telling (Fig.
+Another observation is when using the mode switching feature of \chameleon{}, we find presenting the starting mode and finishing mode of each task and each participant in a correlation matrix is very telling (Fig.
 \ref{fig:r4-mode-switching}).

### 2022-12-12 sections/introduction.tex
Updates
Change figure heads

@@
 They are often found to produce error-prone code \cite{chen_empirical_2020, wang_empirical_2015,xu_python_2016} and require strong programmer discipline to avoid pitfalls \cite{chen_empirical_2020}.
-For these reasons, many modern dynamically-typed languages have introduced static typing annotations as part of the core language features in recent years (e.g.\ \textit{TypeScript} and \textit{mypy}).
+For these reasons, many modern dynamically-typed languages have introduced static typing annotations as part of the core language features in recent years (e.g.\ \textit{TypeScript}~\cite{microsoft_javascript_nodate} and \textit{mypy}~\cite{mypy_mypy_nodate}).
 
@@
 
-The challenge of programming in statically-typed languages is that type errors can sometimes be difficult to resolve~\cite{tirronen_understanding_2015}.
-In particular, they may point to locations that are distant from real error causes, expose errors in hard to comprehend language, or provide misleading fixing suggestions~\cite{wu_how_2017}.
+The challenge of programming in statically-typed languages is that type errors can sometimes be difficult to resolve~\cite{tirronen_understanding_2015, hage_solved_2020}.
+In particular, they may point to locations that are not the real error causes, expose errors in cryptic language, or provide misleading fixing suggestions~\cite{wu_how_2017}.
 % Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}.
@@
 These features are integrated into a unified debugging environment and can be enabled or disabled separately based on the programmers' preferences and debugging needs.
-\chameleon{} is open-source and is available in both web~\cite{shuai_fu_chameleon_2022} and desktop versions.
+\chameleon{} is open-source and is available in both web~\cite{anonymous_chameleon_2022} and desktop versions.
 % \todo{Introduce Haskell earlier?
@@
 \item {
-    We report the results of four experiments designed to evaluate \chameleon{}.}
+    We report the results of three experiments designed to evaluate \chameleon{}.}
 \end{itemize}

### 2022-12-12 sections/related-work.tex
Change figure heads

@@
 \subsection{Finding all type error locations}
-Many have studied the approach of finding all locations that contribute to a type error (\cite{stuckey_interactive_2003, haack_type_2004, pavlinovic_practical_2015, schilling_constraint-free_2012}).
+Many have studied the approach of finding all locations that contribute to a type error~\cite{stuckey_interactive_2003, haack_type_2004, pavlinovic_practical_2015, schilling_constraint-free_2012}.
 Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error.

### 2022-12-12 sections/chameleon.tex
Updates from Overleaf

@@
 Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors.
-To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}''  In the example of Fig \ref{fig:motivation-example}, the type errors can be fixed in multiple locations.
+To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' The type error fig \ref{fig:motivation-example} can be fixed in multiple locations.
 For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}.
@@
 \subsubsection{\textbf{Traditional type errors are biased}}
-In the example of Fig \ref{fig:motivation-example}, the error message arbitrarily focuses the users' attention on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect.
+A common form of bias happens when a type error is reported in one expression, but it can occur in multiple other expressions as well.
+In fig \ref{fig:motivation-example}, the error message arbitrarily focuses the users' attention on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect.
 %Technically, this results from the side effect of different unification orders of the internal type-checking technique and has no bearing on what the programmers think and expect.
-Another related form of bias is that traditional type errors are often framed as conflicts between \texttt{Expected type} and \texttt{Actual type}.
+Another form of bias is that traditional type errors are often framed as conflicts between \texttt{Expected type} and \texttt{Actual type}.
 This framing is standard practice in most typed languages.
@@
 
-Selecting a candidate expression card is a way for programmers to communicate to the type checker the knowledge that the other candidate expressions are in fact well-typed.
-Once a card is selected, the information in the conflicting types block changes according to the change of candidate expression.
-In the editor window, some error locations change highlight colors to reflect the updated knowledge of the program.
+Selecting a candidate expression card is a way for programmers to communicate to the type checker the knowledge that the other candidate expressions are well-typed.
+Once a card is selected, the information in the conflicting types block changes to reflect the change of candidate expression.
+In the editor pane, some error locations change highlight colors based on the updated candidate expression.
 
@@
 
-Programmers interact with the candidate expression cards by clicking on them to commit the change of meaning or hovering on them to peek at the effect of the change; the effect is reverted once the cursor moves away.
+Programmers select a candidate expression card by clicking on one card.
+Alternatively, they can peek at the effect of the change by hovering on one card.
+The hover effect is reverted once the cursor moves away.
 

### 2022-12-12 sections/conclusion.tex

@@
 We present \chameleon{}, a type debugging tool for the Haskell programming language.
-Its constraint-based type inference engine provides unbiased and comprehensive type error reports.
-\chameleon{} also features an interactive debugging interface with three different debugging modes, enabling programmers to interactively explore the stepwise explanations.
+Its constraint-based type inference engine provides unbiased and comprehensive error location reporting.
+\chameleon{} also features an interactive debugging interface with three debugging modes, enabling programmers to interactively explore the stepwise explanations.
 Our studies evaluated the tool's design with programmers.

### 2022-12-12 sections/chameleon.tex
Numbering RQ

@@
     \centering
-    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/intro-compare.pdf}
+    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/intro-compare-2.pdf}
     \caption{
-        \textbf{\chameleon{} minimal output (with candidate expression cards and deduction step disabled)}.
-\chameleon{} identified the conflicting types for the function \texttt{a} and associated the relevant locations with each type.
-In addition, \chameleon{} leaves out the right-hand side of the declaration '\texttt{0}' and '\texttt{1}' because they are unrelated.
+        \textbf{\chameleon{} with type compare tool enabled}.
+\chameleon{} identified the conflicting types for the expression \texttt{u} and associated the relevant locations with each type.
+Compare the output with the traditional type error message in fig.
+\ref{fig:motivation-example}.
 }

### 2022-12-12 sections/evaluation.tex
Numbering RQ

@@
 \begin{itemize}
-    \item Do programmers prefer switching modes during debugging type errors?
-    \item What are programmers' preferences among the three modes provided by \chameleon{} 2?
-    \item How do programmers use the advanced features provided by \chameleon{} 2?
+    \item \textbf{RQ2.1} How do programmers use the advanced features provided by \chameleon{} 2?
+    \item \textbf{RQ2.2} Do programmers prefer switching modes during debugging type errors?
+    \item  \textbf{RQ2.3} What are programmers' preferences among the three modes provided by \chameleon{} 2?
 
@@
 
+\textbf{RQ2.1}.
 The most striking feature of the data is that users tend to vary wildly in their use of the tool.
-Some users used the features extensively while others completed the tasks without actively exploring the given information.
+Some users used the features extensively, while others completed the tasks without actively exploring the given information.
 Based on this discrepancy, we divided the users into three groups:
@@
 This observation suggests two characteristics of using multi-mode debugging tools.
-First, programmers are generally reluctant to switch modes (146/277 changing modes vs.
+First, to answer \textbf{RQ2.2} programmers are generally reluctant to switch modes (146/277 changing modes vs.
 131/277 staying in the same mode).
-Second, when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
+Second, to answer \textbf{RQ2.3} when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
 \begin{figure}

### 2022-12-14 sections/chameleon.tex
Updated bib

@@
 Haack and Wells~\cite{haack_type_2004} noted that ``\textit{Identifying only one node or subtree of the program as the error location makes it difficult for programmers to understand type errors.
-To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' The type error fig \ref{fig:motivation-example} can be fixed in multiple locations.
+To choose the correct place to fix a type error, the programmer must find all the other program points that participate in the error.}'' The type error in Fig.~\ref{fig:motivation-example} can be fixed in multiple locations.
 For instance  replacing \texttt{['0'..'9']} on line 1 with \texttt{[0..9]}, or replacing \texttt{fst x} and \texttt{snd x} on line 2 with \texttt {read (fst x)} and \texttt{read (snd  x)}.
@@
 A common form of bias happens when a type error is reported in one expression, but it can occur in multiple other expressions as well.
-In fig \ref{fig:motivation-example}, the error message arbitrarily focuses the users' attention on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect.
+In Fig.~\ref{fig:motivation-example}, the error message arbitrarily focuses the users' attention on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect.
 %Technically, this results from the side effect of different unification orders of the internal type-checking technique and has no bearing on what the programmers think and expect.
@@
 \ref{fig:expression}, \texttt{x} and \texttt{y} are both candidate expressions.
-Fixing either type error will make both expressions well-typed.
+Fixing either type error can make both expressions well-typed.
 
@@
 \end{listing}
-For instance, for the program in the listing \ref{listing:1}, \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition).
+For instance, for the program in Listing~\ref{listing:1}, \chameleon{}  generates the following constraints and labels (in brackets) $T_a = Bool$ (if condition), $T_b = T_c$ (if branches), $T_a= String$  (definition).
 Clearly, as $T_a$ can not unify with both \textit{Bool} and \textit{String}, this program is not well typed.
 \chameleon{} can construct a human-readable explanation from the MUS.
-An example output for listing \ref{listing:1} can be: \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}.
+An example output for Listing~\ref{listing:1} can be: \texttt{a} has type \texttt{Bool} because \texttt{a} is the condition of an if statement; however, \texttt{a} has type \texttt{String} because \texttt{a} is defined as the string literal \texttt{"True"}.
 This explanation facilitates the deduction steps (Section \ref{sub:deduction-steps}).

### 2022-12-14 sections/conclusion.tex
Updated bib

@@
 
-We present \chameleon{}, a type debugging tool for the Haskell programming language.
-Its constraint-based type inference engine provides unbiased and comprehensive error location reporting.
-\chameleon{} also features an interactive debugging interface with three debugging modes, enabling programmers to interactively explore the stepwise explanations.
+We present \chameleon{}, a type debugging tool for the Haskell programming language. Its constraint-based type inference engine provides unbiased and comprehensive error location reporting. 
+% This sentence could go away:
+%\chameleon{} also features an interactive debugging interface with three debugging modes, enabling programmers to interactively explore the stepwise explanations. 
 Our studies evaluated the tool's design with programmers.
@@
 Further, programmers actively using \chameleon{} interactive features are shown to fix type errors faster than simply reading the type error output.
+\chameleon{} currently works with the Haskell language, but in future we plan to extend the typechecking system to work with other strongly typed languages, such as Rust or TypeScript.

### 2022-12-14 sections/discussion.tex
Updated bib

@@
 We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information.
-With the program showing in the listing \ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}.
-Although, in theory, these two statements are to the same effect, programmers are often required to compute the latter from the former or vice versa.
+With the program shown in Listing~\ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}.
+Although, in theory, these two statements have the same effect, programmers are often required to compute the latter from the former or vice versa.
 And this computation may carry out multiple layers.

### 2022-12-15 sections/chameleon.tex
Some small fixes

@@
 In the type error message, only the \texttt{addPair} expression on line 4 was blamed.
-In this small example, the whole context is visible, but the problem is worse for the programmer in large programs where the lines of code contributing to the type error are far apart in the source code.
+In this small example, the whole context is visible, but it can be more problematic in large programs where the lines contributing to the type error are far apart in the source code.
 
@@
 A common form of bias happens when a type error is reported in one expression, but it can occur in multiple other expressions as well.
-In Fig.~\ref{fig:motivation-example}, the error message arbitrarily focuses the users' attention on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect.
+In Fig.~\ref{fig:motivation-example}, the error message arbitrarily focuses on only \texttt{addPair}, while ignoring that the literals in the definition of \texttt{u} may be incorrect.
 %Technically, this results from the side effect of different unification orders of the internal type-checking technique and has no bearing on what the programmers think and expect.
@@
 However, what is the \texttt{expected} and what is \texttt{actual} are a side effect of different unification orders rather than the intention of the programmer.
-In both of these cases, the error message may lead programmers to falsely believe the validity of parts of code and wrongly accuse others.
+In both forms, the error message may lead programmers to falsely believe the validity of parts of code and wrongly accuse others.
 
 \subsubsection{\textbf{Traditional type errors give poor explanations}}
-When the compiler rejects a program, the internal state of type checking is the result of complex computation.
+When the compiler rejects a program, the internal state of type checking is the result of a complex computation.
 But the details of this process are hard to explain to users and are usually not reported by compilers.
@@
 
-\noindent\textbf{Do not presume} which expression is to blame for the type error based on the order of computation or which possible type for an expression is 'actual' or 'expected'
+\noindent\textbf{Do not presume} which expression is to blame for the type error based on the order of computation or which possible type for an expression is 'actual' or 'expected'.
 
@@
 
-The type compare tool shows conflicting types in different colors, each associated with one or more error locations highlighted in a matching color (Fig.~\ref{fig:compare}).  
+The type compare tool shows conflicting types in different colors, each type associated with one or more error locations highlighted in a matching color (Fig.~\ref{fig:compare}).  
 %The type error must be fixed by modifying at least one of these locations. 
@@
 
-Selecting a candidate expression card is a way for programmers to communicate to the type checker the knowledge that the other candidate expressions are well-typed.
-Once a card is selected, the information in the conflicting types block changes to reflect the change of candidate expression.
-In the editor pane, some error locations change highlight colors based on the updated candidate expression.
-
-This feature can be effective in resolving the problem that traditional tools tend to report errors that happened in source code from third-party libraries.
+% This feature can be effective in resolving the problem that traditional tools tend to report errors that happened in source code from third-party libraries.
 With candidate expression cards, a programmer can freely change the context of the type error to the user-defined variables and functions to gain a better understanding of why their own code does not match the library code instead of the other direction.
@@
 Programmers select a candidate expression card by clicking on one card.
-Alternatively, they can peek at the effect of the change by hovering on one card.
+Once a card is selected, the information in the conflicting types block changes to reflect the change of candidate expression.
+In the editor pane, some error locations change highlight colors based on the updated candidate expression.
+Alternatively, programmers can preview the change of a candidate expression by hovering on one card.
 The hover effect is reverted once the cursor moves away.
@@
 Steps are shown as a list of sequentially  numbered circular buttons (step buttons) and an explanation layer in the editor window.
-In the explanation layer, two highlighted error locations are outlined, and a line is drawn to connect these two locations.
+In the explanation layer, the two locations under examination are outlined, and a line is drawn to connect these two locations.
 This line is accompanied by a human-readable text explanation of their semantic connection.
 Programmers are free to activate any step.
-The active step is shown as the green button.
+The active step is shown in green.
 When activating a step, some highlights switch color.
@@
 Programmers can use mouse and keyboard shortcuts to increment or decrement the step number or jump to any step.
-Typically, programmers resolve  type errors by navigating through all the deduction steps and verifying whether each explanation aligns with the intended meaning of the program.
-Eventually, they will find a step that does not match their intention, and the type error can be fixed by modifying one of the two outlined error locations.
+Programmers resolve type errors by navigating through all the deduction steps and verifying whether each explanation aligns with their intention.
+Eventually, they will find a step that does not match, and the type error can be fixed by modifying one of the two outlined locations.
 
@@
 Unlike traditional type errors produced by the Glasgow Haskell Compiler (GHC)~\cite{ben_gamari_home_2022}, which uses a Hindley–Milner type inference system, Chameleon infers types using constraint solving.
-In Chameleon, Constraints are generated from the source code based on typing rules.
-In addition, each constraint is labeled with the location (line and column numbers) where it is generated.
+In Chameleon, constraints are generated from the source code based on typing rules.
+In addition, each constraint is labeled with the location where it is generated.
 This set of constraints is consistent if the program is well-typed and inconsistent otherwise.
@@
 With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space.
-However, this information is not preserved in \texttt{(a, Float)}.
+However, a type of \texttt{(a, Float)} does not preserve such information.
 If the types get too verbose for programmers to identify the conflict between the two signatures, it is always possible to mute the overlapping information and highlight the differences from the debugging front end.
@@
 Similar to the type explanation system  in ~\cite{jun_explaining_2002},  \chameleon{} is able to produce a human-readable explanation, but for type errors.
-This is achieved by annotating nodes in the abstract syntax tree with constraints and the type of rules applied.
+This is achieved by annotating nodes in the abstract syntax tree with constraints and the typing rules used.
 We generate an inference history from constraints and accompanying annotations.

### 2022-12-15 sections/conclusion.tex
Some small fixes

@@
 Further, programmers actively using \chameleon{} interactive features are shown to fix type errors faster than simply reading the type error output.
-\chameleon{} currently works with the Haskell language, but in future we plan to extend the typechecking system to work with other strongly typed languages, such as Rust or TypeScript.
+\chameleon{} currently works with the Haskell language, but in the future, we plan to extend the type-checking system to work with other strongly typed languages, such as Rust or TypeScript.

### 2022-12-15 sections/discussion.tex
Some small fixes

@@
 
-\subsection{Externalise Intermediate Typing Inforamtion}
+\subsection{Externalize Intermediate Typing Information}
 We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information.
 With the program shown in Listing~\ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}.
-Although, in theory, these two statements have the same effect, programmers are often required to compute the latter from the former or vice versa.
+Although  these two statements are equivalent in theory, programmers are often required to compute the latter from the former or vice versa.
 And this computation may carry out multiple layers.

### 2022-12-15 sections/evaluation.tex
Updated bib
Some small fixes

@@
 %Recruiting from social media allowed us to access a more diverse demographic that better represent the true population of Haskell programmers. 
-Participation is fully anonymized; detailed ethical implications of these experiments are reviewed and approved by the IRB of the institution of the authors.
-
-\todo{Need more detail on participants - years experience, male vs female, etc}
+Participation is fully anonymized; detailed ethical implications of these experiments are reviewed and approved by the IRB of the authors' institution.
 
@@
 We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions.
-This provides us with a "poor man's video recording" to identify usability issues in the study and to recognize general patterns.
+This provides us with a ``poor man's video recording" to identify usability issues in the study and to recognize general patterns.
 %The reason we found this format inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.
@@
 In both studies, the type errors include simple mismatch, confusing syntax, missing instance, precedence and fixation, infinite types, and confusing list versus element.
-The categories of type errors to include in the tasks followed the common type errors in Tirronen's study \cite{tirronen_understanding_2015}.
+These categories follow the common type errors in Tirronen's study \cite{tirronen_understanding_2015}.
 
@@
 The data collected during study 1a, Fig.~\ref{fig:analysis-1a} does not show significant differences across Tasks 1-7.
-In hindsight, these tasks were trivial challenges for most users and the individual differences among participants are generally more significant than the differences between treatments.
-However, one interesting observation is task 8, where the \chameleon{} group outperformed the GHC group (p-value = 0.002456, Wilcoxin signed test).
-We attribute this significant difference to the hardness of Task 8.
+In hindsight, these tasks were trivial challenges for most users, and the individual differences among participants are generally more significant than the differences between treatments.
+However, one interesting observation is task 8, where the \chameleon{} group outperformed the GHC group.
+We attribute this significant difference to the difficulty of Task 8.
 The source file is longer and involves more language features (abstract data types and high-level functions).
 GHC struggles to produce a relevant error message for this type of error.
-From this result, we hypothesized that we may observe a more significant difference using tasks with lengthier and more realistic source code. 
+From this result, we hypothesized that we might observe a more significant difference using tasks with lengthier and more realistic source code. 
 This hypothesis is also supported by the most common feedback (7 out of 35) that the tasks were too trivial to invite meaningful evaluation.
@@
 
-Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group was faster (p = 0.00742, Wilcoxin signed test) than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1.
+Therefore, in study 1b we introduced more difficult challenges and indeed observed that the \chameleon{} group was faster than the GHC group in almost all tasks (figure \ref{fig:analysis-1b}), barring task 1.
+A two-sample paired t-test was performed to compare the completion time between \chameleon{} and GHC groups.
+There was a significant difference between the two groups: $t(23) = -3.86, p = 0007$.
 For task 1, it is suspected that some participants spent more time exploring the interface of \chameleon{} due to its unfamiliarity.
@@
 Based on observations of Study 1 we introduced several new features to \chameleon{}, eventually resulting in the UI depicted in Figs.~(2-13).
-Advanced debugging features such as deduction steps and candidate expressions were available in this iteration.
-In addition, mode switching is enabled by default.
+Interactive features were available in this iteration, such as deduction steps, candidate expressions, and mode switching.
 A few other user interfaces \cite{anonymous_interactive_2021} were designed and prototyped between the development of \chameleon{} 1 and \chameleon{} 2.
-Study 2 therefore addresses the research question:
+Study 2 addresses the research question:
 
@@
 
-During the study, the initial mode of each task alternated through the three different modes and repeated three cycles in nine tasks.
+During each run, the initial mode of each task alternated through the three different modes and repeated three cycles in nine tasks.
 The order of the three modes in each cycle is counterbalanced among all participants.
@@
 Study 2 is more exploratory in methodology than Study 1.
-We were expecting to let programmers discover their own way of using the tool.
+We encouraged programmers  to discover their way of using the tool.
 In post hoc analysis of the collected log data, we were able to extrapolate some interesting patterns of how the tool was used.
-%We share a few of our observations in this section.
 
@@
 As shown in  Fig.~\ref{fig:r4-analysis}, the time to complete each task roughly relates to the interaction level of participants.
-Participants with higher interaction levels generally performed better, and the lowest interaction level was worse (one-way ANOVA since we have three samples, p = 0.0368).
+Participants with higher interaction levels generally performed better, and the lowest interaction level was worse.
+Tukey’s HSD Test for multiple comparisons found that the completion time was significantly different between the minimal interaction group and the high interaction group ($p \le 0.001$, 95\% C.I.
+= [18.26, 31.41]), and between the minimal interaction group and the low interaction group ($p \le 0.001$, 95\% C.I.
+= [11.96, 26.67]).
 The results from three tasks stand out from the general trend: in Tasks 4 and 6, higher interaction users performed worse, and in task 9, the general trend is exaggerated.
@@
 The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code).
-These may be useful because following the natural reading order will allow participants to skip lines on the bottom, and thus the advantage of \chameleon{} is less critical.
+Users simply reading top to bottom could quickly identify the error without needing to skip unrelated sections of code using the information provided by /chameleon{}.
+This reduced the apparent benefit of /chameleon{} in these tasks.
 On the other hand, task 9 is the lengthiest task of all.
@@
     \centering
-    \includegraphics[width=\linewidth,trim=0mm 15mm 0mm 35mm,clip]{images/user-study-2.pdf}
+    \includegraphics[width=\linewidth,trim=0mm 15mm 0mm 50mm,clip]{images/user-study-2.pdf}
     \caption{Study 2 task completion time (secs.) with 95\% confidence intervals.}
@@
 
-Another observation is when using the mode switching feature of \chameleon{}, we find presenting the starting mode and finishing mode of each task and each participant in a correlation matrix is very telling (Fig.
+Another observation is when using the mode switching feature of \chameleon{}, we show this by presenting the starting mode and finishing mode of each task and each participant in a correlation matrix (Fig.
 \ref{fig:r4-mode-switching}).
 This observation suggests two characteristics of using multi-mode debugging tools.
-First, to answer \textbf{RQ2.2} programmers are generally reluctant to switch modes (146/277 changing modes vs.
-131/277 staying in the same mode).
+First, to answer \textbf{RQ2.2} programmers are generally reluctant to switch modes: 53\% changing modes vs.
+47\% staying in the same mode.
 Second, to answer \textbf{RQ2.3} when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
@@
 One threat to the validity of the evaluation is the number of participants.
-Although for each study we received hundreds of online participants, the studies suffer from a high abandonment rate (especially study 1b).
-This was expected: the programming challenges are difficult and our volunteer participants are unremunerated.
-%To encourage completion, we included a cheat sheet containing a summary of the key functionality of the tools and necessary type signatures.
-A skip button is available for when participants are stuck.
+Although for each study we received hundreds of online participants, the studies suffered from a high abandonment rate (especially study 1b).
+This was expected: the programming challenges are difficult, and our volunteer participants are unremunerated.
 Because we recruited participants online and anonymized all the participants, it is possible for participants of a previous study to enter a later one.
@@
 Conducting studies remotely and unsupervised there is no way to intervene when users encounter usability issues.
-To mitigate this we conducted cognitive walkthroughs and sandbox pilots before running each study.
+To mitigate this, we conducted cognitive walkthroughs and sandbox pilots before running each study.
 
@@
 However, the complexity is just short of what professional Haskell programmers may face in the production Haskell codebase.
-It would be interesting to see how \chameleon{} is used against type errors that span multiple files and packages and include confusing abstractions, like Monads, Monad transformers, and Lenses.
+It would be interesting to see how \chameleon{} is used against type errors that span multiple files and packages and include more confusing abstractions, like Monads, Monad transformers, and Lenses.
 

### 2022-12-15 sections/introduction.tex
Some small fixes

@@
 
-Dynamically-typed programming languages such as JavaScript and Python have risen in popularity in recent decades \cite{chatley_next_2019}.
+Dynamically typed programming languages such as JavaScript and Python have risen in popularity in recent decades \cite{chatley_next_2019}.
 These languages present a low barrier of entry, especially to beginner programmers: they require no type declaration, variable types or object structures can be modified dynamically, and functions can deal with dynamic input using ad-hoc polymorphism and runtime reflection.
-However, studies show that dynamically-typed languages negatively affect development productivity \cite{kleinschmager_static_2012}, code usability \cite{mayer_empirical_2012}, and code quality \cite{gao_type_2017, ray_large-scale_2017, meyerovich_empirical_2013}.
+However, studies show that dynamically typed languages negatively affect development productivity \cite{kleinschmager_static_2012}, code usability \cite{mayer_empirical_2012}, and code quality \cite{gao_type_2017, ray_large-scale_2017, meyerovich_empirical_2013}.
 They are often found to produce error-prone code \cite{chen_empirical_2020, wang_empirical_2015,xu_python_2016} and require strong programmer discipline to avoid pitfalls \cite{chen_empirical_2020}.
@@
 Techniques such as type inference and algebraic types have been standard practice for decades in functional languages such as ML and Haskell, and more recently in multi-paradigm languages, such as Rust and TypeScript.
-Haskell is considered a ``type-system laboratory" \cite{hudak_history_2007}.
-Various type system advances were introduced in Haskell and ended up in mainstream languages years or even decades after.
+Various type system advances were introduced in Haskell and ended up in mainstream languages years or even decades after, leading many to consider Haskell the ``type-system laboratory" \cite{hudak_history_2007}.
 Type classes, an implementation of generic programming, were introduced to Haskell in 1988~\cite{hudak_history_2007}, and now can be found in most popular languages such as C\#~\cite{bill_wagner_constraints_2022}, Java~\cite{oracle_generic_2022}, and TypeScript~\cite{microsoft_documentation_2022}.
 
-The challenge of programming in statically-typed languages is that type errors can sometimes be difficult to resolve~\cite{tirronen_understanding_2015, hage_solved_2020}.
-In particular, they may point to locations that are not the real error causes, expose errors in cryptic language, or provide misleading fixing suggestions~\cite{wu_how_2017}.
+One crucial challenge of programming in statically-typed languages is that type errors can sometimes be difficult to resolve~\cite{tirronen_understanding_2015, hage_solved_2020}.
+In particular, they may point to locations that are not the root  causes of the type error, expose errors in cryptic language, or provide misleading fixing suggestions~\cite{wu_how_2017}.
 % Compared to dynamic type systems, static type systems offer programmers opportunities to weed out a large number of errors at compile time (reducing the need for run-time debugging) and to increase usability~\cite{mayer_static_2012} and codebase maintainability~\cite{kleinschmager_static_2012}.
@@
 This paper introduces \chameleon{}, an interactive type debugging tool for Haskell.
-It can visualize the relevant context of a type error (where it happens or could have happened and which parts of code cause it).
+It can visualize the relevant context of a type error: where it happens or could have happened and which parts of the code cause it.
 In addition, \chameleon{} allows programmers to interactively explore all the parts of code where multiple types can be inferred and to resolve ambiguity.
-The most noticeable features provided by \chameleon{} are the type compare tool, the deduction step feature, and the candidate expression card feature.
-These features are integrated into a unified debugging environment and can be enabled or disabled separately based on the programmers' preferences and debugging needs.
-\chameleon{} is open-source and is available in both web~\cite{anonymous_chameleon_2022} and desktop versions.
+The most noticeable features are the type compare tool (Section \ref{sub:type-compare}), the candidate expression card (Section \ref{sub:candidate-expression}), and the deduction step (Section \ref{sub:deduction-steps}).
+These features are integrated into a debugging environment and can be enabled or disabled separately based on the programmers' preferences and debugging needs.
+\chameleon{} is open-source and is available in web~\cite{anonymous_chameleon_2022} and desktop versions.
 % \todo{Introduce Haskell earlier?

### 2022-12-15 sections/related-work.tex
Updated bib
Some small fixes

@@
 The language supported in Haack's work was a subset of Standard ML.
-The original Chameleon~\cite{stuckey_interactive_2003} used a similar technique but generate constraints in Constraint handling rules (CHR).
-As a result,  Chameleon supported advanced type-level features (type classes and functionally dependent types).
+The original Chameleon~\cite{stuckey_interactive_2003} used  Constraint handling rules (CHR) to support the computing of type error slices in Haskell.
+Chameleon also supported advanced type-level features (type classes and functionally dependent types).
 The project also introduced the ability to query type information through a command line-style interface.
@@
 While finding all error locations is useful in comprehending type errors, it is only 1 of the 7 properties listed in the proposed manifesto of good type error reporting~\cite{yang_improved_2000}.
+To the best of our knowledge, ours is the first user-centered evaluation of an interactive type debugging system involving type-error slicing.
 

### 2022-12-15 sections/walkthrough.tex
Some small fixes

@@
 numbers, but \chameleon{} shows there is a type error (Fig.~\ref{fig:basic-mode-1}). 
-After reading the error reports, Maxine realizes that the error possibly revolves 
-around the expression \texttt{xs}.
+After reading the error reports, Maxine realizes that the error revolves 
+around the expression \texttt{xs}. That is: \texttt{xs} can be
+either \texttt{[a]} or \texttt{Int}. By matching the color in the
+conflicting type block (Fig. \ref{fig:anatomy}-H) and the highlighted error locations 
+Maxine knows that the \texttt{[a]} results from the pattern matching of the
+\texttt{:} operator, while \texttt{Int} results from using \texttt{+} to
+ add two expressions.
 
@@
 
-After scanning the type error report, Maxine realizes that \texttt{xs} can be
-either \texttt{[a]} or \texttt{Int}. By matching the color in the
-conflicting type panel (Fig. \ref{fig:anatomy}-G) and the highlighted error locations 
-Maxine knows that the \texttt{[a]} results from the pattern matching of the
-\texttt{:} operator, while \texttt{Int} results from using \texttt{+} to
- add two expressions. 
-At this point, Maxine knows the possible type 1 aligns with her understanding of
-the program, and therefore, the error locations with blue highlights must be
-erroneous. After examining the program, it is obvious that Maxine missed
-applying the \texttt{sum} function recursively at the right-hand side of the
-addition.
+At this point, Maxine knows the possible type 1 aligns with her intention, and therefore, the error locations with blue highlights must be erroneous.
+After examining the program, it comes clear that Maxine forgets to apply the \texttt{sum} function recursively at the right-hand side of the addition.
 
@@
 
-Maxine writes additional code to add only even numbers in a list of integers, reusing the \texttt{sum} function she wrote
-earlier. After saving the file, \chameleon{} shows that there is a type error in the
-expression \texttt{sum} (Fig.~\ref{fig:balance-mode-1}). However, this is not   \todo{Add annotation to figure like prev ones i.e. (1) so can refer to in text without reader searching for it?}
-helpful because Maxine knows that the implementation of  \texttt{sum} is correct 
-because it was verified in the previous task. Switching to
-balanced mode, Maxine noticed that \chameleon{} shows two cards: \texttt{sum}
-and \texttt{evens}.
+Maxine writes additional code to add only even numbers in a list of integers, reusing the \texttt{sum} function she wrote earlier.
+After saving the file, \chameleon{} shows a type error in the expression \texttt{sum} (Fig.~\ref{fig:balance-mode-1}).
+However, this is not helpful because Maxine has just verified the implementation of \texttt{sum}.
+Switching to balanced mode, \chameleon{} shows two cards: \texttt{sum} and \texttt{evens}.
 
@@
             conflicting types panel to show the possible types for \texttt{evens},
-            and the changes highlights color to reflect the assumption that the
+            and the changes highlight color to reflect the assumption that the
             definition of \texttt{evens} is the cause of the error.  
@@
 
-Maxine therefore clicks on the \texttt{evens} card and \chameleon{} shows two
+Maxine therefore clicks on the \texttt{evens} card and \chameleon{} reports two
 possible types for the expression \texttt{[Int]} and \texttt{[Int] -> [Int]}
@@
 a temporary list of even integers (hence it is of \texttt{[Int]} types), Maxine
-knows the Possible type 2 is unintended, and blue in-text highlights must
-contain the cause. It does not take long until Maxine realizes that the list 
+knows the Possible type 2 is unintended. The locations with blue highlights must
+contain the cause. It does not take long for Maxine to realize the list 
 \texttt{l} is not supplied to the \texttt{filter} function.
@@
 the options provided by \chameleon{}, by switching to advanced
-mode, she has access to the debugging steps and the explanation layer.
+mode, she has access to the debugging steps.
 

### 2022-12-19 sections/chameleon.tex
UPdate image ref
Latest change

@@
     \centering
-    \includegraphics[width=\linewidth, trim=0mm 10mm 0mm 0mm]{images/step-interface.pdf}
+    \includegraphics[width=\linewidth, trim=0mm 0mm 0mm 0mm]{images/step-explain.pdf}
     \caption{
-Deduction steps if they are shown all at once (left).
+Deduction steps if they are shown all at once.
 In practice, steps are shown one at a time.
@@
 \ref{fig:anatomy}-F).
-To increment or decrement the deduction step can be intuitively thought of as moving the position of the \textit{split point} where the blue and orange highlights divide (right).
+To increment or decrement the deduction step can be intuitively thought of as moving the position of the \textit{splitting point} (dotted lines) where the blue and orange highlights divide.
         }
@@
 
-Internally, deduction steps are different ways to group the error locations into two highlighted colors.
-All the relevant constraints (highlights) are divided into two groups, denoted by the two colors.
+Internally, deduction steps are different ways to divide the error locations into two  groups, denoted by the two colors.
 Each color infers a different type of the candidate expression.
-Each increment of the step changes the place where the two colors are divided by moving one constraint from one color to the other.
-(Fig.~\ref{fig:step-interface} right).
+Each increment/decrement of the step changes the splitting point (dotted lines in Fig.~\ref{fig:step-interface}) of the two colors.
 

### 2022-12-19 data/v1b-analysis.rmd
Update analysis

@@
 
-attemps <- events %>%
-  filter(participant_id %in% completed) %>% 
-  filter(event_name == "attempts")
+# 
+# attemps <- events %>%
+#   filter(participant_id %in% completed) %>% 
+#   filter(event_name == "attempts")
 
@@
 
-ggsave("r2-data.pdf", device = "pdf", plot = last_plot(), units ="mm", width = 210)
+# ggsave("r2-data.pdf", device = "pdf", plot = last_plot(), units ="mm", width = 210)
 
@@
 
-r <- wilcox.test(d$Chameleon, d$Ghc, paired=TRUE)
-r
+shapiro.test(durations$duration)
+t.test(d$Chameleon, d$Ghc, paired = TRUE, alternative = "two.sided")
+
+# r <- wilcox.test(d$Chameleon, d$Ghc, paired=TRUE)
+# r
 ```
@@
 ```{r}
+```
+
+```{r}
 events |>
@@
   geom_density()
+```
+
+```{r}
+durations |> 
+durations |> 
+  ggplot(mapping=aes(duration)) +
 

### 2023-03-08 response.md
Camera Ready

Dear Reviewers,

Many thanks for your considered feedback and overall positive response.  Below we respond to the main questions raised in the reviews 19A and 10C (as no questions were raised in #19B and #19D).  We are confident that we can address all issues with a minor revision of our manuscript.

# Review #19A
We agree with the reviewer's constructive comments. To further clarify the discussion, we answered some questions raised in the review.

**Question**: "Through three studies involving real programmers" -- Are some programmers not real?

**Answer**: The intention was to differentiate mixed experienced programmers from university students. With that said, we agree with the reviewer that our wording could be more precise.

**Question**: Tasks were less than 1 minute in length? Seems quite short.

**Answer**: The mean task time for user study 1a is less than 1 minute. And admittedly, the user study 1a tasks were relatively trivial, as pointed out in the RQ1 discussion on page 7. The mean time in task 1b and study 2 are substantially longer.

**Question**: What is "/chameleon" mentioned on page 8? Looks like a broken LaTeX macro maybe.

**Answer**: Yes, it's a broken LaTex macro. We will conduct more thorough proofreading should the paper be accepted.

----

# Review #19C

## Questions for authors' response:

**Question**: How did the authors recognize a type error as "fixed"?

**Answer**: A type error is treated as "fixed" when the first time the program type check and some functional tests succeed. Functional tests are in place to prevent participants from "fixing" type errors by deleting offending code, as the reviewer noted. This information is provided in IV-A, data collection subsection (p. 7).

**Question**: In RQ1, why is a paired t-test possible?

**Answer**: The paired t-test is possible because the effect was measured on the same participant receiving both treatments (GHC compiler error message and ChameleonIDE). In section IV-A (p. 7), we noted the with-in-subject design of our human studies and provided justification. To further clarify,  In studies 1a and 1b, each participant received 8 tasks, 4 using the ChameleonIDE and 4 using GHC. The treatment alternated throughout the study (e.g., ChameleonIDE for tasks 1,3,5,7, GHC for tasks 2,4,6,8, or the same configuration with the order of tools reversed). The two configurations were counterbalanced within the population. We agree with the reviewer this information should more explicitly stated in the manuscript.

## Other questions raised in the review
We agree with the constructive comments made by the reviewer.
Some of the review's questions were answered here to help clarify the discussion.

**Question**: What kind of tasks are used in Study 1b and Study 2?
Only the sentence "in study 1b we introduced more difficult challenges" is the explanation. I could see source files on HotCRP but I could not understand whether they are more practical cases or not. The paper should show a summary of type errors and programs.

**Answer**: The tasks given in the supplementary material were the same tasks used in the studies. We agree with the reviewer the paper would benefit from a concise summary of the task programs or a synopsis of the supporting material.

**Question**: How did the authors deal with participants who switched the IDE modes during the tasks in Fig.17? Please clarify the number of data points. In addition, please explain the risk of biased results, as participants were able to switch to advanced modes.

**Answer**: The task time shown in  Fig.17 is not grouped by each mode of ChameleonIDE but rather by different levels of interactivity of participants regardless of their mode assignments. The justification was given in RQ 2.1 discussion on page 8. For instance, if a participant voluntarily switches modes multiple times, even to less advanced modes, during resolving type error, they are placed in the high interactivity group. Conversely, participants in advanced mode can still be placed in the minimal interaction group if they lack interaction with the tool.

**Question**: The availability of the tool is not described.

**Answer**: ChameleonIDE is open source and is available in web and desktop versions. The information is provided at the end of paragraph 4 on page 1.

### 2023-03-08 sections/evaluation.tex
Camera Ready

@@
 Time is measured from the start of each task to the first time the program is successfully type-checked and also passes all the functional tests.
-The data is automatically recorded by the online debugging environment.
+Participants are able to skip a task if they are stuck.
+% The data is automatically recorded by the online debugging environment.
 To not introduce a barrier to completing the study, every task can be skipped if the participant made three failed attempts or is stuck for over 1 minute on the task.
@@
 
-We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions.
-This provides us with a ``poor man's video recording" to identify usability issues in the study and to recognize general patterns.
+We used a browser session recording tool~\cite{openreplay_openreplay_2022} to record the study sessions. This allows us to identify usability issues in the study and to recognize general patterns. 
 %The reason we found this format inferior is that the recording technology does not provide a high enough sampling rate for us to be confidently used for rigorous analysis.
@@
     \centering
-    \includegraphics[width=0.8\linewidth,trim=0mm 12mm 0mm 35mm,clip]{images/user-study-1a.pdf}
+    \includegraphics[width=0.8\linewidth,trim=15mm 12mm 15mm 35mm,clip]{images/user-study-1a.pdf}
     \caption{Study 1a task completion time (secs.) with 95\% confidence interval.}
@@
     \centering
-    \includegraphics[width=0.8\linewidth,trim=0mm 15mm 0mm 35mm,clip]{images/user-study-1b.pdf}
+    \includegraphics[width=0.8\linewidth,trim=12mm 15mm 12mm 35mm,clip]{images/user-study-1b.pdf}
     \caption{Study 1b task completion time (secs.) with 95\% confidence interval.}
@@
 The ideal fixes for these two tasks are placed relatively early in the source code (both in the first two lines of the source code).
-Users simply reading top to bottom could quickly identify the error without needing to skip unrelated sections of code using the information provided by /chameleon{}.
-This reduced the apparent benefit of /chameleon{} in these tasks.
+Users simply reading top to bottom could quickly identify the error without needing to skip unrelated sections of code using the information provided by \chameleon{}.
+This reduced the apparent benefit of \chameleon{} in these tasks.
 On the other hand, task 9 is the lengthiest task of all.
-It also involves more complex features, such as mutually recursive type definitions.
+It also involves deeply nested type definations that are harder to follow in mind.
 
@@
 We offset this by using new code challenges in every study and conducting trial runs before data collection to bring new participants up to speed.
-Conducting studies remotely and unsupervised there is no way to intervene when users encounter usability issues.
+Conducting studies remotely and unsupervised left us no means to intervene when users encounter usability issues.
 To mitigate this, we conducted cognitive walkthroughs and sandbox pilots before running each study.
@@
 Future evaluation would benefit from using more realistic tasks.
-In studies 1b and 2, we used more challenging tasks compared to study 1a.
-However, the complexity is just short of what professional Haskell programmers may face in the production Haskell codebase.
+The tasks in our human studies do not get as complex as professional Haskell programmers may face in a typical production codebase.
 It would be interesting to see how \chameleon{} is used against type errors that span multiple files and packages and include more confusing abstractions, like Monads, Monad transformers, and Lenses.
-
-% The online study limited the amount of qualitative data we could gather from the users.
-Although we handed out debriefing surveys after each study, it is no substitute for more in-depth qualitative study methods.
-Inviting users for speak-aloud walkthroughs and interviews are viable methods to unveil deeper usability issues about \chameleon{}.

### 2023-03-08 sections/introduction.tex
Camera Ready

@@
 These features are integrated into a debugging environment and can be enabled or disabled separately based on the programmers' preferences and debugging needs.
-\chameleon{} is open-source and is available in web~\cite{anonymous_chameleon_2022} and desktop versions.
+\chameleon{} is open-source and is available at ~\cite{anonymous_chameleon_2022}.
 % \todo{Introduce Haskell earlier?
@@
 Further, programmers who actively use \chameleon{} interactive features fix type errors faster than simply reading the type error output.
+Although \chameleon{} is designed to work with
+the Haskell language, we plan to extend the underlying ideas to work with other strongly typed languages.

### 2023-03-08 sections/related-work.tex
Camera Ready

@@
 \subsection{Interactive Debugging}
-The original Chameleon provides a certain level of ``interactive'' debugging through a command-line interface~\cite{stuckey_interactive_2003}, allowing the programmer to query different parts of the program.
+% The original Chameleon provides a certain level of ``interactive'' debugging through a command-line interface~\cite{stuckey_interactive_2003}, allowing the programmer to query different parts of the program.
 A recent study \cite{tsushima_type_2021} provided an interactive type debugging system that restores type information based on the output from CFT (Counter-factual typing)~\cite{chen_counter-factual_2014, chen_efficient_2020, chen_improving_2022} inference.
@@
 
-A GUI provides programming tools with immediate feedback on changes and the ability to show complex visualizations.
-Many tools aim to improve the debugging experience.
+Modern programming tools can offer alternative methods of code authoring, display real-time feedback and reveal complex programming contexts through visualizations.
+Many tools aim to improve the debugging experience using such capabilities.
 We list two.

### 2023-03-08 sections/walkthrough.tex
Camera Ready

@@
 
-For the task shown in section \ref{sub:balanced}, if Maxine is not satisfied by
-the options provided by \chameleon{}, by switching to advanced
-mode, she has access to the debugging steps.
+% For the task shown in section \ref{sub:balanced}, if Maxine is not satisfied by
+% the options provided by \chameleon{}, by switching to advanced
+% mode, she has access to the debugging steps.
 
-First, Maxine clicks on step 5 (Fig. \ref{fig:advanced-mode-step5}) and verifies
+To illustrate the deduction steps with the task shown in section \ref{sub:balanced},  first, Maxine clicks on step 5 (Fig. \ref{fig:advanced-mode-step5}) and verifies
 that the two occurrences of \texttt{evens} are supposed to be identical, and the

### 2023-03-08 main.tex

@@
 
+\makeatletter % changes the catcode of @ to 11
+\newcommand{\linebreakand}{%
+  \end{@IEEEauthorhalign}
+  \hfill\mbox{}\par
+  \mbox{}\hfill\begin{@IEEEauthorhalign}
+}
+\makeatother % changes the catcode of @ back to 12
+
 % Comment out second line to disable.
@@
 
-\ignore{
-\author{\IEEEauthorblockN{1\textsuperscript{st} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+%\ignore{
+\author{\IEEEauthorblockN{Shuai Fu}
+\IEEEauthorblockA{
+\textit{Faculty of Information Technology} \\
+\textit{Monash University}\\
+Clayton, Australiay \\
+\texttt{shuai.fu@monash.edu}}
 \and
-\IEEEauthorblockN{2\textsuperscript{nd} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\IEEEauthorblockN{Tim Dwyer}
+\IEEEauthorblockA{
+\textit{Faculty of Information Technology} \\
+\textit{Monash University}\\
+Clayton, Australiay \\
+\texttt{tim.dwyer@monash.edu}}
 \and
-\IEEEauthorblockN{3\textsuperscript{rd} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\IEEEauthorblockN{Peter J. Stuckey}
+\IEEEauthorblockA{
+\textit{Faculty of Information Technology} \\
+%\textit{Data Science and Artificial Intelligence} \\
+\textit{Monash University}\\
+Clayton, Australia \\
+\texttt{peter.stuckey@monash.edu}}
 \and
-\IEEEauthorblockN{4\textsuperscript{th} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\linebreakand 
+\IEEEauthorblockN{Jackson Wain}
+\IEEEauthorblockA{\textit{Faculty of Information Technology} \\
+\textit{Monash University}\\
+Clayton, Australia \\
+\texttt{0000-0003-2006-3538}
+}
 \and
-\IEEEauthorblockN{5\textsuperscript{th} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
-\and
-\IEEEauthorblockN{6\textsuperscript{th} Given Name Surname}
-\IEEEauthorblockA{\textit{dept. name of organization (of Aff.)} \\
-\textit{name of organization (of Aff.)}\\
-City, Country \\
-email address or ORCID}
+\IEEEauthorblockN{Jesse Linossier}
+\IEEEauthorblockA{\textit{Faculty of Information Technology} \\
+\textit{Monash University}\\
+Clayton, Australia \\
+\texttt{0000-0001-6782-7019}
 }
+
 }
+%}
 
@@
 Programmers can interactively verify the steps of reasoning against their intention.
-Through three studies involving real programmers, we showed that  \chameleon{} is more effective in fixing type errors than traditional text-based error messages.
+Through three studies involving actual programmers, we showed that \chameleon{} is more effective in fixing type errors than traditional text-based error messages.
 This difference is more significant in harder tasks.

### 2023-03-08 sections/chameleon.tex

@@
 
-The \chameleon{} debugging interface provides three main features to visualize the relation between the error statement and locations in code and to explore the explanation of each error location given by the type inference engine.
+The \chameleon{} debugging interface provides three main features to visualize and explain type errors.
+
+% the relation between the error statement and locations in code and to explore the explanation of each error location given by the type inference engine. 
 %Further, programmers  can turn on and off features to suit their preferences and debugging needs.
@@
 The types reconstructed from Chameleon may be \texttt{(a, Int)} and \texttt{(a, String)}.
-Although the recovered types are theoretically correct, they introduce a new type variable \texttt{a}, making the error message harder to understand.
+Although the recovered types are theoretically correct, they introduce the notation \texttt{a}, which denotes a generic type variable that can be any type, making the error message harder to understand.
 To solve this issue in \chameleon{}, for each constraint \texttt{c} in the MUS, we find a maximally satisfiable subset (MSS) from all the constraints that contain every other element of MUS but not \texttt{c}.
-These maximally satisfiable subsets, while not helpful to inform error locations, will produce the most concrete conflicting types, see Fig.~\ref{fig:compare-to-original}.
+These maximally satisfiable subsets, while not helpful in error localization, will produce the most concrete types, see Fig.~\ref{fig:compare-to-original}.
+Using concrete types, such as \texttt{Int} and \texttt{String},  provides extra information to programmers.
+With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space.
+However, a type of \texttt{(a, Float)} does not preserve such information.
 
@@
     \caption{
-Reporting the same type error, Chameleon uses
+Reporting the same type error, Chameleon uses more abstract types
 \texttt{Int -> a} and \texttt{Char -> a}, while \chameleon{} uses the 
-concrete types \texttt{Int -> Bool} and \texttt{Char -> Bool}.
+concrete types (types that do not contain type variables) \texttt{Int -> Bool} and \texttt{Char -> Bool}.
     }
@@
 
-Using concrete types provides extra information to programmers.
-With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space.
-However, a type of \texttt{(a, Float)} does not preserve such information.
-If the types get too verbose for programmers to identify the conflict between the two signatures, it is always possible to mute the overlapping information and highlight the differences from the debugging front end.
+% If the types get too verbose for programmers to identify the conflict between the two signatures, it is always possible to mute the overlapping information and highlight the differences from the debugging front end.
 
@@
 Similar to the type explanation system  in ~\cite{jun_explaining_2002},  \chameleon{} is able to produce a human-readable explanation, but for type errors.
-This is achieved by annotating nodes in the abstract syntax tree with constraints and the typing rules used.
+This is achieved by annotating nodes in the abstract syntax tree with constraints and the type inference rules used.
 We generate an inference history from constraints and accompanying annotations.

### 2023-03-08 sections/discussion.tex

@@
 We speculate that one factor of this result is that  \chameleon{} helps to develop debugging plans.
-We observed that when working with \chameleon{}, programmers form different debugging tactics to attack the problem.
-Among the \textit{high} interactivity participants in user study 4, some programmers cycle through deduction steps as a guide to reading source code; some programmers navigate to both ends of the deduction steps where types are normally grounded and concrete.
+We observed that when working with \chameleon{}, programmers form different debugging plans to attack the problem.
+Among the \textit{high} interactivity participants in user study 4, some programmers cycle through deduction steps as a guide to reading source code; some navigate to both ends of the deduction chain where types are normally grounded and concrete.
 In contrast, \textit{minimal} interactive participants generally form similar plans, including carefully reading the program text and manually annotating expressions based on their understanding of the program.
@@
 A recent study on working memory \cite{crichton_role_2021} suggested this approach may provide a positive effect in helping programmers manage cognitive load and free up working-memory space for high-level thinking.
+
+% \subsection{Implication for future studies}
+% Our study aims to contribute to the wider understanding of program comprehension in the context of advanced type systems.
+One promising research area to extend our results is to apply the design of \chameleon{} to other languages with high demand for explainability\cite{ferdowsi_usability_2023}.
+
+% Colorization and overlay widgets are used to visualize relations and orders that may not conform to the source code structures.
+These techniques fit well to types because the introduction and elimination of typing constraints tend to be scattered throughout the program.
+A interesting question is whether these techniques can also be applied to other program properties, such as memory allocation and freeing, as well as message dispatching and consumption, which also exhibit similar patterns.
+
+% Finally, we employed MUS (minimal unsatisfiable subset) for fault localization and types recovery, but we are also curious about the possibility of further analysis based on MUS to power advanced debugging systems, such as efficient and accurate type error repair by combining our results with mutation testing or solution searching based on machine learning.
 

### 2023-03-09 sections/chameleon.tex
Minor changes

@@
 This framing is standard practice in most typed languages.
-However, what is the \texttt{expected} and what is \texttt{actual} are a side effect of different unification orders rather than the intention of the programmer.
+However, what is \texttt{expected} and what is \texttt{actual} are a side effect of different unification orders rather than the intention of the programmer.
 In both forms, the error message may lead programmers to falsely believe the validity of parts of code and wrongly accuse others.
@@
 \ref{fig:anatomy}-C).
-The features accessible from different modes are summarised in the following table:%shown in table~\ref{tab:chameleon-features}.
+The features accessible from different modes are summarized in table~\ref{tab:chameleon-features}.%shown in table~\ref{tab:chameleon-features}.
 
-% \begin{table}
-%     \centering
+\begin{table}
+    \centering
 \begin{scriptsize}
@@
 \end{scriptsize}
-%     \caption{\chameleon{} modes and features}
-%     \label {tab:chameleon-features}
-% \end{table}
+    \caption{\chameleon{} modes and features}
+    \label {tab:chameleon-features}
+\end{table}
 

### 2023-03-09 sections/discussion.tex
Minor changes

@@
 
-This paper presents the interactive type debugging tool \chameleon{} and charts the evolution of its design across several iterations in response to user evaluation and feedback, as well as examining the effectiveness of the general approach compared to traditional static type error messages.
-We found that programmers using \chameleon{} type compare tool are able to debug errors faster than using traditional text-based error messages.
+This paper presents the interactive type debugging tool \chameleon{} and charts the evolution of its design across several iterations in response to user evaluation and feedback, as well as examines   the effectiveness of the general approach compared to traditional static type error messages.
+We found that programmers using the \chameleon{} type compare tool are able to debug errors faster than using traditional text-based error messages.
 This effect is shown more clearly when the task is not trivial.
@@
 \subsection{Externalize Intermediate Typing Information}
-We speculate another factor of the effectiveness \chameleon{} interactive debugging tools is they help programmers effectively chunk intermediate information.
+We speculate another factor of the effectiveness of \chameleon{} interactive debugging tools is that they help programmers effectively chunk intermediate information.
 With the program shown in Listing~\ref{listing:2}, \chameleon{} offers two candidate expressions: \texttt{f} can be typed as \texttt{Int -> Bool} or \texttt{Char -> Bool}; \texttt{z} can be typed as \texttt{Int} or \texttt{Char}.

### 2023-03-09 sections/evaluation.tex
Minor changes

@@
 At each task, a participant receives a single Haskell file that contains one or more type errors.
-They was then asked to correct the code with the help of the given tool.
+They were then asked to correct the code with the help of the given tool.
 
@@
 Some users used the features extensively, while others completed the tasks without actively exploring the given information.
-Based on this discrepancy, we divided the users into three groups:
-
+Based on this discrepancy, we divided the users into three groups in table~\ref{tab:interaction-level}.
+\begin{table}
+    \centering
+\begin{scriptsize}
 \begin{small}
@@
   | >{\hsize=.7\hsize \raggedright\arraybackslash}X  | }
-
     \hline
@@
 \end{small}
+\end{scriptsize}
+    \caption{Levels of programmer interaction and their description}
+    \label {tab:interaction-level}
+\end{table}
 
@@
 On the other hand, task 9 is the lengthiest task of all.
-It also involves deeply nested type definations that are harder to follow in mind.
+It also involves deeply nested type definitions that are harder to follow in mind.
 
@@
 47\% staying in the same mode.
-Second, to answer \textbf{RQ2.3} when changing modes, programmers generally switch to the more informative modes instead of, the more concise ones.
+Second, to answer \textbf{RQ2.3} when changing modes, programmers generally switch to the more informative modes instead of the more concise ones.
 \begin{figure}

### 2023-03-09 sections/related-work.tex
Minor changes

@@
 Through a series of mixed-method studies, Prather showed~\cite{prather_novices_2017} that ECEM has a positive result in understanding compiler errors.
-Decafe~\cite{becker_effective_2016} is a tool that can rephrase Java compiler error messages into an enhanced version.
+Decaf~\cite{becker_effective_2016} is a tool that can rephrase Java compiler error messages into an enhanced version.
 In a study of over 200 CS1 students, Decaf was shown to reduce overall errors in their coding practices.
@@
 Most principles and suggestions are followed in \chameleon{} in constructing error statements.
-However, these earlier studies were not targeting type errors alone but general compiler errors (some even include run time errors).
+However, these earlier studies were not targeting type errors alone but general compiler errors (some even include runtime errors).
 The nuances of type errors, such as alternative typing, were not considered.
@@
 However, they focus on different aspects of the debugging process.
-Java Whyline mainly tackles the problem of unintended run-time behavior, while Hazel Tutor specializes in development assistance supported by type holes.
+Java Whyline mainly tackles the problem of unintended runtime behavior, while Hazel Tutor specializes in development assistance supported by type holes.
 

### 2023-03-10 sections/introduction.tex
Latest

@@
 Although \chameleon{} is designed to work with
-the Haskell language, we plan to extend the underlying ideas to work with other strongly typed languages.
+the Haskell language, we plan to extend the underlying ideas to work with other strongly typed languages, such as Rust or TypeScript..

### 2023-03-10 sections/chameleon.tex
Latest
Squeeze

@@
 In the type error message, only the \texttt{addPair} expression on line 4 was blamed.
-In this small example, the whole context is visible, but it can be more problematic in large programs where the lines contributing to the type error are far apart in the source code.
+In this small example, the whole context is visible, but it can become problematic in large programs where the lines contributing to the type error are far apart in the source code.
 
@@
 But the details of this process are hard to explain to users and are usually not reported by compilers.
-For the typical type error shown in Fig.~\ref{fig:motivation-example}, the evidence for the type error is gathered from the previous two declarations.
+For the typical type error shown in Fig.~\ref{fig:motivation-example}, the evidence for the type error is gathered from the previous declarations.
 These have to be rediscovered by programmers using less rigorous methods.
@@
 
-\noindent\textbf{Do not presume} which expression is to blame for the type error based on the order of computation or which possible type for an expression is 'actual' or 'expected'.
+\noindent\textbf{Do not presume} which expression is to blame for the type error based on the order of computation or which possible type for an expression is `actual' or `expected'.
 
@@
 \chameleon{} comprises two parts: a type inference engine and a novel interactive debugging interface. 
-This separation of concerns allows us to (1) adapt \chameleon{} into multiple front ends, such as IDE extensions, desktop applications, and web applications; and (2) reuse the debugging interface for different programming languages by replacing the back-end type inference engine.
+% This separation of concerns allows us to (1) adapt \chameleon{} into multiple front ends, such as IDE extensions, desktop applications, and web applications; and (2) reuse the debugging interface for different programming languages by replacing the back-end type inference engine.
 The debugging interface is designed from the ground up; the type inference engine is a re-implementation of the original Chameleon with several novel improvements, as described in Section \ref{sec:typeinferenceengine}.
@@
 \begin{center}
+
     \begin{tabular}{ l l  }
@@
      Balanced mode & Type Compare Tool \\
-     & Candidate Expression Card \\  \hline
+     & Candidate Expression Cards \\  \hline
      Advanced mode & Type Compare Tool \\
-     & Candidate Expression Card \\
+     & Candidate Expression Cards \\
      & Deduction Steps \\
@@
 % \end{lstlisting}
+\definecolor{bg}{rgb}{0.95,0.95,0.95}
 \begin{listing}[!ht]
-\begin{minted}{haskell}
+\begin{minted}[bgcolor=bg,linenos]{haskell}
     if a then b else c
@@
 \end{minted}
-\vspace{-3mm}
+\vspace*{-9mm}
 \caption{A simple program that is ill-typed. It generates two constraints from line 1 and one constraint from line 2. }
@@
 This explanation facilitates the deduction steps (Section \ref{sub:deduction-steps}).
+\looseness=-1

### 2023-03-10 sections/discussion.tex
Squeeze

@@
 This paper presents the interactive type debugging tool \chameleon{} and charts the evolution of its design across several iterations in response to user evaluation and feedback, as well as examines   the effectiveness of the general approach compared to traditional static type error messages.
-We found that programmers using the \chameleon{} type compare tool are able to debug errors faster than using traditional text-based error messages.
+We found that programmers using \chameleon{} are able to debug errors faster than using traditional text-based error messages.
 This effect is shown more clearly when the task is not trivial.
-We found that programmers who actively use \chameleon{}'s interactive features (candidate expression cards and deduction steps) are more efficient in fixing type errors than passively reading the type error output.
+We found that programmers who actively use \chameleon{}'s interactive features are more efficient in fixing type errors than passively reading the type error output.
 In this section, we will discuss a few interpretations of the results.
@@
 Although traditional compiler error message tools initially show fewer locations, these may be incomplete, meaning that programmers have to expand the reading span without clear guidance.
-In contrast, Chameleon shows more error locations initially.
+In contrast, \chameleon{} shows more error locations initially.
 However, the completeness of error locations assures programmers which part of the source code can be safely skipped.
@@
 \begin{listing}
-\begin{minted}{haskell}
+\begin{minted}[bgcolor=bg,linenos]{haskell}
 f z     
@@
 \end{minted}
-\caption{In this simple program, \chameleon{} reports the error happened in \texttt{f} and \texttt{z}.}
+\vspace*{-9mm}
+\caption{A simple Haskell program with a type error}
 \label{listing:2}

### 2023-03-10 sections/evaluation.tex
Squeeze

@@
 From this result, we hypothesized that we might observe a more significant difference using tasks with lengthier and more realistic source code. 
-This hypothesis is also supported by the most common feedback (7 out of 35) that the tasks were too trivial to invite meaningful evaluation.
+This hypothesis is also supported by the most common feedback that the tasks were too trivial to invite meaningful evaluation.
 One participant said, ``Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."
@@
 \noindent\begin{tabularx}{\linewidth}{ 
-  | >{\hsize=.3\hsize}X 
-  | >{\hsize=.7\hsize \raggedright\arraybackslash}X  | }
+  | >{\hsize=.26\hsize}X 
+  | >{\hsize=.74\hsize \raggedright\arraybackslash}X  | }
     \hline
@@
 This observation suggests two characteristics of using multi-mode debugging tools.
-First, to answer \textbf{RQ2.2} programmers are generally reluctant to switch modes: 53\% changing modes vs.
+First, to answer \textbf{RQ2.2} programmers are roughly splitted in this matter: 53\% changing modes vs.
 47\% staying in the same mode.

### 2023-03-10 sections/related-work.tex
Squeeze

@@
 Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error.
-Internally labeled constraints and Minimal Unsatisfiable Set (MUS) manipulation are used to generate these slices.
+Internally labeled constraints and Minimal Unsatisfiable Subset (MUS) manipulation are used to generate these slices.
 The language supported in Haack's work was a subset of Standard ML.
@@
 Chameleon also supported advanced type-level features (type classes and functionally dependent types).
-The project also introduced the ability to query type information through a command line-style interface.
+The project also introduced the ability to query type information through a command line interface.
 Although Chameleon was firmly grounded in results from type theory, its designs were never evaluated with user studies.

### 2023-03-10 sections/chameleon.tex
Minor fixes

@@
 When activating a step, some highlights switch color.
-The message in the explanation layer changes as well.
+The message in the explanation layer changes accordingly.
 A program in Fig.
@@
 These maximally satisfiable subsets, while not helpful in error localization, will produce the most concrete types, see Fig.~\ref{fig:compare-to-original}.
-Using concrete types, such as \texttt{Int} and \texttt{String},  provides extra information to programmers.
+Concrete types, such as \texttt{Int} and \texttt{String},  often provides extra information to programmers.
 With a type of \texttt{(Float, Float)}, programmers may want to convey a point in 2d space.

### 2023-03-10 sections/discussion.tex
Minor fixes

@@
 We observed that when working with \chameleon{}, programmers form different debugging plans to attack the problem.
-Among the \textit{high} interactivity participants in user study 4, some programmers cycle through deduction steps as a guide to reading source code; some navigate to both ends of the deduction chain where types are normally grounded and concrete.
+Among the \textit{high} interactivity participants in user study 2, some programmers cycle through deduction steps as a guide to reading source code; some navigate to both ends of the deduction chain where types are normally grounded and concrete.
 In contrast, \textit{minimal} interactive participants generally form similar plans, including carefully reading the program text and manually annotating expressions based on their understanding of the program.
@@
 \vspace*{-9mm}
-\caption{A simple Haskell program with a type error}
+\caption{\chameleon{} reports an error in the expressions \texttt{f} and \texttt{z}}
 \label{listing:2}

### 2023-03-10 sections/evaluation.tex
Minor fixes

@@
 In study 1a, the tasks were taken from the exercises of the Haskell programming class in the authors' institute.
-These tasks cover a variety of type errors.
 In the second study, the tasks are sourced from the top 20 Haskell topics on GitHub~\cite{github_github_2022}.
@@
 From this result, we hypothesized that we might observe a more significant difference using tasks with lengthier and more realistic source code. 
-This hypothesis is also supported by the most common feedback that the tasks were too trivial to invite meaningful evaluation.
+This hypothesis is also supported by the most common feedback claiming that the tasks were too trivial to invite meaningful evaluation.
 One participant said, ``Looks nicer than GHC, but without trying it on something more complicated, I cannot conclude whether it would help me in practice."

### 2023-03-10 sections/related-work.tex
Minor fixes

@@
 Type error slicing~\cite{haack_type_2004} is a technique that finds locations that are complete and minimal for the type error.
-Internally labeled constraints and Minimal Unsatisfiable Subset (MUS) manipulation are used to generate these slices.
+Internally labeled constraints and Minimal Unsatisfiable Subset (MUS) generation are used to generate these slices.
 The language supported in Haack's work was a subset of Standard ML.
@@
 In a study of over 200 CS1 students, Decaf was shown to reduce overall errors in their coding practices.
-Berik proposed a framework~\cite{barik_how_2018} for constructing compiler error messages based on argumentation theory.
-In this proposal,  Berik argued that error messages following a simple argumentation layout or an extended argumentation layout are more human-friendly.
+Berik proposed a framework~\cite{barik_how_2018} for constructing compiler error messages based on argumentation theory, and showed that error messages following a simple argumentation layout or an extended argumentation layout are more human-friendly.
 These works show the significance of improving the language in the compiler error messages.

### 2023-03-10 sections/walkthrough.tex
Minor changes
Minor fixes

@@
 a temporary list of even integers (hence it is of \texttt{[Int]} types), Maxine
-knows the Possible type 2 is unintended. The locations with blue highlights must
+concludes that the Possible type 2 is unintended. The locations with blue highlights must
 contain the cause. It does not take long for Maxine to realize the list 
@@
             explains that \texttt{filter} is applied to 
-            the function \texttt{isEven}. Assisting by 
+            the function \texttt{isEven}. Assisted by 
             the type of \texttt{filter} in the 
